Roles of G-protein coupled receptors and mechanosensitive ion channels in pressure-induced chronotropy of lymphatic vessels

Version 1

Abstract

Active lymph pumping relies on the spontaneous contractions of collecting lymphatic vessels, with frequencies that are exquisitely sensitive to changes in intraluminal pressure. This homeostatic and mechanosensitive mechanism, termed pressure-induced lymphatic chronotropy, enables lymph transport to be matched to the filling state of the lymphatic capillaries. We investigated the mechanistic basis of pressure-induced chronotropy using ex vivo contraction assays of mouse popliteal collecting vessels, in which contraction frequency increased >10-fold with a 5 cmH2O pressure change. The contractile, electrophysiological and transcriptional similarities between lymphatic muscle and arterial smooth muscle led us to hypothesize that pressure-dependent chronotropy shares a parallel signaling process to pressure-induce arterial depolarization/constriction. Thus, we investigated two major mechanisms: 1) pressure-induced activation of mechanosensitive cation channels, including TRPC6, TRPM4, PKD1/2, TRPV2 and ENaC, and 2) mechano-activation of GNAQ/GNA11-coupled GPCRs that would generate second messengers to activate those channels. We combined contraction assays with scRNAseq analysis of the respective targets and made maximum use of transgenic mice to avoid non-specific effects of pharmacological inhibitors, particularly those used to block TRP channels. Our findings rule out significant roles for TRP and other mechanosensitive channels implicated in myogenic constriction, as well as channels implicated in ionic pacemaking of other tissues, and instead support a scheme whereby mechano-activation of GNAQ/GNA11-coupled GPCRs generates IP3, which induces SR Ca2+ release through IP3R1 and drives depolarization through the activation of ANO1 Cl− channels.

Introduction

The lymphatic system is critically important for the maintenance of interstitial fluid balance. Because the Starling forces governing fluid movement across blood capillaries favor net filtration in most tissues (Levick & Michel, 2010), filtered fluid and protein will accumulate in the interstitium unless it is removed by the lymphatic system. Lymphatic capillaries are optimized to reabsorb that fluid, which can exceed 8 L·day−1 (Renkin, 1986). Once reabsorbed, lymph is transported by a combination of passive and active transport mechanisms through the collecting lymphatic vessel network, first to lymph nodes and eventually back into the venous system. Active lymph pumping relies on the spontaneous contractions of collecting lymphatic vessels—events that are initiated by action potentials (APs) in lymphatic muscle cells (LMCs). These contractions are relatively large in amplitude, with ejection fractions often exceeding 80% and sufficient to propel lymph through a system of one-way valves that ensure its unidirectional movement (Davis et al., 2025). Active pumping accounts for the majority of lymphatic transport in the lower legs of humans during quiet standing (Olszewski & Engeset, 1980) and facilitates lymph return against adverse hydrostatic pressure gradients that develop under gravitational loads (Davis et al., 2012; Scallan et al., 2016; Li et al., 2022). An important feature of the active lymph pump system is the exquisite sensitivity of the spontaneous contraction frequency of collecting lymphatic vessels to changes in their intraluminal pressure, enabling lymph transport to be matched to the filling state of the lymphatic capillary network (Scallan et al., 2012). This homeostatic mechanism is termed “pressure-induced lymphatic chronotropy” (Zawieja et al., 2019). In popliteal collecting vessels of the mouse hindlimb, the spontaneous contraction frequency can increase from 2 to 20 contractions·min−1 as pressure is elevated from 0.5 to 10 cmH2O, i.e., a 10-fold increase over the physiological pressure range of those vessels (Davis et al., 2020; Davis et al., 2023a; Davis et al., 2023b). Contraction amplitude increases modestly over a portion of the same pressure range (Scallan et al., 2012), but modulation of contraction frequency is the primary determinant of active lymph transport.

Pressure-induced chronotropy reflects an underlying mechanotransduction process that ultimately involves control of an intrinsic pacemaker in LMCs by intraluminal pressure. Although the specific force sensing mechanism is unknown, contractile, electrophysiological and transcriptional similarities between LMCs and vascular smooth muscle cells (VSMCs) support the hypothesis that pressure-dependent chronotropy shares a parallel signaling process to the myogenic response of VSMCs. The myogenic response refers to in rapid increase in arterial pressure induces distention of a small artery / arteriole, which then constricts over time until a new, stable diameter lower than the original diameter is reached (Davis, 2012). A similar constriction often occurs in collecting lymphatic vessels simultaneously with pressure-induced chronotropy (Davis et al., 2009). Pressure-induced arterial constriction is preceded by VSM depolarization (Harder, 1984) and several lines of evidence support a critical role for the activation of (putative) mechanosensitive cation channels in that process—in particular the involvement of three TRP channel family members: TRPC6, TRPM4 and PKD1/2.

The first evidence for a molecularly identified ion channel involved in myogenic constriction came from the demonstration that single channel currents in VSMCs isolated from rat cerebral arteries were activated by patch pipette suction and those currents, along with pressure-induced cerebral artery constrictions and depolarizations, were attenuated after downregulation of TRPC6with antisense oligonucleotides (Welsh et al., 2002). TRPM4 channels were likewise implicated after inward, TRPM4-like currents in rat pial artery myocytes were found to be activated by membrane stretch, and because antisense- or siRNA-mediated knock down of TRPM4 significantly attenuated pressure-induced depolarization and myogenic tone development in those arteries (Earley et al., 2004b; Gonzales et al., 2014). However, results from Trpc6−/− and Trpm4−/− mice tended to contradict the above studies in finding no differences or even enhanced arterial myogenic constriction in the global knock out animals (Dietrich et al., 2005; Mathar et al., 2010; Schleifenbaum et al., 2014). Evidence also supports roles for PKD2 (previously described as TRPP1) channels in myogenic constriction, but it, too, is controversial. Jaggar and colleagues reported that shRNA-mediated down-regulation of Pkd2attenuated pressure-induced depolarization and myogenic tone in rat pial arteries (Narayanan et al., 2013) and that pressure-induced constriction was blunted in hindlimb arteries isolated from Myh11-CreERT2;Pkd2f/f (smooth muscle knockout, smKO) mice (Bulley et al., 2018). However, a study by Honoré and coworkers suggested that stretch-activated cation currents and myogenic constriction in mouse mesenteric artery smooth muscle were mediated by PKD1 channels (previously termed TRPP2), and that PKD2 channels played a counteracting role (Sharif-Naeini et al., 2009). Myogenic responses could not be tested in global Pkd2 KO mice because constitutive deletion of Pkd2 is embryonically lethal (Wu et al., 1998).

More recent work points to G-protein-coupled receptors (GPCRs) rather than TRP or other ion channels as the primary mechanosensing elements in the arterial myogenic response. Studies of cardiomyocyctes first demonstrated that the angiotensin receptor, AT1R, a GNAQ/GNA11-coupled GPCR, could be activated by mechanical stress through a mechanism independent of its ligand (Zou et al., 2004). Gudermann and coworkers provided compelling support for this mechanism using patch-clamped HEK293 cells, co-expressing AT1R and TRPC6 channels, with hypoosmotic swelling as a mechanical stimulus (Mederos y Schnitzler et al., 2008). Two critical findings were 1) that TRPC6 current was activated by hypoosmotic swelling only if the cells co-expressed Agtr1 (Mederos y Schnitzler et al., 2008) and 2) that the swelling-induced currents were largely prevented by inverse AT1R agonists. Agtr1 could be replaced by another GNAQ/GNA11-coupled GPCR, or Trpc6 by another DAG-sensitive TRP channel (Mederos y Schnitzler et al., 2008). Subsequent studies from multiple laboratories, using selective pharmacological inhibitors of AT1R, acute Agtr1 knockdown, and/or mice deficient in Agtr1 or other GNAQ/GNA11-coupled GPCRs, confirmed that basic principle in VSMCs: that pressure-induced constriction of many different types of arteries is transduced primarily, or exclusively, by GNAQ/GNA11-coupled GPCRs (Blodow et al., 2014; Harraz et al., 2014; Li et al., 2014b; Schleifenbaum et al., 2014; Pires et al., 2017; Bjorling et al., 2018). Helix 8 in the GPCR c-terminus was subsequently identified as the essential structural motif endowing mechanosensitivity (Erdogmus et al., 2019). The collective data support a model in which phospholipase C is activated downstream from GNAQ/GNA11-coupled GPCRs in response to elevated pressure, leading to increased production of DAG and IP3/Ca2+ release, which activate TRPC6 and TRPM4 channels, respectively, to depolarize the cell and increase the open probability of L-type voltage-gated Ca2+ channels, thereby promoting global Ca2+ influx and contraction (Davis et al., 2023c). In a similar manner, TRPM4 and PKD2/2 channels might be activated by Ca2+ influx/release downstream from mechano-activation of a GPCR or another Ca2+-permeable ion channel such as PIEZO1 (Peyronnet et al., 2013) or TRPV4 (Swain & Liddle, 2021).

Whether a similar scheme is operative in LMCs is unknown. Each lymphatic contraction is preceded by a near-linear depolarization from the “resting” potential to threshold potential, termed the diastolic depolarization, which is the primary determinant of AP/contraction frequency (Zawieja et al., 2025). Thus, lymphatic pressure elevation manifests not as a stepwise increase in baseline potential LMCs (von der Weid et al., 2014; Davis & Zawieja, 2018) but rather as an increase in the diastolic depolarization slope and a reduction in time required to reach threshold for AP firing (Zawieja et al., 2018b; Zawieja et al., 2019). Previously, we found that regulation of the diastolic depolarization slope was mediated largely by activation of the Ca2+ activated chloride channel ANOCTAMIN 1 (ANO1, also referred to as TMEM16A). ANO1 is not intrinsically mechanosensitive, but is regulated by Ca2+ and, in LMCs specifically, by IP3 receptor 1-mediated Ca2+ release from sarcoplasmic reticulum (Zawieja et al., 2019; Zawieja et al., 2023), possibly downstream from one or more mechanosensitive GPCRs. However, other mechanisms are also involved because Ano1 deletion from smooth muscle or pharmacological inhibition of ANO1 does not completely eliminate spontaneous APs or contractions (Zawieja et al., 2019). The primary goal of the present study was to test the roles of putative VSMC mechanosensitive ion channels, channels implicated in the regulation of pacemaking in other cell types, and GNAQ/GNA11-coupled GPCRs in pressure-induced chronotropy of popliteal lymphatic vessels. An advantage in studying this particular aspect of lymphatic vessel mechanotransduction is the high sensitivity of the spontaneous contraction frequency to pressure, increasing ∼10-fold over only a 5 cmH2O pressure range in popliteal lymphatics compared to relatively modest levels of myogenic constriction (50% change) over a 60 mmHg pressure range, even in the most myogenically reactive arteries (Davis et al., 2023c). As part of our strategy, we sought to make maximum use of transgenic mice to avoid non-specific effects of pharmacological inhibitors, particularly those used to block TRP channels.

Materials and Methods

Protocol approval

All procedures were reviewed and approved by the animal care committee at the University of Missouri and complied with the standards stated in the “Guide for the Care and Use of Laboratory Animals” (National Institutes of Health, revised 2011).

Mice

WT mice [C57Bl/6 (#000664) and SVF129 (#101043) strains], Slc8a1−/− (#025943) mice and AT1aR−/−(#002682) mice were purchased from The Jackson Laboratory (Bar Harbor, Maine). Myh11-CreERT2mice were gifts from Stefan Offermans (Max Plank Institute, GDR). Sperm from Gna12−/−;Gna13f/f and Gna11−/− ;Gnaqf/f mice were gifts from Stefan Offermanns, after which the respective mice were rederived at MMRC, Columbia, MO to generate Myh11-CreERT2;Gna13f/f;Gna12−/−mice and Myh11-CreERT2;Gnaqf/f;Gna11−/−mice. Trpc6−/− and Trpc3−/− mice were gifts from Lutz Birnbaumer (NIH); Trpc6−/− ;Trpc3−/− double KO mice were generated by crossing those two strains. Myh11-CreERT2;Kcnq4f/f (encoding Kv7.4) mice were a gift of Thomas Jentsch (Free Univ. of Berlin, GDR). Piezo1f/f mice were a gift from Ardem Pataputian (UCSD) and were bred to Myh11-CreERT2 mice to generate Myh11-CreERT2; Piezo1f/f mice. Ano1f/fmice, generated in the lab of Jonathan Jaggar (Leo et al., 2021) were bred to Myh11-CreERT2 mice to obtain Myh11-CreERT2;Ano1f/f mice. Itpr1f/f mice were obtained from Ju Chen (UCSD) and bred to Myh11-CreERT2mice to generate Myh11-CreERT2;Itpr1f/fmice. Pkd2f/f mice were generated in the University of Maryland PKD Core by Jonathan Jaggar, as previously described (Bulley et al., 2018), and bred in the Jaggar laboratory to Myh11-CreERT2mice to generate Myh11-CreERT2;Pkd2f/f mice. Trpm4f/fmice were generated in the laboratory of Scott Earley and bred to Myh11-CreERT2 mice to produce Myh11-CreERT2;Trpm4f/fmice. Trpv4f/f mice were generated in the laboratory of Timothy Domeier (University of Missouri) and bred to Myh11-CreERT2 mice to produce Myh11-CreERT2;Trpv4f/f mice. Rosa26mTmGf/f mice were purchased from JAX (#007676) and bred to Myh11-CreERT2 and Myh11-CreERT2;Ano1f/f mice to produce Myh11-CreERT2;Rosa26mTmGf/f and Myh11-CreERT2;Ano1f/f; Rosa26mTmGf/fmice.

For genotyping, genomic DNA was extracted from tail clips using the HotSHOT method. Genotypes were determined by PCR with 2x PCR Super Master Polymerase Mix (Catalog # B46019, Bimake, Houston, TX) according to the provider’s instructions. Mice containing Myh11-CreERT2 were induced with five consecutive daily injections of 100 mg tamoxifen (10mg/kg i.p.; in safflower oil). Those mice were studied a minimum of 2 weeks after induction. All floxed control mice were subjected to the same induction protocol. Because the Myh11-CreERT2 used for SM-specific deletion of floxed genes is carried on the y chromosome, only male mice were used for experiments involving Myh11-CreERT2 mice or their floxed controls. For other strains, both male and female animals were used.

Vessel isolation, pressure myography, and data acquisition

Mice were anesthetized with ketamine/xylazine (100/10 mg/kg, i.p.) and placed face down on a heated tissue dissection/isolation pad. The saphenous vein was exposed by a proximal-to-distal incision along the skin of the calf and the popliteal afferent lymphatic vessels on each side of the vein were isolated as previously described (Scallan & Davis, 2013). Each vessel was then pinned with short segments of 40 µm stainless steel wire onto the SYLGARD-coated surface of a dissection chamber filled with BSA-supplemented Krebs buffer at room temperature. Once secured, the surrounding adipose and connective tissue were removed by microdissection. An isolated popliteal lymphatic vessel was then transferred to a 3-mL observation chamber, cannulated, pressurized to 3 cmH2O using two glass micropipettes (50-60 µm outside diameter) and moved to the stage of a Zeiss inverted microscope. Lymphatic segments used in these studies typically contained a single valve. Polyethylene tubing attached to the back of each glass micropipette was connected to a two-channel computerized pressure controller (Scallan et al., 2013). To minimize diameter-tracking artifacts associated with longitudinal bowing at higher intraluminal pressures, input and output pressures were briefly set to 10 cmH2O at the beginning of every experiment, and the vessel segment was stretched axially to remove longitudinal slack. The lymphatic vessel was then allowed to equilibrate at 37°C with inflow and outflow pressures set to 3 cmH2O. Constant exchange of Krebs buffer was maintained using a peristaltic pump at a rate of 0.5 mL/min. After temperature stabilized, popliteal lymphatics typically began to exhibit spontaneous contractions within 10-15 min and stabilized in a consistent contraction pattern within ∼30 minutes. Custom LabVIEW programs (National Instruments; Austin, TX) acquired real-time analog data and digital video through an A-D interface (USB-6216, National Instruments) and detected the inner diameter of the vessel at 30 fps using a Basler A641fm firewire camera (Davis, 2005). Videos of the contractile activity of lymphatic vessels were recorded for further analysis, if needed.

Assessment of ex vivo contractile function

The contractile parameters of each vessel were characterized at different levels of intraluminal pressure spanning the physiological range from 0.5 to 10 cmH2O (in successive steps as follows: 3, 2, 1, 0.5, 3, 5, 8, and 10 cmH2O). Spontaneous contractions were recorded at each pressure in typical intervals of 2 minutes, although a period of 4-5 min was sometimes required at the lowest pressure to obtain multiple contractions. During the pressure response protocol, both the input and output pressures were maintained at equal levels so that there was no imposed pressure gradient for forward flow. At the end of every experiment, all vessels were equilibrated by perfusion with calcium-free Krebs buffer containing 3 mM EGTA for 30 minutes, after which passive diameters were obtained at each level of intraluminal pressure.

Contractile function parameters

Once an experiment was completed, internal diameter traces and/or 30-fps brightfield videos of spontaneous contractions were analyzed using custom-written LabVIEW programs (Davis, 2023b, a) to detect end diastolic diameter (EDD), end systolic diameter (ESD), and contraction frequency (FREQ, computed on a contraction-by-contraction basis), with each parameter averaged over a 2-5 min period. These data were used to calculate the following commonly reported parameters that characterize the contractile function of lymphatic vessels:

where DMAX represents the maximum passive diameter (obtained after incubation with calcium-free Krebs solution) at a given level of intraluminal pressure. When frequency was zero, a value for amplitude was omitted. The relative paucity of some amplitude data at low pressures in some genotypes reflects the fact that this parameter could not be determined when frequency was zero. Each of the contractile parameters represented the average of all the recorded contractions during the measurement period at each intraluminal pressure.

Analysis of pressure-induced chronotropy

Quantification of the frequency-pressure (F-P) relationship for comparisons across genotypes required additional analysis steps, as illustrated in Fig. 1. Raw recordings of pressure and diameter (Fig. 1A) show the responses of a WT popliteal vessel to descending pressure steps from 5 to 0.5 cmH2O, with the average frequency stated for each pressure. The frequency data were then plotted as a function of pressure (Fig. 1B). As evident from the behavior of this representative vessel, frequency tended to plateau above 5 cmH2O, so the subset of data in which the response was nearly linear over the pressure range 0.5-5 cmH2O was fitted with a first-order polynomial to obtain the slope and intercept (Fig. 1C). As an alternative method, we considered calculating the ratio of the respective contraction frequencies at 0.5 and 5 cmH2O; however, in many cases no contractions occurred at P = 0.5 cmH2O, which would equate to a ratio of infinity. To avoid this problem, we computed the difference (ΔF) between the frequencies at 5 and 0.5 cmH2O. This approach gave consistent values that, like the slope, could detect blunting of the F-P relationship, but in essentially every case the results matched those of the curve fitting procedure; thus, only the F-P slope was used for statistical analyses and shown in subsequent figure plots. To enable a more comprehensive assessment of contractile function, full plots of amplitude, frequency, FPF and tone were also plotted as a function of pressure for each genotype (Fig. 1D).

Quantification of the F-P relationship.

Steps for quantitative assessment of pressure-induced chronotropy. A) Representative recording of pressure and diameter changes in a WT popliteal lymphatic vessel during inflow (Pin) and outflow (Pout) pressure steps between 0.5 and 5 cmH2O. Frequencies stated at the bottom of the trace were determined off-line using a custom peak detection program written in LabVIEW. B) Plot of frequency vs pressure for the vessel in A, showing how frequency reaches a plateau above 5 cmH2O. C) Fit of the frequency data with a first-order equation and calculation of the frequency difference (ΔF) between 5 and 0.5 cmH2O. D) Summary of contraction data for this representative WT vessel over the stated pressure range; see Methods for description of parameter calculations.

FACS

Inguinal-axillary lymphatic vessels from tamoxifen-treated Myh11-CreERT2;mTmGf/f and Myh11-CreERT2;Ano1f/f;mTmGf/fmice were dissected and cleaned of fat and connective tissue as described previously (Zawieja et al., 2018a). Cleaned vessel segments were transferred to a 1-ml tube of low-Ca2+ PSS containing (in mM): 137 NaCl, 5.0 KCl, 0.1CaCl2, 1.0 MgCl2, 10 HEPES, 10 Glucose, and 1 mg/ml BSA at room temperature for 10 min. The solution was decanted and replaced with a similar solution containing 26 U/ml papain (Sigma, St. Louis, MO) and 1 mg/ml dithioerythritol. The vessels were incubated for 30 min at 37°C with occasional agitation, then transferred to a new tube containing low-Ca2+ PSS containing 25 mg/ml collagenase H (FALGPA U/ml, Sigma), 0.7 mg/ml collagenase F (Sigma), 20 mg/ml trypsin inhibitor (Sigma), 1mg/ml elastase (Worthington), and incubated for 6 min at 37°C. The dispersed cells were sedimented by centrifugation (300 g, 4 min), resuspended in 0.6 ml PSS containing 1mM Ca solution, and filtered through a 35-µm mesh nylon filter to obtain a single cell suspension. GFP+ LMCs were sorted by fluorescence-activated cell sorting (FACS) using a Beckman-Coulter MoFlo XDP instrument with excitation laser (488 nm) and emission filter (530 ± 40 nm), a 70 µm nozzle, a sheath pressure of 45 psi and sort rate of 100 events per second. Sorting was performed at the Cell and Immunobiology Core facility at the University of Missouri.

RNA isolation and quantitative, real-time PCR

Total RNA was extracted from FACS-sorted cells using the Arcturus PicoPure RNA isolation kit (ThermoFisher Scientific, Waltham, MA) with on-column DNase I treatment (Qiagen, Valencia, CA) according to manufacturer’s instructions. RNA was eluted with nuclease-free water. Purified RNA was transcribed into cDNA using High-Capacity cDNA Reverse Transcription kit (ThermoFisher Scientific, Waltham, MA). Real-time PCR (qPCR) was performed on cDNAs prepared from each sample using 2x PrimeTime Gene Expression Master Mix (IDT, Coralville, IA) with predesigned TaqMan probes as listed in Suppl. Table 1 (IDT, Coralville, IA). Real-time PCR protocols were as follows: preheating at 95°C for 3 min, 45 cycles of two-step cycling of denaturation at 95°C for 15 sec and annealing/extension steps of 30 sec at 60°C. Data collection was carried out using a Bio-Rad CFX 96 Real-Time Detection System (software version Bio-Rad CFX Manager 3.1; Bio-Rad, Hercules, CA, USA). For analysis, the result was expressed as a ratio of target gene/reference gene (β-actin).

scRNAseq analysis

A total of 10 Rosa26mTmG mice (5 males and 5 females), without Cre and without tamoxifen treatment were used for scRNAseq analysis. Inguinal-axillary lymphatic vessels (IALVs) from both sides of each mouse were isolated and cleaned of connective and adipose tissue. The pooled vessels were digested into a single cell suspension and the cells were kept on ice until all tissues had been processed. Cells from all vessels were combined and sorted for tdTomato expression to remove debris and concentrate the cells for downstream single cell 3’ RNA-Seq libraries creation with 10x Genomics Chromium Chip and Chromium Next GEM Single Cell 3’ RNA-Seq reagents (Zawieja et al., 2025). Samples were sequenced with the NovaSeq 6000 S4-PE100 flow cell.

Mus musculus genome GRCm39 and annotation GTF (v106) from Ensembl (https://useast.ensembl.org/Mus_musculus/Info/Index) were used to build the reference index and the reads were processed using Cell Ranger (v7.0.1) using the default parameters. The quality control and filtering steps were performed using R (v4.2.1; https://www.r-project.org/). Ambient RNA was removed from the Cell Ranger output with SoupX (Young & Behjati, 2020). A doublet score for each cell was estimated using scDBlFinder [v1.12.0; (Germain et al., 2021)]. Non-expressed genes (sum zero across all samples) and low-quality cells (>10% mitochondrial genes, < 500 genes, < 1,000 UMIs per cell and doublet score <0.5) were removed with custom R scripts. Cells passing filtering were normalized/scaled (SCTransformation), dimensionally reduced (t-distributed stochastic neighbor embedding (t-SNE), uniform manifold approximation and projection (UMAP)) clustered, and hierarchically analyzed with Seurat (Hao et al., 2021; Hao et al., 2024) using the default parameters. Marker gene expression profile on cell clusters and gene co-expression were visualized using Seurat and ShinyCell R app (Ouyang et al., 2021 PMID: 33774659). The raw scRNAseq dataset is available on the NIH GEO #GSE277843 (eLIFE ref).

Solutions and chemicals

Krebs buffer contained: 146.9 mM NaCl, 4.7 mM KCl, 2 mM CaCl2·2H2O, 1.2 mM MgSO4, 1.2 mM NaH2PO4·H2O, 3 mM NaHCO3, 1.5 mM Na-HEPES, and 5 mM D-glucose (pH = 7.4). An identical buffer was prepared with the addition of 0.5% bovine serum albumin (“Krebs-BSA”). During cannulation Krebs-BSA buffer was present both luminally and abluminally; however, during the experiment the abluminal solution was constantly exchanged with plain Krebs. Ca2+-free Krebs was Krebs with 3 mM EGTA replacing CaCl2·2H2O. All chemicals were obtained from Sigma-Aldrich (St. Louis, MO), with the following exceptions: BSA (US Biochemicals; Cleveland, OH), MgSO4 and Na-HEPES (ThermoFisher Scientific; Pittsburgh, PA), ivabradine, zatebradine and losartan (Tocris). Amiloride, benzamil, ivabradine, zatebradine, BaCl2 and losartan were dissolved in water to make stock solutions. Ani9, tranilast and SET2 were dissolved in DMSO to make stock solutions. Each inhibitor was then further diluted in Krebs solution to reach the final concentrations stated.

Statistical analysis

The data were analyzed in LabVIEW, compiled in Excel and statistical analyses were performed using Prism (v.10.2; Graphpad, San Diego, CA). Statistical differences in the slopes of the F-P relationships between two groups were assessed by an unpaired, two-tailed t-test, and differences between three or more groups were tested using a one-way ANOVA with Dunnett’s post-hoc tests. Comparisons between contraction parameters as a function of pressure were made using two-way or mixed-model ANOVAs with repeated measures and Sidák’s post-hoc tests. Data are plotted as mean ± SEM. Significance levels (compared to the control group) are either stated or indicated as follows: *p<0.05; **p<0.01; ***p<0.001; ****p<0.0001; “n.s.” p≥0.05, with the color-coding of the symbols matching the respective group. In the figure legends, N refers to the number of animals and n refers to the number of vessels per group.

Results

Role of ANO1 in pressure-induced chronotropy

Single cell RNAseq analyses from dissected IALVs revealed 19 distinct cell clusters (Fig. 2A), segregating into prominent clusters/subclusters of lymphatic endothelial cells (LECs), LMCs, adventitial cells (AdvCs) and immune cells. Further analysis confirmed that Ano1 was expressed at moderate levels (relative to Acta2) in >75% of LMCs and to a lesser extent in AdvCs, but not in LECs (Fig. 2A). AdvCs contained subclusters of adventitial fibroblasts and other cell types expressing multiple stem cell markers (Zawieja et al., 2019).

Consequences of ANO1 deletion and/or inhibition.

A) UMAP plot of Ano1 expression in the various cell populations within the walls of murine IALVs that were thoroughly cleaned prior to dissociation and scRNAseq analysis, as described in (Zawieja et al., 2025). The LMC cluster (composed of at least two subclusters of 978 cells) was identified by the expression of canonical smooth muscle cell markers Acta2 (SM α-actin), Myh11 (myosin heavy chain 11), Itga8 (integrin alpha 8). Gray color in this and subsequent figures represents Acta2 expression, red/brown in this panel indicates Ano1 expression. Bubble plot shows Ano1 expression in the LMC and other cell clusters in terms of percent cells in an individual cluster expressing Ano1 (size of dot) and the Ano1 expression level relative to Acta2 (color of dot). B) Slope of frequency vs pressure (F-P) for WT (C57Bl/6) vessels with / without treatment by the ANO1 inhibitor Ani9 (1 μM), for Ano1f/f control vessels, for Ano1 smKO vessels and for the latter treated with Ani9. C) Summary plots of contraction parameters as a function of pressure for the 5 respective groups of vessels. WT N=15, n=20; WT+Ani9 N=5, n=6; Ano1 smKO N=14, n=23; Ano1f/f N=8, n=16.

Previously, we reported that ANO1 inhibition or SM-specific Ano1 deletion caused a nearly complete abrogation of pressure-induced chronotropy in mouse IALVs. However, the contraction frequency of IALVs changes only ∼1.7 fold over the pressure range from 0.5 to 10 cmH2O (Zawieja et al., 2019), compared to the much larger dynamic range (≥ 10-fold) characteristic of popliteal lymphatics (Fig. 1). Thus, we used popliteal lymphatics throughout this study to assess pressure-induced chronotropy, in anticipation that we would be able to detect even subtle changes in the response after deletion of specific genes. The F-P analysis for 20 WT popliteal lymphatic vessels is shown in Fig. 2. C57Bl/6 mice served as the WT control group for many comparisons because all transgenic strains used in the present study (with one exception) were derived from, or backcrossed into, that background. The average slope (± SEM) of the F-P relationship between 0.5 and 5 cmH2O was 2.8 ± 0.3 for WT vessels (Fig. 2B), equating to an average frequency difference of 12.1 ± 1.0 contractions·min−1 over that pressure range. The average r2 value for the linear regressions was 0.93 ± 0.02, indicating a consistently high quality of the curve fits and confirming that the slope was approximately linear between 0.5 and 5 cmH2O. The role of ANO1 in pressure-induced chronotropy of mouse popliteal lymphatic vessels was then tested after pharmacologic inhibition of ANO1 with Ani9, one of the most selective small molecule inhibitors of ANO1 (Hwang et al., 2016). Initial assessment of the Ani9 concentration-response relationship (from 30 nM to 10 μM) in WT popliteal vessels revealed that 3 μM Ani9, the concentration previously used to inhibit ANO1 in IALVs (Zawieja et al., 2019; Zawieja et al., 2023), did indeed reduce the basal contraction frequency of popliteal lymphatics (in the lower pressure range), but also caused a loss of tone and marked reduction in their contraction amplitude (not shown)—suggestive of possible off-target effects; therefore, for subsequent F-P protocols we used a lower concentration (1 μM) of Ani9, which had insignificant effects on contraction amplitude and tone (Fig. 2C) but significantly reduced the average slope of the F-P relationship from 2.8 ± 0.3 to 0.3 ± 0.2 (p<0.001; Fig. 2B).

We then tested the effects of SM-specific Ano1 deletion by comparing the F-P relationship of popliteal lymphatic vessels from tamoxifen-treated Myh11-CreERT2;Ano1f/f (Ano1 smKO) mice to Ano1f/f control mice. The results are also shown in Fig. 2B. The average slope of the F-P relationship between 0.5 and 5 cmH2O for Ano1 smKO vessels was 0.4 ± 0.2, significantly lower (p<0.001) than that (2.8 ± 0.2) for Ano1f/f vessels. Complete sets of contraction data for untreated and Ani9-treated WT popliteal lymphatics, Myh11-CreERT2;Ano1f/f and Ano1f/fpopliteal vessels, are shown in Fig. 2C. The frequency of Ani9-treated vessels was significantly reduced at most pressures but increased at higher pressures. Normalized contraction amplitudes were unchanged by Ani9. Tone was slightly but not significantly higher in Ani9-treated vessels. After Ano1 deletion, frequency was reduced to <3 contractions·min−1 at all pressures, in contrast to Ani9-treated WT vessels in which frequency was comparably reduced at low pressures but not at elevated pressures (Fig. 2C). Ano1 smKO vessels had significantly elevated levels of tone at 5 of 7 pressures, suggesting that Ani9 treatment may have interfered with tone development in a way that Ano1 deletion did not. Collectively, these results show that deletion / inhibition of ANO1 significantly impaired pressure-induced chronotropy in popliteal lymphatic vessels; however, SM-specific Ano1 deletion failed to completely abolish spontaneous contractions in 16 of 21 Ano1 smKO vessels. To test the possibility that residual pacemaking activity in Ano1 smKO vessels might be due to incomplete recombination by the inducible Myh11-CreERT2 driver, Ano1 mRNA levels in IALVs were quantified using qPCR analysis. Smooth muscle cells were digested from IALVs of Myh11-CreERT2;mTmGf/for Myh11-CreERT2;mTmGf/f;Ano1f/f mice and purified by FACS using the GFP signal. Ano1 message (referenced to α-actin levels) in LMCs from Myh11-CreERT2;mTmGf/f;Ano1f/f vessels was undetectable compared to that in LMCs from Myh11-CreERT2;mTmGf/f vessels (0 ± 0 and 1006 ± 80, respectively; mean ± SD, n=3). As a further test for possible residual ANO1 activity in Ano1 smKO vessels, we treated another group of Ano1 smKO vessels with Ani9 (1 μM), and those results are shown in Fig. 2B-C. Pacemaking activity persisted even after combined deletion and inhibition of ANO1, without significant further reductions in frequency at any pressure, suggesting that residual pacemaking in Ano1 KO vessels was not due to incomplete ANO1 knock down. Collectively, these results point to ANO1 as the major component of pressure-induced lymphatic chronotropy but also to the involvement of one or more additional mechanisms, possibly involving other LMC ion channels.

Role of PIEZO1

PIEZO1 and PIEZO2 are two ion channels that satisfy the full set of criteria for true mechanosensitivity (Coste et al., 2010; Davis et al., 2023c). PIEZO2 is expressed primarily in neurons (Woo et al., 2014; Woo et al., 2015) whereas PIEZO1 is more widely expressed (Murthy et al., 2017). PIEZO1 is critically important for shear-stress induced Ca2+ signaling in endothelium (Li et al., 2014a), for lymphatic valve and vessel development (Nonomura et al., 2018; Choi et al., 2019; Choi et al., 2024) and for hypertension-dependent arterial wall remodeling (Retailleau et al., 2015). The role of PIEZO1 in the arterial myogenic response has not been systematically tested, except for a single report showing that SM-specific Piezo1 deletion did not alter myogenic tone development in either caudal and rostral cerebellar arteries of mice [Suppl. Figs. 2-3 in (Retailleau et al., 2015)]. The possible involvement of PIEZO1 in lymphatic vessel contractility has not been reported. Our scRNAseq analysis revealed strong expression of Piezo1 message in LECs but very weak expression in LMCs (Fig. 3A). Piezo2 expression was barely detectable in either cell type. Popliteal lymphatic vessels from Myh11-CreERT2;Piezo1f/f (Piezo1 smKO) and Piezo1f/fcontrol mice were subjected to the protocol for assessing pressure-induced chronotropy. However, the data revealed no significant difference in the F-P slope between Piezo1 smKO vessels and Piezo1f/f control vessels (Fig. 3B), suggesting that PIEZO1 channels in LMCs are not required for pressure-induced lymphatic chronotropy. As a further test, we used Nestin-Cre to generate Nestin-Cre;Piezo1f/f (Piezo1 nestinKO) mice with constitutive deletion of Piezo1 both in smooth muscle and other cells expressing intermediate filaments. An analysis of vessel responses from Piezo1 nestinKO mice indicated that there was no significant impairment in pressure-induced chronotropy (Fig. 3B). Although we did not suspect the endothelium would be involved in pressure-induced chronotropy, we also generated and tested Prox1-CreERT2;Piezo1f/f (Piezo1 lecKO) mice. As expected, there was no significant difference in the average F-P slope of vessels from Piezo1 lecKO mice compared to tamoxifen-treated Piezo1f/f controls (Fig. 3B). Complete sets of contraction data for popliteal vessels from Piezo1 smKO are shown in Fig. 3C and revealed no significant differences in any of the lymphatic contractile parameters at any pressure compared to their floxed controls.

Consequences of Piezo1 deletion.

A) UMAP and bubble plots of Piezo1 and Piezo2 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Piezo1f/f, Piezo1 smKO, Piezo1 nestin KO, Piezo1 lecKO mice. C) Summary plots of contraction parameters as a function of pressure for the 4 respective groups of vessels. Piezo1f/f N=9, n=17; Piezo1 smKO N=7, n=13; Piezo1 nestin KO N=5, n=10; Piezo1 lecKO N=5, n=6.

Roles of TRPC6 and TRPC3

TRPC6 is thought to be a mechanosensitive cation channel whose activation underlies pressure-induced constriction in cerebral arteries (Welsh et al., 2002). However, studies of other arteries from Trpc6−/− mice found either no differences from control vessels or even enhanced myogenic responsiveness (Dietrich et al., 2005; Schleifenbaum et al., 2014). Because global TRPC6 deletion results in compensatory upregulation of TRPC3 (Dietrich et al., 2005), a related TRPC family member with similar properties and which is also activated by DAG, it is possible that TRPC3 could compensate for loss of TRPC6. The expression patterns for TRPC6 and TRPC3, based on scRNAseq analyses, are shown in Fig. 4A. TRPC6 was expressed at modest levels in ∼50% of LMCs and was almost exclusively confined to LMCs. TRPC3 was expressed at very low levels in only a few LMCs (from control mice). We tested the effects of global TRPC6 deletion (Trpc6−/−), global TRPC3 deletion (Trpc3−/−) and global deletion of both isoforms (Trpc6−/−;Trpc3−/−) on pressure-induced lymphatic chronotropy. As these mice were all generated in the SV129 background, control measurements were performed on popliteal vessels from SV129 mice, which showed a lower average F-P slope than WT (C57Bl/6) popliteal vessels. There were no significant differences in F-P slope between SV129 controls and Trpc6−/− or Trpc3−/− vessels, but the F-P slope of Trpc6−/−;Trpc3−/− vessels was significantly higher than that of the SV129 controls (Fig. 4B). Complete sets of contraction data for Trpc6−/−;Trpc3−/− vessels, compared to their SV129 controls, are shown in Fig. 4C. Trpc6−/−;Trpc3−/−double knock-out vessels had significantly impaired contraction amplitudes at several pressures and increased frequencies and increased levels of tone at most pressures. However, relevant to the main objective of this study, there were no significant impairments in pressure-induced chronotropy in vessels deficient in TRPC6, TRPC3 or both TRPC6 and TRPC3 (Fig. 4B-C).

Consequences of Trpc6 and/or Trpc3 deletion.

A) UMAP and bubble plots of Trpc6 and Trpc3 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from SV129 controls, Trpc6−/−, Trpc3−/− and Trpc6−/−;Trpc3−/− mice. C) Summary plots of contraction parameters as a function of pressure for the 4 respective groups of vessels. SV129 control N=8, n=11; Trpc6−/− N=5, n=10; Trpc3−/− N=7, n=11; Trpc6−/−;Trpc3−/− N=5, n=10.

Role of TRPM4

Multiple studies have implicated TRPM4 channels in the arterial myogenic response (Gonzales & Earley, 2013; Earley & Brayden, 2015), but to date no published studies have examined the role of TRPM4 in lymphatic vessel contractility. scRNAseq analysis revealed that Trpm4 was expressed at modest levels in only ∼10% of LMCs, and in a subset of LECs, but at higher levels in some AdvCs (Fig. 5A). We tested the role of TRPM4 in pressure-induced lymphatic chronotropy by comparing the responses of Myh11-CreERT2;Trpm4f/f vessels to Trpm4f/fvessels. The lack of significant reductions in F-P slope of Trpm4 smKO vessels (Fig. 5B) suggests that TRPM4 is not required for pressure-induced lymphatic chronotropy. Contraction data comparing Trpm4 smKO vessels to Trpm4f/f popliteal vessels are shown in Fig. 4C. There were no significant differences between the contraction parameters of Trpm4 smKO and Trpm4f/f vessels at any pressure.

Consequences of Trpm4 deletion.

A) UMAP and bubble plots of Trpm4 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Trpm4f/f and Trpm4 smKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. Trpm4 smKO N=6, n=13; Trpm4f/f N=4, n=7.

Role of PKD2

Studies by Jaggar and colleagues indicate that SM-specific deletion of PKD2 attenuates pressure-induced depolarization and myogenic tone in rat pial arteries (Narayanan et al., 2013) and blunts pressure-induced constriction in hindlimb arteries (Bulley et al., 2018). The roles of PKD2/2 channels in lymphatic vessel contractility are not known. scRNAseq analysis revealed modest Pkd2 and Pkd1 expression in LMCs and AdvCs (Fig. 6A). We tested the possible involvement of PKD2 channels in pressure-induced lymphatic chronotropy by comparing the responses of Myh11-CreERT2;Pkd2f/f(Pkd2 smKO) vessels to Pkd2f/f vessels (Fig. 6B). The lack of a significant reduction in the F-P slope for Pkd2 smKO vessels suggests that PKD2 is not required for pressure-induced lymphatic chronotropy. Contraction data comparing Pkd2 smKO vessels to Pkd2f/f popliteal vessels are shown in Fig. 6C. There were no significant differences between the contraction parameters of Pkd2 smKO and Pkd2f/f vessels at any pressure.

Consequences of Pkd2 (TRPP1) deletion.

UMAP and bubble plots of Pkd2 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Pkd2 f/f and Pkd2 smKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. Pkd2 smKO N=6, n=13; Pkd2 f/f N=4, n=8.

Role of TRPV4

In the blood vasculature, TPRV4 has been studied extensively in the context of Ca2+ influx pathways in endothelium (Chen & Sonkusare, 2020). However, TRPV4 is also expressed in VSM cells of some arteries (Mercado et al., 2014) and one report indicates that pharmacological inhibition of TRPV4 channels blunts pressure-induced VSM cell depolarization and myogenic vasoconstriction of preglomerular arteries from neonatal pigs (Soni et al., 2017). Our scRNAseq analysis revealed that TRPV4 was largely undetectable in LMCs but expressed at higher levels in almost all LECs, macrophages and some AdvCs (Fig. 7A). Nonetheless, we tested the role of TRPV4 channels in pressure-induced lymphatic chronotropy by comparing the responses of Myh11-CreERT2;Trpv4f/f vessels to Trpv4f/f vessels. The lack of a significant reduction in the F-P slope for Trpv4 smKO vessels suggests that TRPV4 is not required for pressure-induced lymphatic chronotropy (Fig. 7B). Contraction data comparing Trpv4 smKO and Trpv4f/f popliteal vessels are shown in Fig. 7C. Contraction amplitude was impaired at several pressures in Trpv4 smKO vessels, suggesting some involvement of TRPV4 channels in control of contractile strength, possibly as a result of macrophage-released products (Schulz et al., 2025). Frequency was significantly different at two lower pressures in Trpv4 smKO vessels (Fig. 7C), where it was elevated rather than reduced. We also tested the F-P relationship in Trpv4−/− (global KO) vessels and confirmed that there was no significant impairment in pressure-induced lymphatic chronotropy (average slope = 2.7 ± 0.2; n=18) compared to WT vessels (average slope = 2.8 ± 0.3; data not shown). Collectively, these results indicate that TRPV4 channels are not required for pressure-induced lymphatic chronotropy.

Consequences of Trpv4 deletion.

UMAP and bubble plots of Trpv4 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Trpv4f/f and Trpv4 smKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. TRPV4 smKO N=5, n=10; TRPV4f/f N=3, n=7.

Role of TRPV2

TRPV2 is an osmosensitive cation channel (Nilius & Droogmans, 2001). Down-regulation of TRPV2 expression in mouse aortic VSM cells reduced the amplitude of cation currents activated by hypotonic swelling (Muraki et al., 2003). In another study, application of a hypotonic bath solution stimulated TRPV2-like whole-cell cation currents and Ca2+ influx in VSMCs isolated from rat retinal arteries—responses that were attenuated by the TRPV2 inhibitor tranilast (McGahon et al., 2016). Additionally, the administration of tranilast to rat retinal arteries resulted in loss of pre-existing myogenic tone whereas pre-incubation with a TRPV2 blocking antibody prevented the development of myogenic tone (McGahon et al., 2016). Our scRNAseq analysis revealed that TRPV2 was highly expressed in 63% of Myh11-expressing LMCs (Fig. 8A). As we did not have access to TRPV2 null or floxed mice, we tested the potential role of TRPV2 channels in pressure-induced lymphatic chronotropy by bath application of tranilast (100 μM) to WT popliteal lymphatic vessels. There was no significant reduction in the F-P slope for tranilast-treated vessels (Fig. 8B). We also tested the effects of SET2, a recently reported TRPV2 inhibitor with a lower IC50 (460 nM) than tranilast. The application of SET2 (1 μM) caused a transient cessation of spontaneous contractions for 2-5 min in 5 of 6 vessels, but the contractions returned within 10 min in the continued presence of SET2. Subsequent pressure steps revealed that SET2 did not significantly alter the F-P slope (Fig. 8B). Contraction data comparing WT and tranilast-and SET2-treated WT popliteal vessels are shown in Figs. 8C. These results suggest that TRPV2 is not required for pressure-induced lymphatic chronotropy.

Consequences of Trpv2 inhibition.

UMAP and bubble plots of Trpv2 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from WT mice with/without treatment with the TRPV2 inhibitors tranilast (10 μM) and SET2 (1 μM). C) Summary plots of contraction parameters as a function of pressure for the three respective groups of vessels. WT+tranilast N=5, n=10; WT+SET2 N=3, n=6.

Role of ENaC

The epithelial Na+ channel (ENaC), comprised of a heterotrimer under the expression of the genes Scnn1a, Scnn1b, Scnn1g, Scnn1d, is expressed in both the endothelial and smooth muscle layers of arteries. αENaC is the principal isoform in endothelium and studies of native epithelial or heterologous channels reconstituted into bilayers suggest that αENaC is intrinsically mechanosensitive (Price et al., 2000; Benos, 2004). Work from the Drummond laboratory suggests that the βENaC isoform is expressed in VSM cells and plays a critically important role in myogenic constriction (Drummond, 2012). However, our scRNAseq analysis revealed almost no expression of αENaC subunits in LMCs and only weak expression in the LEC population (Fig. 9A); βENaC was essentially undetectable. Nevertheless, we tested the possible role of ENaC in pressure-induced lymphatic chronotropy by administering two widely used ENaC inhibitors, amiloride and benzamil, to WT popliteal lymphatic vessels. The data in Fig. 9B show that neither amiloride nor benzamil, at concentrations known to inhibit arterial myogenic constriction [5 and 1 μM, respectively (Jernigan & Drummond, 2006)], caused a significant impairment in pressure-induced lymphatic chronotropy. Contraction data for amiloride- and benzamil-treated WT popliteal vessels are shown in Fig. 9C. Both ENaC inhibitors tended to impair contraction amplitude to some degree (significant only for amiloride), with benzamil also significantly increasing tone at almost all pressure levels. However, the F-P relationships were nearly identical between WT and amiloride- or benzamil-treated vessels, reinforcing the conclusion that ENaC is not involved in pressure-induced lymphatic chronotropy.

Consequences of ENaC inhibition.

UMAP and bubble plots of ENaC subsunit expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from WT mice with/without treatment with the ENaC inhibitors amiloride (5 μM) and benzamil (1 μM). WT N=15, n=20, WT+amiloride N=5, n=9; WT+benzamil N=6, n=11.

Role of NCX1

The sodium-calcium exchanger (NCX) has a well-documented role in controlling the pacemaker activity of sinoatrial nodal cells, where it couples SR calcium oscillations, termed local calcium releases (LCRs) caused by calcium movement through ryanodine channels (Vinogradova et al., 2004; Vinogradova et al., 2005), to AP ignition. LCRs may occur throughout diastole (Monfredi et al., 2013), but their role is particularly critical in late diastole where LCRs activate NCX and the ensuing addition of inward current, and positive feedback of further calcium influx through L-type channels, provides sufficient depolarization to cross the voltage threshold for initiating AP firing (Bogdanov et al., 2001; Lyashkov et al., 2018b). This process is possibly analogous to the electrophysiological pacemaking mechanism in LMCs. scRNAseq analysis revealed strong expression of Slc8a, which encodes NCX1, in >80% of both LMCs and expression was also noted in LECs and some immune cells (Fig. 10A). We tested the role of NCX1 in pressure-induced chronotropy by comparing the responses of Myh11-CreERT2;Slc8a1f/f (Slc8a1 smKO) and Slc8a1f/f vessels. The lack of a significant reduction in the F-P slope for Slc8a1 smKO vessels suggests that NCX1 is not required for pressure-induced lymphatic chronotropy (Fig. 10B). Contraction data for Slc8a1 smKO and Slc8a1f/f popliteal vessels are shown in Fig. 10C and reveal no significant differences between the contraction parameters of Slc8a1 smKO and Slc8a1f/f vessels.

Consequences of Slc8a1 (NCX1) deletion.

UMAP and bubble plots of Slc8a1 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Slc8a1f/f and Slc8a1 smKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. Slc8a1f/f N=7, n=12; Slc8a1 smKO N=7, n=13.

Role of HCN

HCN (Hyperpolarization-activated cyclic-nucleotide-gated) channels contribute an inward “funny” current (If) to the diastolic pacemaking potential of sinoatrial node cells (Accili et al., 2002) and are also involved in the intrinsic rhythmicity of bladder and uterine smooth muscle (Alotaibi et al., 2017; Mader et al., 2018). In LMCs they could potentially contribute to diastolic depolarization. The HCN blockers CsCl, ZD7288 and ivabradine all lower the rate of spontaneous contractions of lymphatic vessels in rat diaphragm (Negrini et al., 2016), but inhibition requires substantially higher concentrations than the documented IC50 values for HCN channel blockade [see discussion in (Davis & Zawieja, 2024)]. In contrast, all three inhibitors increase the spontaneous contraction frequency of human lymphatic vessels at concentrations near or slightly higher than the IC50 values for HCN inhibition (Majgaard et al., 2022), suggesting off-target effects in some lymphatic vessels. Our scRNAseq analysis revealed essentially no expression of any of the four HCN isoforms in LMCs (Fig. 11A), however it is possible the expression is too low to detect without deep sequencing. Due to the proven role of HCN channels in cardiac pacemaking, we tested the role of HCN channels in pressure-induced lymphatic chronotropy of WT mouse popliteal lymphatics using the HCN inhibitor ivabradine (at a concentration of 3 μM, which is slightly higher than the IC50 of 2.5 μM), and the more effective inhibitor, zatebradine, at a concentration of 3 μM (IC50 = 480 nM). The lack of significant reductions in the F-P slope for ivabradine- or zatebradine-treated WT vessels suggests that HCN channels are not required for pressure-induced lymphatic chronotropy (Fig. 11B). Contraction data for ivabradine- and zatebradine-treated WT vessels are shown in Fig. 11C and reveal no substantial differences from WT vessels, except for a single pressure in which each HCN channel inhibitor actually increased the contraction frequency. Collectively, these results recapitulate the findings of Majgaard et al (Majgaard et al., 2022) and reinforce the conclusion that HCN channels are not required for pressure-induced lymphatic chronotropy.

Consequences of HCN inhibition.

UMAP and bubble plots of Hcn gene family expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from WT mice with/without treatment with the HCN inhibitors ivabradine (3 μM) and zatebradine (3 μM). WT N=15, n=20, WT+zatebradine. N=5, n=10; WT+ivabradine N=5, n=10.

Role of Kv7.4

Kv7 channels are activated at more negative membrane potentials than Kv1-family delayed rectifier K+ channels and thus are implicated in control of VSMC excitability near the resting membrane potential (Yeung & Greenwood, 2005; Zhong et al., 2010; Mani & Byron, 2011). The Kv7 blocker XE991 enhances myogenic tone in isolated mesenteric and renal arteries and hypotonic swelling suppresses an XE991-sensitive Kv current in patch-clamped VSM cells (Schleifenbaum et al., 2014). The collective implication from these studies is that mechanosensitive inhibition of Kv7.4 may increase VSM excitability. The effects of Kv7 channel inhibition on lymphatic function are unknown but the pan-Kv7 channel inhibitors linopirdine and XE991 cause concentration-dependent increases in lymphatic contraction frequency while the Kv7 activator ML-277 inhibits spontaneous lymphatic contractions (our unpublished observations in rat mesenteric lymphatic vessels). Our scRNAseq analysis revealed moderately strong expression of Kcnq4 in >50% LMCs and in some AdvCs (Fig. 12A). We tested whether pressure-induced lymphatic chronotropy was altered in Myh11-CreERT2;Kcnq4f/f (Kcnq4 smKO) mice. However, the lack of a significant reduction in the F-P slope for Kcnq4 smKO vessels suggests that Kv7.4 channels are not required for pressure-induced lymphatic chronotropy (Fig. 12B). Contraction data for Kcnq4 smKO and WT vessels are compared in Fig. 12C. There were no significant differences in any of the parameters at any pressure.

Consequences of Kcnq1 deletion.

UMAP and bubble plots of Kcnq4 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from WT and Kcnq4 smKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. Kcnq4 smKO N=5, n=7.

Role of Kir channels

Inwardly rectifying K+ channels, including Kir2.1 and Kir2.2 which are encoded by Kcnj2 and Kcnj12 respectively, have been implicated in shear-stress-induced responses of arterial endothelium (Ahn et al., 2017) and swelling-induced currents in arterial smooth muscle cells (Sancho et al., 2019). However, their potential roles in controlling LMC membrane potential and pressure-induced lymphatic chronotropy are unknown. Our scRNAseq analysis revealed modest expression of Kcnj8 (Kir6.1), an inward rectifier which forms the pore forming unit of the KATP channel, in >50% LMCs (Fig. 13A), but very little expression of Kcnj2 or Kcnj12 in LMCs. We previously showed that deletion of Kcnj8 does not alter the F-P relationship in popliteal lymphatics (Davis et al., 2020). Here, we tested the effects of BaCl2, a selective inhibitor of Kir2.1/2.2 at a concentration of 100 μM (Longden et al., 2017; Sancho et al., 2017; Sancho et al., 2019), on the F-P relationship of WT popliteal lymphatic vessels. Kir inhibition resulted in a significantly reduced slope of the F-P relationship compared to WT vessels (Fig. 13B), but the consistent effect of Ba2+ was to flatten the slope of the F-P relationship by substantially increasing the contraction frequency at low pressures while only slightly increasing the contraction frequency at elevated pressures (Fig. 13C). This contrasts with the effect of ANO1 inhibition or deletion: in the absence of ANO1 activity, the contraction frequency was significantly reduced at all pressures, particularly at low pressures (Fig. 2C). Thus, the inhibition of Kir channels does impair pressure-induced chronotropy but does so by differentially elevating contraction frequency as a function of pressure.

Consequences of Kir2.x inhibition.

UMAP and bubble plots of Kcnj2 (Kir2.1), Kcnj12 (Kir2.2) and Kcnj8 (Kir6.1) expression in various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from WT mice with/without treatment with the Kir2 inhibitor BaCl2 (100 μM). WT N=15, n=20, WT+BaCl2 N=4, n=11.

Roles of L- and T-type Ca2+ channels

The L-type, voltage-gated Ca2+ channel (Cav1.2) carries the majority of current contributing to the action potential in LMCs (van Helden, 1993; To et al., 2020). scRNAseq analysis confirmed that >90% of LMCs expressed message for the pore forming subunit of Cav1.2 (Cacna1c), along with other accessory subunits of the channel responsible for either gating characteristics (α2δ) or membrane targeting (β) (Fig. 14A). SM-specific knockout of Cacna1, using either the Myh11-CreERT2 or Itga8-CreERT2, eliminated essentially all propulsive contractions from popliteal lymphatic vessels (Warthi et al., 2022; Davis et al., 2023d). Residual activity in some vessels was limited to small, localized diameter oscillations (<5 μm in amplitude) with irregular frequencies that were largely independent of pressure. As expected, the slope of the F-P relationship for Cacna1c smKO vessels was significantly reduced compared to Cacna1cf/f control vessels (Fig. 14B).

Consequences of Cacna1c (Cav1.2) and Cacna1g/Cacna1h (Cav3.1/3.2) deletion.

UMAP and bubble plots of the expression of voltage-gated Ca2+ channels and their accessory subunits in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Cacan1cf/f, Cacna1c smKO and Cacna1g−/−;Cacna1h−/− mice. C) Summary plots of contraction parameters as a function of pressure for the three respective groups of vessels. Cacna1c: alpha-1c, Cacna2d1: α2/δ1, Cacnb2: β2, Cacnb3: β3, Cacna1d: alpha-1d (Cav1.3), Cacna1a: Cav2.1 (P/Q), Cacna1g: Cav3.1, Cacna1h: Cav3.2. Cacna1cf/f N=10, n=20; Cacna1c smKO N=14, n=24; Cacna1g−/−;Cacna1h−/− N=16, n=25.

The T-type Ca2+ channel (Cav3.x) is critical for pacemaking in SA node (Lyashkov et al., 2018a), GI smooth muscle (Sanders, 2019) and renal pelvis smooth muscle (Grainger et al., 2022). Two Cav3 isoforms are expressed in LMCs: Cacna1g and Cacna1h, encoding Cav3.1 and Cav3.2 respectively (Lee et al., 2014; To et al., 2020). One report suggests that Cav3 channels are selectively involved in regulating the contraction frequency (but not amplitude) of rat mesenteric lymphatic collectors (Lee et al., 2014). We previously tested the roles of Cav3.1 and Cav3.2 in mouse lymphatic collectors and isolated LMCs after generating Cacna1g−/−;Cacna1h−/− double KO mice (To et al., 2020). Here, we analyzed the slope of the F-P relationship in popliteal lymphatics from Cacna1g−/−;Cacna1h−/− mice but found no significant differences from either WT or Cav1.2f/f control vessels (Fig. 14B), indicating that Cav3 channels are not involved in determining the F-P relationship.

Upstream pathways mediating pressure-induced chronotropy

Having established that ANO1, out of all the cation and K+ channels tested, is the only current mediating pressure-induced lymphatic chronotropy, we then turned to an investigation of mechanosensitive pathways upstream from ANO1 channel activation.

Role of IP3R in pressure-induced chronotropy

Calcium release as a consequence of IP3 generation is known to activate a number of ion channels and other proteins, including ANO1. We recently showed that genetic deletion of Itpr1 (encoding IP3R1) from mouse IALVs results in a reduction in spontaneous contraction frequency and a blunting of pressure-induced chronotropy similar to what is observed after Ano1 deletion (Zawieja et al., 2023). scRNAseq analysis confirmed expression of Itpr1 in >50% of LMCs (Fig. 15A). Here, we examined the consequences of Itpr1 deletion in popliteal lymphatics by comparing the F-P relationship for vessels from Myh11-CreERT2;Itpr1f/f (Itpr1 smKO) and Itpr1f/fmice. SM-specific deletion of Itpr1 resulted in a significant reduction in the slope of the F-P relationship from 2.6 ± 0.3 to 0.6 ± 0.1 (Fig. 15B). Contraction data for Itpr1 smKO and Itpr1f/f control vessels are shown in Fig. 15C. The most striking features are 1) the significant reductions in frequency at all pressures and 2) a significant reduction in tone at elevated pressures in Itpr1 smKO vessels. The latter effect contrasts with the increases in tone observed after ANO1 inhibition or deletion (Fig. 2C). Collectively, these results show that IP3R1 is critically important for pressure-induced lymphatic chronotropy in popliteal collecting vessels in much the same way it was previously documented in IALVs. Deletion of Itpr1 in LMCs produces an effect on pressure-induced chronotropy similar to that of ANO1 deletion.

Consequences of Itpr1 (IP3R1) deletion.

UMAP and bubble plots of Ip3r1 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Itpr1f/f and Itpr1 smKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. Itpr1 smKO N=5, n=8; Itpr1f/f N=5, n=10.

Role of GNAQ/GNA11 in pressure-induced chronotropy

Evidence from multiple studies specifically points to GNAQ/GNA11-coupled GPCRs as the upstream mechanosensing elements of myogenic constriction in arteries (Mederos y Schnitzler et al., 2008; Blodow et al., 2014; Harraz et al., 2014; Li et al., 2014b; Schleifenbaum et al., 2014; Pires et al., 2017; Bjorling et al., 2018). Mechanosensitive activation of one or more GPCRs independent of ligand binding (Zou et al., 2004; Erdogmus et al., 2019) could account for pressure-induced increases in IP3 levels in LMCs. Our scRNAseq analysis revealed strong expression of Gnaq and Gna11 in ∼75% of LMCs, LECs, and AdvCs as well as very strong expression of Gnaq in some immune cell clusters (Fig. 16A). We tested the effects of SM-specific Gnaq knock out (Myh11-CreERT2;Gnaqf/f), global Gna11 knock out (Gna11−/−) and Gnaq/Gna11 double knock out (Myh11-CreERT2;Gnaqf/f;Gna11−/−) on pressure-induced lymphatic chronotropy. There was a trend for Gnaq smKO vessels to have a reduced slope in the F-P relationship in comparison to Gnaqf/f;Gna11+/+ control vessels, but it did not reach statistical significance (p=0.1296). However, global deletion of Gna11 significantly reduced the slope of the F-P relationship and the reduction was even more significant and pronounced in Gnaq/Gna11 double KO vessels (Fig. 16B). However, even combined knock out of both proteins did not cause an attenuation of pressure-induced chronotropy that was comparable to that observed after SM-specific Itpr1 or Ano1 deletion, particularly at pressures > 2 cmH2O (compare Fig. 16B to Figs. 2B, 15B). Contraction data for Gnaq smKO, Gna11−/− and Gnaq/Gna11double KO vessels, compared to tamoxifen-treated Gnaqf/f;Gna11+/+controls, are shown in Fig. 16C. The full F-P relationships for each genotype showed significant reductions in frequency at a few pressures for Gnaq smKO vessels compared to Gnaqf/f;Gna11+/+ controls, significant differences for Gna11−/− vessels at more pressures and significant differences at even more pressures for Gnaq/Gna11 DKO vessels. Collectively, these results suggest that both GNAQ and GNA11-coupled G-proteins are involved in pressure-induced lymphatic chronotropy, with G11 being more important. However, the escape from inhibition at higher pressures points to the possible involvement of other mechanisms activating phospholipase(s) to generate IP3.

Consequences of Gnaq and/or Gna11 deletion.

UMAP and bubble plots of Gnaq and Gna11 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Gnaqf/f;Gna11+/+, Myh11-CreERT2;Gnaqf/f (Gnaq smKO), Gna11−/− and Myh11-CreERT2;Gnaqf/f;Gna11−/− DKO mice. C) Summary plots of contraction parameters as a function of pressure for the four respective groups of vessels. Gnaqf/f;Gna11+/+ control N=3, n=6; Gnaq smKO N=4, n=8; Gna11−/− N=3, n=6; Myh11-CreERT2;Gnaqf/f;Gna11−/−N=8, n=13.

Role of GNA12/GNA13 in pressure-induced chronotropy

Chennupati et al. (Chennupati et al., 2019) made the surprising finding that GNA12/GNA13-coupled GPCRs, rather than GNAQ/GNA11-coupled GPCRs, mediated myogenic constrictions in mesenteric and cerebral arteries of mice. GNA12/GNA13-coupled GPCRs are known to regulate contraction through Rho kinase (McDuffie et al., 2024) and one study of rat mesenteric lymphatics suggests that RhoK is essential for phasic lymphatic contractions (Kurtz et al., 2014). We investigated the effects of deleting both Gna12 and Gna13 from LMCs (Myh11-CreT2;Gna12f/f;Gna13−/− mice) on the F-P relationship of popliteal lymphatics. In the Gna12/Gna13 double knock out, contraction amplitude was significantly blunted and tone was significantly elevated at several pressures, but the F-P relationship was nearly identical to that of vessels from control Gna12f/f;Gna13+/+ mice (Fig. 17B-C).

Consequences of Gna12/Gna13 deletion.

UMAP and bubble plots of Gna12 and Gna13 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Gna12f/f;Gna13+/+ and Myh11-CreERT2;Gna12f/f;Gna13−/− DKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. Gna12f/f;Gna13+/+ controls N=5, n=10; Myh11-CreERT2;Gna12f/f;Gna13−/−N=7, N=15.

Role of AT1R in pressure-induced chronotropy

In many arteries, AT1R is one of the predominantly expressed GPCRs and deletion of Agtr1 results in loss or impairment of myogenic constriction in several types of arteries (Mederos y Schnitzler et al., 2008; Storch et al., 2012; Blodow et al., 2014; Schleifenbaum et al., 2014; Storch et al., 2015; Pires et al., 2017; Yamasaki et al., 2020). Selective deletion of the AT1R isoform (AT1aR or AT1bR) is sufficient to abrogate myogenic constriction in some arteries (Schleifenbaum et al., 2014; Pires et al., 2017). However, our scRNAseq analysis revealed almost no Agtr1a expression in LMCs, except in one possible subcluster (Fig. 18A). Pressure-induced chronotropy was measured in popliteal vessels from Agtr1a−/− mice and in vessels from WT mice treated with the AT1R inhibitor losartan, which blocks both AT1aR and AT1bR. We used a concentration of losartan (10 μM) known to be sufficient to block myogenic constriction in several types of arteries (Blodow et al., 2014; Jackson & Boerman, 2017; Yamasaki et al., 2020). However, neither inhibition of AT1R nor deletion of Agtr1a impaired pressure-induced chronotropy, as evident by the absence of significant reductions in slope of the F-P relationship (Fig. 18B). Contraction data for these three groups are shown in Fig. 18C. Losartan treatment caused a slight impairment of contraction amplitude at several pressures but neither AT1R blockade nor Agtr1a deletion significantly inhibited the F-P relationship, suggesting that AT1R is not required for pressure-induced lymphatic chronotropy.

Consequences of Agtr1 (AT1R) deletion/inhibition.

UMAP and bubble plots of Agtr1a expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from WT mice with/without treatment with the AT1R inhibitor losartan (10 μM) and Agtr1a−/− vessels. WT+losartan N=5, n=10; Agtr1a−/− N=3, n=6.

Discussion

Summary

The major goal of this study was to investigate the mechanistic basis of the F-P relationship in lymphatic muscle. The primary means by which lymphatic pumping is matched to excess filling of the interstitium is pressure-induced chronotropy—a >10-fold or more increase in contraction frequency with over the physiological pressure range (Davis et al., 2025). Because little was previously known regarding the mechanisms of mechanosensitivity underlying pressure-induced lymphatic chronotropy, we took cues from a well-studied mechanosensitive mechanism in vascular smooth muscle—the arterial myogenic response. In both arteries and collecting lymphatic vessels (Davis et al., 2009; Davis et al., 2023c), pressure elevation leads to vascular muscle cell depolarization and constriction, reflecting the activation of a depolarizing current that triggers Ca2+ entry through the opening of L-type, voltage-gated Ca2+ channels. Several ionic mechanisms have been previously implicated in this process for arterial smooth muscle, specifically the activation of TRPC6, TRPM4, PKD2, TRPV2 and ENaC cation channels (Welsh et al., 2002; Earley et al., 2004a; Jernigan & Drummond, 2005; Sharif-Naeini et al., 2009; McGahon et al., 2016), with supporting evidence for each obtained from gene knockdown / knockout studies. Additional ion channels may also be involved, including the activation of channels promoting Cl− efflux (Boedtkjer et al., 2008) and the inhibition of constitutively active K+ channels (Sancho et al., 2019), but evidence for those is based largely on the use of pharmacological inhibitors and therefore subject to concerns about off-target effects (Maroto et al., 2005; Dietrich et al., 2007; Gottlieb et al., 2008; Sancho et al., 2019). In the present study, we primarily used genetic knock out approaches to test the possible roles of the major ion channels contributing to the mechanosensitive chronotropy of lymphatic vessels from the mouse. Our results strongly support the conclusion that pressure-induced lymphatic chronotropy does not involve acute activation of any of the putative mechanosensitive (and/or second-messenger gated) cation channels previously implicated in myogenic constriction (TRPC6, TRPM4, PKD2, TRPV2, ENaC), nor does it require PIEZO1, the bona fide mechanosensitive cation channel implicated in other aspects of mechanotransduction in the cardiovascular system [see references in (Davis et al., 2023c)]. Rather, our results align with evidence from related studies of arterial smooth muscle showing that mechanotransduction of stretch/pressure is mediated by GNAQ/GNA11-coupled GPCRs, leading to downstream activation of second messenger-gated channels. In the case of LMCs, mechano-activation of GNAQ/GNA11-coupled GPCRs promotes IP3 production that triggers Ca2+ release from stores through IP3R1; Ca2+ then activates ANO1 Cl− channels, which accelerate the rate of spontaneous diastolic depolarization, increasing the firing rate of action potentials that control phasic lymphatic contractions (Fig. 19). Repolarization is controlled in part by Kv channels (Cotton et al., 1997), including Kv11 channels (Kim et al., 2023). As summarized in Table 1, knockout of each element in the GNAQ/GNA11-IP3-IP3R1/Ca2+-ANO1-Cav1.2 pathway (and only those elements out of all that were tested, except Kcnj2/ Kcnj12, which can be alternatively explained) abrogated or significantly attenuated pressure-induced chronotropy. Mechanosensitive regulation of ANO1 in this context is likely mediated by Ca2+ release from stores rather than possible direct (Ca2+-independent) mechanosensitivity of ANO1 (Zou et al., 2025), because the similar degree of attenuation in the F-P relationship with Iptr1 knockout as with Ano1 smKO knock out shown here points to the involvement of an upstream mechanosensor.

Proposed mechanotransduction mechanism underlying pressure-induced chronotropy.

Summary diagram depicting the primary signaling pathway through which changes in pressure regulate the frequency of the ionic pacemaker. GNAQ/GNA11-coupled GPCRs are the primary, but not exclusive mechanosensor. ANO1 is the primary, but not exclusive effector. Kv11 provides at least one source of repolarizing signal (Kim et al., 2023).

Possible involvement of other G-coupled proteins

One inconsistency with the scheme suggested in Table 1 is that combined knock out of Gnaq/Gna11 did not cause an attenuation of pressure-induced chronotropy comparable to that observed after SM-specific Itpr1 or Ano1 deletion, particularly at pressures > 2 cmH2O (compare Fig. 16B to Figs. 2B, 15B). A possible explanation is that other G-coupled proteins are involved in mechanosensitive regulation of the GPCR-IP3R1-Ca2+-ANO1 signaling in LMCs. Indeed, two other G protein-coupled effectors, Gna14 and Gna15 in the GNAQ family, could also be activated downstream of pressure-induced GPCR activation and potentially mediate IP3/DAG production. We have included Gna14 (Gna14 is expressed in 28% of LMCs in our dataset) in the summary scheme shown in Fig. 19 but its role in IP3 production and ANO1 activation remains to be investigated. Even so, inclusion of other G protein-coupled effectors would still fail to account for the residual pacemaking observed after apparently complete deletion of ANO1. We use the phrase “apparently complete” because incomplete recombination by the inducible Myh11-CreERT2 remains a possibility. Although we could not detect residual message for Ano1 in Myh11-CreERT2;Ano1f/f mice after tamoxifen treatment (Fig. 3B), that message was measured in LMCs digested from IALVs inguinal-axillary lymphatic vessels (chosen because of a higher cell yield), whereas the F-P relationship was measured in popliteal lymphatic vessels (chosen because of the greater dynamic range in F-P).

We also investigated the possible role of GNA12/GNA13-coupled GPCRs in the F-P relationship, based on a study showing that GNA12/GNA13, rather than GNAQ/GNA11, is required for myogenic constriction in arterial smooth muscle (Chennupati et al., 2019) and a paper showing that RhoK (known to be activated downstream from GNA12/GNA13 activation) is critical for phasic lymphatic contractions and associated phasic LMC Ca2+ signaling (Kurtz et al., 2014). However, Gna12/Gna13 double knock out resulted in no significant change in the F-P slope (Fig. 17B), and there were no differences in frequency at any pressure compared to Gna12f/f; Gna13+/+ control vessels (Fig. 17C).

Roles for ion channels in other aspects of LMC contractile function

Originally, we hypothesized that an ion channel (other than ANO1) was involved in LMC pacemaking. If IP3 production in LMCs is enhanced after pressure elevation, as it is in arterial smooth muscle (Narayanan et al., 1994; Mederos y Schnitzler et al., 2008), then concomitant increases in DAG, the other product of PIP2 hydrolysis, would also be predicted to occur, and DAG should then activate TRP6C and related channels. However, if this occurs in popliteal lymphatic vessels, the activation of those channels apparently does not significantly influence the F-P relationship (Fig. 4). Likewise, the Ca2+ increase resulting from mechanosensitive activation of the GNAQ/GNA11-IP3-IP3R1 signaling axis would potentially activate TRPM4 and/or PKD2 (assuming those channels are localized to the appropriate LMC microdomains), but we find that their deletion does not alter the F-P relationship (Figs. 5-6). Because TRPM4, PKD2 and PIEZO1 expression levels are very low in LMCs (Figs. 3A,4A, 5A), further activation of those channels may have little effect. It is possible that another unidentified ion channel could be mediating the pressure-driven frequency changes at higher pressures—perhaps a channel with a higher mechanosensitive threshold or requiring higher levels of IP3, DAG or Ca2+. Candidates include TRPV1, which has also been implicated in arterial myogenic constriction (Phan et al., 2022).

Although the deletion/inhibition of several TRP channel family members and other channels did not significantly impact pressure-induced chronotropy, it did impact other aspects of lymphatic contractile function, namely contraction amplitude and/or tone. The respective channels and their statistically significant contraction effects are summarized in Table 2. Deletion of TRPC6/TRPC3 impaired contraction amplitude by ∼40% and increased tone by ∼2.5-fold (Fig. 4C). Deletion of TRPV4 from LMCs impaired contraction amplitude by ∼45% without an effect on tone (Fig. 7C). TRPV4 channel activation has recently been implicated in the regulation of lymphatic contraction, but primarily via the production of vasoactive products from LECs and resident macrophages (Schulz et al., 2025). There were three other channels whose inhibition led to significant changes in amplitude and/or tone. Specifically, one of the two TRPV2 inhibitors (SET2) inhibited tone at 4 of the 5 lowest pressures. One of the two ENaC inhibitors (amiloride) inhibited amplitude by 20-30% while the other one (benzamil) enhanced tone by ∼2-fold (Fig. 9C). The Kir2 blocker Ba2+ impaired amplitude by ∼40% and increased tone by 2-fold (Fig. 13C). Collectively, these effects suggest that the respective channels are each involved in some aspect of lymphatic contractile function (other than pressure-induced chronotropy), with the more compelling evidence being that obtained after by gene deletion rather than after acute application of soluble inhibitors, which may have off-target effects.

Future directions

Lastly, the critical mechanosensitive GNAQ/GNA11-coupled GPCR(s) in LMCs remain(s) to be identified. Although evidence points to AT1R as the key GNAQ/GNA11 –coupled mechanosensitive GPCR in arterial smooth muscle (Mederos y Schnitzler et al., 2011; Davis et al., 2023c), message for AT1R is negligible in LMCs (Fig. 18A) and neither Agtr1 knock out or pharmacologic AT1R blockade altered pressure-induced chronotropy (Fig. 18B-C). No studies to date have attempted to identify the major GPCRs expressed in LMCs, e.g., by analyzing the various published scRNAseq databases for LMCs (Kenney et al., 2022; Lei et al., 2025; Schulz et al., 2025; Zawieja et al., 2025), but doing so might provide a starting point for identifying the specific mechanosensitive GNAQ/GNA11 - coupled (and other) GPCRs that mediate pressure-induced chronotropy.

Physiological Relevance

LMC function is impaired in Cantú syndrome, the only known example of primary lymphedema produced by a gene mutation (Kcnj8/Abbc9) in LMCs (Davis et al., 2023b). This impairment is evident both in lower contraction frequencies and amplitudes at multiple pressure levels. The lower frequency is consistent with parallel observations that gain-of-function mutations in KATP channels increase metabolic stress (Kim et al., 2024), whereas the lower amplitude suggests that other aspects of LMC function are altered as a consequence of chronic KATP channel hyperactivation. It is interesting that lymphatic contraction frequency and amplitude are also impaired in secondary lymphedema, in aging and in various models of metabolic disease [(Castorena-Gonzalez et al., 2025) and see discussion and references in (Davis et al., 2023e)]. Potential therapies for reversing LMC dysfunction under these conditions will depend on identifying the key ion channels and signaling pathways that drive pacemaking and control contraction amplitude. The present study represents a major step forward in the identification of specific LMC targets that could be used for that purpose.

eLife Assessment

Davis and colleagues describe findings that are fundamental to the understanding of pressure mechanosensation in lymphatic vessels and are of significant importance to other areas of mechanosensory physiology. Based on many different knockout mouse models and rigorous state-of-the-art pressure myography recordings, they present compelling evidence that mechano-activation of GNAQ/GNA11-coupled GPCRs generates IP3, which induces Ca2+ release from internal stores through IP3R1 and drives depolarization through the activation of ANO1 Cl- channels to induce lymphatic vessel contractility. Nevertheless, some aspects of the manuscript are incomplete. The specific identity of the GPCR(s) involved remains to be uncovered, as evidence of frequency-pressure impairment is only demonstrated with abolition of GNAQ/GNA11action, not the receptors per se.

Reviewer #1 (Public review):

Summary:

Davis and co-authors used many mouse models to investigate mechanisms that regulate the contractility of mouse popliteal collecting vessels, primarily chronotropy. Many of the mechanisms studied were previously shown to regulate pressure-induced constriction in small arteries. The authors use prior literature from the vasculature as a framework to test similar concepts in lymphatic vessels. The mouse models used provide evidence for and against the involvement of multiple proteins in regulating chronotropy and other contractile properties in lymphatic vessels. They propose that mechano-activation of GNAQ/GNA11-coupled GPCRs generates IP3, which induces Ca2+ release through IP3R1 and drives depolarization through the activation of ANO1 Cl- channels. Major concerns include the author's major conclusion that GNAQ/GNA11-coupled GPCRs contribute to chronotropy. This conclusion is not supported by the data presented.

Strengths:

One major strength of the study lies in the vast number of mouse knockout models that were used to test the importance of ion channels and G protein signaling pathways in the regulation of lymphatic vessel contractility. In this regard, the study is a valiant effort. The authors achieved several objectives to find that ANO1 and IP3R1 regulate chronotropy, and many other potential proteins do not regulate chronotropy. This study will have a major impact on the field if additional support for G proteins is provided.

Weaknesses:

Major conclusions concerning the involvement of G proteins are drawn from the global Gna11 knockout mouse models. This conclusion is weak. Global Gna11 knockout mice are highly likely to have a multifactorial phenotype that could create significant differences in the data. Control experiments need to be performed on vessels from the global knockout mice if these major conclusions are to be made. Similarly, pharmacological tools or alternative approaches to manipulate G proteins should be used to support the data from these mouse models to draw these major conclusions.

The Gnaq smKO mice are the most specific G protein model studied here. However, there is no phenotype. Do not discuss trends in the data. If the data are not significant, conclude so. If more experiments are required to reach significance, provide more data in the manuscript.

The conclusions repeatedly refer to a signaling pathway wherein the upstream component is GPCRs, which activate G proteins. While this may be the case, no GPCRs were identified here, and the involvement of G proteins is questionable, as the authors outline in lines 693-695 and noted above. The conclusions should be tempered, including in the abstract, unless additional experiments are performed to support the involvement of G proteins. Perhaps then the authors may be able to infer that GPCRs are involved.

Line 318. The point regarding the choice to use popliteal vessels versus IALVs will be unclear to the uninitiated, particularly as the authors previously used IALVs. Including additional justification in the text and/or data from IALVs in Figure 1, which compares IALVs to popliteal vessels, would better explain the logic.

The conclusions drawn for TRPC6 and TRPC3 are less convincing. Germline global knockout mice, which are known to undergo compensation, were used, and high data variability is apparent. Using TRPC3 and TRPC6 blockers in the mouse models studied in Figure 4 would strengthen the arguments made regarding these proteins.

Did you perform power analysis to ensure that experimental numbers were sufficient to conclude that no statistical difference exists between datasets? If not, this needs to be done. For example, data shown in Figure 5C for tone and 6C for frequency and tone appear to be significantly different, but are concluded not to be so.

At the end of each result section, a concluding statement is made regarding the effects on pressure-induced chronotrophy. In many cases, there are additional effects of manipulating protein expression on other contractile properties. One example is for TRPC3 and TRPC6 (lines 414-416), but others are TRPV4, TRPV3, ENaC, Kir, Cav3.1/3.2, etc. Some interpretation is in the Discussion, but the concluding statements at the end of each result section should be expanded to summarize what the authors think the other significant differences in the data represent.

Kv7.4 channels. You state you have data (not shown) with linopiridine and XE991. Why not show those results here to support the experiments with the Kcnq4 smKO mice? Otherwise, I suggest you remove the statement from the unpublished data.

Figure 13A. Kcnj2 is modestly expressed in LECs, but very little is present in LMCs. This likely underlies the effect of barium. If you remove the endothelium, does the effect of barium disappear? While this is not the major focus of the study, the effects of barium are dramatic, and it should be made clear whether this is due to inhibition of Kir channels in smooth muscle or endothelial cells.

Figure 18C tone. Several values for losartan look different but are not labelled as such. Please clarify and discuss if different.

The manuscript should include raw data traces in figures that show the major pathways that you conclude regulate chronotropy.

Reviewer #2 (Public review):

Summary:

In this study, Davis et al. embarked on the quest for the molecular elements responsible for the regulation of lymphatic phasic contractile activity in response to variation of transmural pressure, a mechanism (termed pressure-induced lymphatic chronotropy by the authors) critical for drainage of interstitial fluid from the tissue and transport of lymph back to the blood circulation. Their aim was to investigate the mechanism(s) involved in the pressure-induced regulation of lymphatic pumping, and test whether activation of cation channels, shown in other systems to play mechanosensitive roles are directly at play, and/or whether mechano-activation of GNAQ/GNA11-coupled GPCRs is necessary to generate second messengers to activate those channels, as it has been suggested for the regulation of myogenic tone in arteries. To achieve their goal, the authors used their well-described, highly reliable protocols of mouse lymphatic vessel isolation, pressure myography, and data acquisition to obtain frequency-pressure relationships and other contractile function parameters from transgenic mice where specific channels or molecular elements of interest have been ablated. They combined these data with scRNAseq analysis of these gene targets to determine their respective role and levels of expression in lymphatic muscle cells. Their conclusion is that none of the exhaustive list of tested ion channels was critical, except ANO1 Cl channels, part of the contractile pacemaker mechanism, but that transmural pressure activates GNAQ/GNA11-coupled GPCRs, which generate IP3 to induce SR Ca2+ release through IP3R1 and activate ANO1-mediated depolarization.

Strengths:

The manuscript's strengths reside primarily in very robust, clean, and unequivocal pressure myography data and analysis. The research team is mastering these techniques they developed more than a decade ago and have implemented in mouse lymphatics to study their contractile properties, with consistent and convincing outcomes. They also provide data from an impressive list of transgenic mice in order to determine the role of the targeted gene in pressure-induced lymphatic chronotropy, relying on pharmacological small molecule inhibitors only when necessary. Finally, the use of scRNAseq analysis they gathered from previously published datasets brings novelty with respect to the expression of the genes of interest in all populations of cells comprising the lymphatic vessels, but more critically, to validate or contrast the potential impact of genetic alteration of the given gene on the ability of lymphatic muscles to respond to a change in pressure.

Weaknesses:

The main weakness may reside in the fact that while the authors provide a convincing demonstration that GNAQ/GNA11 are involved in the regulation of the F-P relationship, they give little evidence of the involvement of "upstream" receptors. Indeed, inhibition of AT1R, shown to be involved in myogenic regulation of arteries (a phenomenon the authors rightfully compare to pressure-induced lymphatic chronotropy), didn't lead toa similar effect (decrease in F-P) in lymphatic vessels. Arguably, other GPCRs might be involved in lymphatic vessels, but as such information is not provided in the manuscript, the author's conclusions should be dampened. More in-depth discussion would be required. In fact, it can be argued that the discussion is very restricted with respect to the amount of data and information the manuscript provides.

Overall, the authors convincingly achieved their aim by performing an impressive number of technically challenging experiments, leading to solid datasets. While these support their main conclusions, a more elaborate discussion might be required to refine them.

This study is likely to have an important impact on the field as it provides some answers to the lingering question of how lymphatic vessels regulate their contractile activity to variation in transmural pressure and certainly proposes an experimental means to further explore and address that question.

Reviewer #3 (Public review):

In this manuscript, Davis and colleagues aimed to identify the molecular sensors and signaling cascade that enable collecting lymphatic vessels to increase their spontaneous contraction frequency in response to intraluminal pressure (pressure-induced chronotropy). They tested whether the process is similar to blood vessel myogenic constriction by relying on cation channels (TRPC6, TRPM4, PKD2, PIEZO1, etc.) or instead require the activation of G-protein-coupled receptors (presumably mechanosensitive GNAQ/GNA11-coupled receptors), using ex vivo pressure myography of mouse popliteal lymphatics, smooth muscle-specific conditional knockouts, quantitative PCR validation, and single-cell RNA sequencing for target prioritization. The authors convincingly demonstrate that pressure-induced chronotropy does not require the cation channels implicated in arterial myogenic tone but is blunted by deletion of GNAQ/GNA11 or IP3 receptor 1, supporting a model of GPCR > IP3 > Ca2+ release > Cl⁻ channel activation > depolarization. The core conclusion is robust. The work redefines lymphatic pacemaking as G-protein-coupled receptor-dependent mechanotransduction, distinct from arterial mechanisms, and provides a genetically validated toolkit that is useful for studying lymphatic function and dysfunction.

Strengths:

(1) The data are of high quality and highly sensitive functional readouts

(2) The systematic genetic targeting is a major strength that overcomes pharmacological artifacts

(3) Careful quantitative analyses of frequency-pressure slopes

Weaknesses:

(1) The use of inguinal-axillary vessels for single-cell RNA sequencing rather than the popliteal segment studied functionally.

(2) No direct testing of the specific G-protein-coupled receptor involved.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

(1) Davis and co-authors used many mouse models to investigate mechanisms that regulate the contractility of mouse popliteal collecting vessels, primarily chronotropy. Many of the mechanisms studied were previously shown to regulate pressure-induced constriction in small arteries. The authors use prior literature from the vasculature as a framework to test similar concepts in lymphatic vessels. The mouse models used provide evidence for and against the involvement of multiple proteins in regulating chronotropy and other contractile properties in lymphatic vessels. They propose that mechano-activation of GNAQ/GNA11-coupled GPCRs generates IP3, which induces Ca2+ release through IP3R1 and drives depolarization through the activation of ANO1 Cl- channels. Major concerns include the author's major conclusion that GNAQ/GNA11-coupled GPCRs contribute to chronotropy. This conclusion is not supported by the data presented.

Although we have not yet identified a specific GPCR in LMCs that mediates pressure-induced chronotropy, our experiments have established that Gq/11-coupled GPCRs are critical for this response. The almost identical phenotypes of GNAQ/GNA11 double KO vessels with ANO1 KO and IP3R1 KO vessels point to an upstream, mechanosensitive GPCR that regulates ANO1-regulated pacemaking activity through calcium release by IP3R1. New experiments show that the highly specific GNAQ/GNA11 inhibitor YM254890 decreases contraction frequency to the same degree as, or even stronger than, that observed in Gq/11 double KO mice (Fig. 22). Additional data assessing the responses of inguinal-axillary lymphatic vessels (IALVs) from G11-/- and Gq/11 double KO further confirm our findings and conclusions (Suppl. Fig. 1). To address your concern about identifying a specific GPCR, we have performed new scRNAseq analyses to identify and rank the most prominently expressed GPCRs in LMCs (Fig. 21) along with new experiments to test the most likely GPCR candidates that emerged from that analysis (Fig. 22). Please see our specific responses below and the additional data that we now provide.

While our results still fall short of identifying a specific Gq/11-coupled GPCR involved in this LMC mechanosensitive response, they rule out all of the major mechanosensitive ion channels previously implicated in this process (Figs. 4-15), rule out G12/13-coupled GPCRs (Fig.20) and strongly support a critical role for the canonical Gq/11-coupled mechanism (Fig. 18 and Suppl. Fig. 1). New data to eliminate contributions of the 7 most consistently expressed LMC GPCRs, along with the most likely mechanosensory GPCR in VSMCs (AT1R) (Fig. 22), out of the 136 GPCRs detected in our LMC scRNAseq dataset, raise interesting GPCR candidates that require future testing. The remaining list of 11 strongly or moderately expressed candidates includes 3 frizzled GPCRs that typically do not signal through G proteins, an olfactory GPCR, and several adhesion GPCRs (aGPCRs) that encode their own agonist (containing the Stachel peptide sequence). Some of these adhesion GPCRs recruit and signal through Gq/11, and aGPCRs have been implicated in mechanosensation in other cell types (PMID: 25937282, PMID: 40395494). However, pharmacological tools to specifically test most of these aGPCRs are currently lacking. Another possibility, of course, is that the key LMC protein may be a GPCR with a very low expression level.

At present, the only way to test most of those candidates would be to generate new mice with smooth muscle KO of the specific GPCRs, as Offermann’s group has done in a few other cases (PMID: 40623996, PMID: 38597112, PMID: 39920161), but we are without the resources to embark on such an expedition. We are therefore unable to resolve this last issue with currently available resources and/or technology but now acknowledge this as a shortcoming of our study and a future direction. We have adjusted our title to be more congruent with the data presented in the revised manuscript.

Strengths:

(2) One major strength of the study lies in the vast number of mouse knockout models that were used to test the importance of ion channels and G protein signaling pathways in the regulation of lymphatic vessel contractility. In this regard, the study is a valiant effort. The authors achieved several objectives to find that ANO1 and IP3R1 regulate chronotropy, and many other potential proteins do not regulate chronotropy. This study will have a major impact on the field if additional support for G proteins is provided.

Please see our specific responses below and the additional data that we now provide to strengthen support for a GNAQ/GNA11 mechanism.

Weaknesses:

(3) Major conclusions concerning the involvement of G proteins are drawn from the global Gna11 knockout mouse models. This conclusion is weak. Global Gna11 knockout mice are highly likely to have a multifactorial phenotype that could create significant differences in the data.

We agree that global GNAQ/11 DKO mice likely have a multifactorial phenotype and that, IF our hypothesis involved testing some aspect of in vivo function with multiple cell types or organs, this would complicate our conclusions. However, because our approach focuses on ex vivo testing of the F-P relationship of isolated lymphatic vessels, a response mediated specifically by LMCs (i.e., independent of other circulating factors, endothelium, neural control, etc.), our results are NOT likely to be influenced by such extrinsic factors. Stefan Offermanns and colleagues have used similar mice (single floxed gene + Cre + global KO) and ex vivo assays of arteries in multiple peer-reviewed studies (e.g. PMID 31549965), so there is clear precedent for our approach. An alternative approach to the global GNAQ/11 DKO mice would be to use Myh11Cre; GNAQf/f; GNA11f/f double floxed mice, but then those would likely suffer from the potential limitation of competing recombination between the two floxed genes, as we and others have encountered this problem in previous studies using a single Cre with multiple floxed alleles (PMID: 29336844, PMID: 36868507).

(4) Control experiments need to be performed on vessels from the global knockout mice if these major conclusions are to be made.

Control experiments were conducted on popliteal vessels from Myh11Cre; GNAQf/f; GNA11+/+ mice (uninduced littermates) and from GNAQf/f; GNA11+/+ mice (lacking Myh11Cre), the combination of which had a normal F-P relationship (Fig. 18). We have also conducted experiments on Myh11Cre; GNAQf/f; GNA11-/- mice that were not induced with tamoxifen (GNA11-/-). It is not clear if the reviewer is suggesting that we include GNAQ-/- mice, but we did not have access to those mice. Gnaq-/- mice exhibit significant developmental abnormalities (PMID: 22723772), although according to Offermanns and colleagues (PMID: 9687499), only GNAQ homozygous deficient mice displayed “obvious phenotypic defects”.

(5) Similarly, pharmacological tools or alternative approaches to manipulate G proteins should be used to support the data from these mouse models to draw these major conclusions.

To address this comment, we have performed new experiments on WT mice using the potent GNAQ/11 inhibitor YM254890 (at 100 nM, the concentration used in many previous publications by other groups), which had nearly the same effect on abrogating the F-P relationship as induced Myh11Cre; GNAQf/f; GNA11-/- mice (see Fig. 22). The specificity of YM-254890 for the Gq family has recently been verified by excellent work out of David Yule’s lab (PMID: 40563241).

(6) The Gnaq smKO mice are the most specific G protein model studied here. However, there is no phenotype. Do not discuss trends in the data. If the data are not significant, conclude so. If more experiments are required to reach significance, provide more data in the manuscript.

With respect to the Gq smKO data:

(1) We have added more data to the control group.

(2) We have now listed the p value (0.0819 in Fig. 18) for the Gq/11 controls vs Gq smKO, instead of simply designating it “n.s.” with a cut-off of p<0.05; the readers can now judge for themselves how close the values come to being significant.

(3) Although the F-P slope does not reach significance (Fig. 18B), the frequency is significantly different between the Gq/11 controls and Gq smKO at 3 pressures (see the frequency graph in Fig. 18C), so significant differences ARE indeed detected in Gq smKO vessels.

(4) The F-P slope for the DKO is lower than that of the G11 KO and it appears the two knockouts are cumulative, suggesting that Gnaq deletion is contributing to the lower values of the Gq/11 DKO. As further support for this conclusion, we have performed new experiments on IALVs (in part also to address concern #8 below) and that analysis is shown in Suppl. Fig. 1. We include those data as a Supplemental Figure because the contraction data in all the other figures are for popliteal lymphatic vessels. Note that the F-P analysis is not appropriate for IALVs because the control vessels typically show only a 2-2.5x change in frequency over the entire pressure range and frequency essentially reaches a plateau at P = 2 cm H2O. The frequency plots for control, G11-/- and Gq/11 DKO IALV vessels show the same trend as for popliteal vessels (significantly lower than control vessels at all pressures, with only a single exception), but in this case the DKO vessels have significantly lower frequencies than G11-/- vessels at the lower pressures (1-2-3 cm H2O), further supporting the results from popliteal vessels.

(7) The conclusions repeatedly refer to a signaling pathway wherein the upstream component is GPCRs, which activate G proteins. While this may be the case, no GPCRs were identified here, and the involvement of G proteins is questionable, as the authors outline in lines 693-695 and noted above. The conclusions should be tempered, including in the abstract, unless additional experiments are performed to support the involvement of G proteins. Perhaps then the authors may be able to infer that GPCRs are involved.

Please see our response to your comment #1. We have now included additional experiments using the Gq/11 inhibitor YM254890 (at a concentration of 100 nM that is reputed to be “selective” for Gq/11, according to the literature (PMID: 40563241). The blunting of the F-P relationship in the presence of that compound is comparable to, and perhaps even more significant than, the blunting observed for the DKO (the summary analysis is shown for P= 3 cm H2O in the middle of Fig. 22, in a format that matches the other analyses in that figure, but the frequencies were significantly attenuated over the entire lower pressure range). These results strengthen support for a ligand-independent mechanotransduction mechanism through one or more GPCRs of the Gq/11 family. However, because 1) there are at least 136 possible specific GPCR candidates and 2) our new specific tests of the 7 most abundantly expressed GPCRs were negative, pinning down the exact GPCR(s) involved will require much more work. Given the amount of data already presented in this study, and the possibility that the unidentified GPCR is an orphan receptor or other GPCR with very low expression, we think it is appropriate to defer that line of experimentation to a future study. We now clearly state that this is a limitation of the present work and a future direction.

(8) Line 318. The point regarding the choice to use popliteal vessels versus IALVs will be unclear to the uninitiated, particularly as the authors previously used IALVs. Including additional justification in the text and/or data from IALVs in Figure 1, which compares IALVs to popliteal vessels, would better explain the logic.

We now include additional justification in the text by citing Zawieja 2018 and the similarities to Ano1 KO and IP3R1 KO findings in IALVs in our other papers. Note also the very close agreement between the Gq/11 data for popliteal vessels vs IALVs in Fig. 18 C and Suppl. Fig. 1).

(9) The conclusions drawn for TRPC6 and TRPC3 are less convincing. Germline global knockout mice, which are known to undergo compensation, were used, and high data variability is apparent. Using TRPC3 and TRPC6 blockers in the mouse models studied in Figure 4 would strengthen the arguments made regarding these proteins.

Our primary reason for using the global KO mice is that pharmacological inhibitors of TRP channels, particularly TRPC6, are quite non-specific. We did not have access to Trpc6 floxed mice but our ex vivo assay focuses a specific aspect of LMC function in isolation from extrinsic influences (see answer to comment #3 above), and of the 18 cell subtypes detected in these vessels by scRNAseq, Trpc6 is only detectable in LMCs (Fig. 5 panel A). With regard to consequences of global knockouts and possible compensation, that is why we generated the Trpc6/Trpc3 DKO mice, as upregulation of Trpc3 is a known compensatory mechanism for Trpc6 deletion (PMID: 16055711); to our knowledge only one other group has taken the trouble to test this (PMID: 16055711) with respect to vascular smooth muscle function.

(10) Did you perform power analysis to ensure that experimental numbers were sufficient to conclude that no statistical difference exists between datasets? If not, this needs to be done. For example, data shown in Figure 5C for tone and 6C for frequency and tone appear to be significantly different, but are concluded not to be so.

With respect to Fig. 5, there are significant differences in contraction frequency between the TrpC6/C3 DKO mice and their SV129 controls at 5 of 7 pressures, but the frequencies of the KO mice were actually HIGHER—the opposite of the result expected if TRPC6 or TRPC3 were critical for pacemaking. With respect to the effects of TRPC6 KO on tone, there were no significant differences in tone between TRPC6-/- vessels and their SV129 controls, but there were indeed significant differences between the TrpC6/C3 DKO mice and controls at 4 of 7 pressures, but with tone actually being HIGHER than normal (opposite to what would be predicted if those channels were critical for the development of myogenic tone). We point out how that observation is at odds with the conclusions of at least one study on arterial myogenic tone (PMID: 11861411), but that topic is not the focus of our study.

(11) At the end of each result section, a concluding statement is made regarding the effects on pressure-induced chronotropy. In many cases, there are additional effects of manipulating protein expression on other contractile properties. One example is for TRPC3 and TRPC6 (lines 414-416), but others are TRPV4, TRPV3, ENaC, Kir, Cav3.1/3.2, etc. Some interpretation is in the Discussion, but the concluding statements at the end of each result section should be expanded to summarize what the authors think the other significant differences in the data represent.

Thank you for allowing us the flexibility to address this at the end of each relevant section in the Results. We have now expanded several of these sections as you suggest.

(12) Kv7.4 channels. You state you have data (not shown) with linopiridine and XE991. Why not show those results here to support the experiments with the Kcnq4 smKO mice? Otherwise, I suggest you remove the statement from the unpublished data.

We have removed the comments about the unpublished data with other Kv7.4 inhibitors.

(13) Figure 13A. Kcnj2 is modestly expressed in LECs, but very little is present in LMCs. This likely underlies the effect of barium. If you remove the endothelium, does the effect of barium disappear? While this is not the major focus of the study, the effects of barium are dramatic, and it should be made clear whether this is due to inhibition of Kir channels in smooth muscle or endothelial cells.

Thank you for making this very good point. We cannot rule out an effect on LECs in this context and denuding popliteal LVs is nearly impossible without compromising their contractile function because of their small size and the presence of intraluminal valves. The valves prevent easy passage of an air bubble (the opening is only 1/4 the area of the lumen) and attempts at mechanical removal result in extreme valve damage and/or residual LECs at valve sites. If there is an influence of inhibiting Kir channels in LECs, it is not via electrical communication between LECs and LMCs, which is minimal or non-existent as we have shown in previous studies (PMID: 30143234, PMID: 28994159, PMID: 30355030). Thus, a potential LEC influence would be mediated by release/production of vasoactive factors, which is a point that we now discuss.

(14) Figure 18C tone. Several values for losartan look different but are not labelled as such. Please clarify and discuss if different.

All the significantly different values were labeled as such, but that panel is no longer in the revised paper as the AT1R frequency data have been included in a different form (Fig. 22).

(15) The manuscript should include raw data traces in figures that show the major pathways that you conclude regulate chronotropy.

Thank you for this suggestion. Three new figures have now been added to illustrate the blunting of F-P relationship in Ano1 smKO (Fig. 3), IP3R1 smKO (Fig. 17) and GNAQ/11 DKO vessels (Fig. 19).

Reviewer #2 (Public review):

Summary:

In this study, Davis et al. embarked on the quest for the molecular elements responsible for the regulation of lymphatic phasic contractile activity in response to variation of transmural pressure, a mechanism (termed pressure-induced lymphatic chronotropy by the authors) critical for drainage of interstitial fluid from the tissue and transport of lymph back to the blood circulation. Their aim was to investigate the mechanism(s) involved in the pressure-induced regulation of lymphatic pumping, and test whether activation of cation channels, shown in other systems to play mechanosensitive roles are directly at play, and/or whether mechano-activation of GNAQ/GNA11-coupled GPCRs is necessary to generate second messengers to activate those channels, as it has been suggested for the regulation of myogenic tone in arteries. To achieve their goal, the authors used their well-described, highly reliable protocols of mouse lymphatic vessel isolation, pressure myography, and data acquisition to obtain frequency-pressure relationships and other contractile function parameters from transgenic mice where specific channels or molecular elements of interest have been ablated. They combined these data with scRNAseq analysis of these gene targets to determine their respective role and levels of expression in lymphatic muscle cells. Their conclusion is that none of the exhaustive list of tested ion channels was critical, except ANO1 Cl channels, part of the contractile pacemaker mechanism, but that transmural pressure activates GNAQ/GNA11-coupled GPCRs, which generate IP3 to induce SR Ca2+ release through IP3R1 and activate ANO1-mediated depolarization.

Strengths:

The manuscript's strengths reside primarily in very robust, clean, and unequivocal pressure myography data and analysis. The research team is mastering these techniques they developed more than a decade ago and have implemented in mouse lymphatics to study their contractile properties, with consistent and convincing outcomes. They also provide data from an impressive list of transgenic mice in order to determine the role of the targeted gene in pressure-induced lymphatic chronotropy, relying on pharmacological small molecule inhibitors only when necessary. Finally, the use of scRNAseq analysis they gathered from previously published datasets brings novelty with respect to the expression of the genes of interest in all populations of cells comprising the lymphatic vessels, but more critically, to validate or contrast the potential impact of genetic alteration of the given gene on the ability of lymphatic muscles to respond to a change in pressure.

Weaknesses:

(1) The main weakness may reside in the fact that while the authors provide a convincing demonstration that GNAQ/GNA11 are involved in the regulation of the F-P relationship, they give little evidence of the involvement of "upstream" receptors. Indeed, inhibition of AT1R, shown to be involved in myogenic regulation of arteries (a phenomenon the authors rightfully compare to pressure-induced lymphatic chronotropy), didn't lead to a similar effect (decrease in F-P) in lymphatic vessels. Arguably, other GPCRs might be involved in lymphatic vessels, but as such information is not provided in the manuscript, the author's conclusions should be dampened. More in-depth discussion would be required. In fact, it can be argued that the discussion is very restricted with respect to the amount of data and information the manuscript provides.

To address these valid concerns, we performed a detailed scRNA-seq analysis of the GPCRs expressed in mouse LMCs. Approximately 136 GPCRs were identified as being expressed in over 0.5% of LMCs, the top 20 of which are expressed at moderate levels in a substantial percentage (>36%) of LMCs. We have added a new Figure showing bubble plots of these 20 GPCRs in the various cell populations of IALVs, ranked according to their expression level (Fig. 21); we also show their expression in the other cell types. We then experimentally tested 7 of the most likely candidates for which inhibitors were available by determining whether blocking the GPCR would alter pacemaking by lowering the contraction frequency at P=3 cm H2O. This was a streamlined assay compared to measuring the complete F-P relationship but produced similar results; to demonstrate this point, we also performed the streamlined analysis for the ANO1 inhibitor, Ano1 smKO, IP3R1 smKO and Gq/11 DKO, all of which showed significantly reduced frequencies, compared to their respective controls at P=3 cm H2O (see Fig. 22, black symbols). The alignment of statistical significance for the latter results with the corresponding F-P analyses shown in Figs. 2, 16 and 18) demonstrates that this assay is sufficiently powerful to detect a significant effect on pacemaking. Unfortunately, the results for each of the GPCRs tested were negative (Fig. 22, red symbols). We conclude that none of those particular GPCRs (NPY1R, ETAR, NPR3, TBA2XR, S1PR1, AT1AR, AT1BR), or two additional GPCRs (HT-2R, HRH1) that were tested in previous studies (PMID: 12770929, 28453392), are critical for pacemaking or the F-P relationship in LMCs. Of the remaining 11 GPCRs, several are adhesion receptors (Adgrl1, Adgrl2, Adgrl3, Adga2), two are orphan receptors (Gprc5b, Olfr1033), one codes for the GABA-B1 receptor (Gabbr1, with unknown function in the lymphatic system), two code for proteins in the Wnt signaling pathway (Fzd4, Fzd2), one codes for a protein in the Hedgehog signaling pathway (Smo) and one encodes a beta-amyloid binding protein (Tm2d1). The possible roles of these GPCRs, and the other 100+ GPCRs expressed at much lower levels remain to be tested in future studies. At any rate, systematic investigation of these GPCRs is well beyond the scope of this study, and in many cases not possible due the lack of known soluble ligands/antagonists. We have significantly expanded the Discussion with regard to this topic and stated that our inability to identify a specific Gq/11-coupled mechanosensory GPCR in LMCs is a limitation of our study.

(2) Overall, the authors convincingly achieved their aim by performing an impressive number of technically challenging experiments, leading to solid datasets. While these support their main conclusions, a more elaborate discussion might be required to refine them.

We hope this limitation has been tempered by the addition of the new data and the expanded Discussion section mentioned above.

(3) This study is likely to have an important impact on the field as it provides some answers to the lingering question of how lymphatic vessels regulate their contractile activity to variation in transmural pressure and certainly proposes an experimental means to further explore and address that question.

Thank you.

Reviewer #3 (Public review):

In this manuscript, Davis and colleagues aimed to identify the molecular sensors and signaling cascade that enable collecting lymphatic vessels to increase their spontaneous contraction frequency in response to intraluminal pressure (pressure-induced chronotropy). They tested whether the process is similar to blood vessel myogenic constriction by relying on cation channels (TRPC6, TRPM4, PKD2, PIEZO1, etc.) or instead require the activation of G-protein-coupled receptors (presumably mechanosensitive GNAQ/GNA11-coupled receptors), using ex vivo pressure myography of mouse popliteal lymphatics, smooth muscle-specific conditional knockouts, quantitative PCR validation, and single-cell RNA sequencing for target prioritization. The authors convincingly demonstrate that pressure-induced chronotropy does not require the cation channels implicated in arterial myogenic tone but is blunted by deletion of GNAQ/GNA11 or IP3 receptor 1, supporting a model of GPCR > IP3 > Ca2+ release > Cl⁻ channel activation > depolarization. The core conclusion is robust. The work redefines lymphatic pacemaking as G-protein-coupled receptor-dependent mechanotransduction, distinct from arterial mechanisms, and provides a genetically validated toolkit that is useful for studying lymphatic function and dysfunction.

Strengths:

(1) The data are of high quality and highly sensitive functional readouts

(2) The systematic genetic targeting is a major strength that overcomes pharmacological artifacts

(3) Careful quantitative analyses of frequency-pressure slopes

Weaknesses:

(1) The use of inguinal-axillary vessels for single-cell RNA sequencing rather than the popliteal segment studied functionally.

Agreed. The need to use IALVs derives from the much smaller amount of tissue needed for scRNAseq analysis of individual popliteal vessels (and 3-5x longer time required for cleaning) before pooling them. We have tried to clearly state this limitation. However, functional comparisons between popliteal and IALV function are quite similar in almost all aspects that we have studied. In the majority of our protocols, preliminary experiments using IALVs were performed side by side with popliteal vessels (for Ncx1 smKO, Trpc6-/-, Trpc6/3 DKO, G12/13 DKO, Piezo1 smKO, Trmp4 smKO) without detection of a significant difference in phenotype. In addition, we have performed new experiments using IALVs from G11-/- and GNAQ/GNA11 DKO mice to support the major positive findings drawn from the popliteal vessel studies (compare Fig. 18 with Suppl. Fig. 1).

(2) No direct testing of the specific G-protein-coupled receptor involved.

Agreed. As described in our response to the other reviewers (see point 1 to Reviewer 1 and point 1 to Reviewer 2), to address these concerns we have added a new figure (Fig. 21) ranking the top-expressed GPCRs in LMCs and have experimentally tested several of the most likely candidates (Fig. 22). Unfortunately, our results were still negative, allowing us to rule out those particular GPCRs but not to pinpoint a specific GPCR. The possible roles of other 100+ GPCRs in LMCs remain to be investigated in future studies. We have adjusted our title to better reflect our results.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) The authors state multiple times that several cation channels have been shown to contribute to vascular tone. Arterial smooth muscle cell ANO1 channels also contribute to arterial myogenic tone (PMID: 22872152). An ANO1-dependent mechanism is proposed here in lymphatic vessels, and the same ANO1 mouse models are studied. PMID: 22872152 is not discussed or cited and should be.

Now cited.

(2) Lines 88-97. It is important to note that the Bulley paper studied tamoxifen-inducible smooth muscle cell-specific knockout mice, whereas the Sharif-Naeini study used germline SM22-driven PKD1 knockout mice. The mice used in the Sharif-Naeini paper were subsequently found to have an ADPKD-like phenotype due to non-specific knockout of PKD1 in other cell types (PMID: 20856231). That should also be noted. The authors also state "PKD1 channel" in this paragraph when PKD1 does not form a channel.

Corrected and clarified. A new paper on PKD1 by Jaggar’s group is now cited in that context.

(3) Line 119. Perhaps you mean PKD1/2 channels?

Corrected.

(4) Line 312. It is unusual that data from Figure 2A is described before that in Figure 1. Can you reorder the figures to fit the text?

Corrected.

(5) Line 130. Can you provide references that show that ANO1 is not intrinsically mechanosensitive?

Added.

(6) Line 160. I believe the Pkd2f/f mice are available from the Maryland PKD Core and were not generated by the Jaggar lab.

Corrected.

(7) Line 319 needs a reference.

Added.

(8) The color choice for the symbols in many figures, particularly Figures 2, 3, and 4, makes it difficult to know which condition is which.

Corrected.

(9) I may have missed it, but the Figure 1A legend seems to need a definition for every cell type abbreviation.

It is shown in panel A (Figure 2).

(10) In several places, the authors state that they deleted gene expression in LMCs, when the mouse model deletes expression from all smooth muscle cells. Please refine wording.

Clarified.

(11) Line 664-666. Sharif-Naeini did not propose that PKD2 channels contribute to myogenic tone. That was Bulley et al.

Corrected.

Reviewer #2 (Recommendations for the authors):

(1) The role of TRPV4: It is interesting to note that despite a quasi-non-expression of the TRPV4 gene in LMC, KO this gene in these cells results in an increase in contraction frequency and a strong decrease in amplitude at low pressures, suggesting the channel is expressed and active in LMCs. It is not too clear how macrophages could be indirectly involved in this process by releasing products. Are they also sensitive to pressure? How? And why should they be more involved than any other cell type present in the vessel wall (i.e, LECs)?

Macrophages definitely release products that affect LMC activity (PMID: 38826322) and they may be sensitive to pressure but that has not yet been demonstrated. However, as shown in the paper cited, selective KO of Trpv4 in macrophages does not abrogate the F-P relationship in lymphatic vessels.

(2) The role of AT1R in pressure-induced chronotropy: The lack of effect of AT1R inhibition on F-P doesn't mean other GPCRs could play a similar role in lymphatics. A more in-depth discussion could be provided. Furthermore, could G-proteins be activated by pressure independently of coupled receptors?

We have extensively expanded our discussion of possible mechanosensitive GPCRs in LMCs. We have also cited at least one paper providing evidence for G-proteins being activated by pressure independently of coupled receptors (PMID 28148497), although we believe our added assessment of LMC GPCR expression reveals novel targets to explore for mechanotransduction in LMCs.

(3) Figure 1: Please explain the apparent discrepancy between the described protocol and the experiment displayed in A. Please explain and confirm that the experiments presented in the study followed the protocol described in the Methods section. Add text to Methods.

The example does indeed follow the stated protocol in the “METHODS section” for the sequence and duration of pressure steps. Note that we stated 2 min was typical at each pressure level but for some of the lower pressures 4-5 min were required to get a sufficient number of contractions, as in this case. The diameter and pressure traces at P=8, 10 cm H2O are not shown in the example so that we could expand the time axis for better resolution.

It can also be inferred from the trace that the F-P curve was built from frequencies obtained when pressure was lowered and increased. Are values at a given pressure identical if recorded when pressures were lowered or increased? Does it influence the relationship?

There is indeed some variability in the response to stepping pressure up vs down because of rate-sensitive effects, as described in one of our previous studies (PMID: 19001046). In the protocol we implemented here, the FREQ at P=3 cm H2O upon step up from 0.5 cm H2O was usually slightly higher than the initial FREQ at P=3 cm H2O. Those two values were averaged together and this is now stated.

The FREQ graph in B displays values up to 10 cm H2O, which are not shown in A.

The data at P=8, 10 cm H2O were obtained for that same vessel. The complete range is shown to illustrate that the relationship reaches a plateau above 5 cm H2O.

(4) Figure 1: Is there a need to repeat/duplicate the FREQ graph in B and D?

It has now been replaced with Ejection Fraction.

(5) Figure 19: With respect to point #2, and whether G-proteins are involved/important in the process, given the current state of understanding, that part of the illustration might need to be amended.

The figure has been revised.

Reviewer #3 (Recommendations for the authors):

Below are some suggestions for the authors to consider:

(1) Test specific GPCRs pharmacologically (e.g., AT1R) on wild-type vessels to determine if pressure-induced chronotropy is blunted.

Done, Figure 22.

The authors could also rank Gq/11 receptors based on the expression in LMCs (e.g., qPCR using FACS-sorted cells).

Done, Figure 21.

(2) Residual chronotropy in ANO1 knockout mice: Quantify the remaining frequency-pressure slope in Myh11-CreERᵀ²;Ano1ᶠ/ᶠ vessels and test whether it is sensitive to low-chloride buffer or chloride channel blockers to probe an alternative Cl⁻ conductance.

The role of chloride in lymphatic pacemaking is well appreciated and the reviewer offers some intriguing experiments for future experimentation. The use of low Cl- solutions was part of the fundamental discovery of lymphatic muscle pacemaking and the role of a calcium-activated chloride channel as the basis for spontaneous transient depolarizations (STDs) by Dirk Van Helden in 1993 who reported that “Lymphatic STDs reversed at potentials near -35 mV (range -40 to -25 mV; n = 4)” and were “suppressed by exposure to low chloride (replaced with sodium isethionate) solution [(> 5 minutes)]”, although more details were not provided. The effect of low Cl- and lymphatic muscle excitability was followed up in a 2008 paper by von der Weid et al., which showed that STD frequency and amplitude increased acutely (within 2-3 min) after the bath solution was changed to a solution containing 10% of the initial Cl- concentration (replaced with methane sulphonate). This result likely suggests that the acute reduction of extracellular Cl- increased the reversal potential for Cl- and that within 2-3 minutes the intracellular Cl- content had not quite run down. Work by Boedtkjer’s group (Mohanakumar et al. 2018) also demonstrated that human lymphatic pacemaking was also heavily dependent on the bath Cl- concentration. Using wire myography preparations, they showed that human thoracic rings and mesenteric vessels ceased spontaneous contractions when Cl- was removed (replaced with aspartate), and that contractions returned upon washout with a normal Cl- containing solution. In that study, the authors noted that there was some variability in the timing required for contractions to cease with the mesenteric LVs. In some cases, an acute large constriction immediately followed the exchange with Cl- free solution, which would also point to an acute depolarizing/stimulatory effect of a rapid resetting of ECl to a more positive value when external Cl- was removed acutely removed. A limitation to the approach is the use of wire myographs as opposed to pressurized vessels, as the wire method does not allow the vessels to be stretched in a physiological manner. Similarly, diastolic depolarization was not typically observed by Van Helden and von der Weid, as they used pinned-out tissue sections (required for recording STDs from short vessel sections), which highlights the necessity of using a physiological stretch stimulus.

On the face of it, the complete cessation of contractions reported in the Boedtkjer paper is at odds with our previous and present results using Ani9 and Ano1 smKO and IP3R1 smKO vessels, which retain some baseline level of pacemaking, albeit largely pressure-independent. Thus, the hypothesis that a separate Cl- channel may be active and important holds some merit. However, it is worth noting that we do not expect this basal pacemaker drive to be calcium-dependent, as IP3R1 calcium oscillations are maintained. IP3R1 smKO vessels, which largely lack subcellular calcium transients in diastole, also have a lower-than-expected contraction frequency compared to Ano1 smKO vessels, although this is in part due to an elongation of the action potential plateau. We also do not think the residual activity is simply due to the inability of inducible Cres to drive complete recombination, as we performed experiments with Ani9-treated Ano1 smKO vessels (Fig. 2B) and did not observe a significant additional reduction in the concentrations at which maximal inhibition is assumed, 2-10 mM (see also Harlow et al. 2025). We do not see a robust presence of Ano2 (TMEM16b) and while Ano6 (TMEM16) is expressed, it is a calcium-activated scramblase with permeability to both cations and Cl- (Ye et al 2019).

We also have personal experience with Cl- substitution (replaced to 10 mM by aspartate), originally obtained as part of our work identifying Ano1 as the primary calcium-activated chloride channel, although the inability to ascertain the intracellular Cl- at any point in time while the bath exchange is occurring significantly limited the interpretability of that data. During review of that manuscript, we were asked to remove the Cl- substitution experiments but spontaneous contractions still persisted in the majority of the vessels (n=6). Notably this was not a complete removal of Cl- as was done in Mohanakumar et al. 2018. However, the complete removal of Cl- is a much more significant intervention than the inhibition of a single channel, likely altering HCO3- and Na+ flux through the associated Cl- antiporters and symporters. How Cl- removal affects LMC intracellular pH regulation via impaired bicarbonate transport also remains unknown. Nor is it well appreciated how Cl- removal affects mitochondrial function and ROS production in LMCs, which regulate potentially significant confounding pathways related to lymphatic muscle pacemaking. Without control of these significant confounding circumstances, the hypothesis that the baseline depolarization is still a Cl- channel is far from a certainty.

In Author response image 1, we provide the reviewer with the expression data for other members of the Anoctamin family, the potential Anoctamin-regulating CLCA family, and other documented chloride channels in LMCs (clusters 5-6).

(3) I understand this would be a huge undertaking, but aligning scRNAseq with functional vessels using popliteal LMCs or inguinal-axillary LMCs could help rule out potential tissue-specific differences.

We understand the reviewer’s concerns and, owing to the small amount of cells that can be recovered from popliteal vessels, we instead now provide new data using pressure myography of inguinal axillary vessels (IALVs) for the data with positive findings (Gq/11 inhibition) that support our findings in popliteal vessels (Suppl. Fig. 1). This points to a fundamental role for Gaq, Ga11, and possibly Ga14, in signaling upstream of IP3R1 mobilization and Ano1 activation. We are aware of the length of this manuscript (now further increased in order to respond to the reviewers’ suggestions) and, given the negative data for the vast majority of putative mechanosensory pathways, we have omitted some of the parallel data collected in IALVs. In the majority of our protocols, preliminary experiments using IALVs were tested side by side with popliteal vessels (from Ncx1 smKO, Trpc6-/-, Trpc3/6-/-, G12/13 DKO, Piezo1smKO, Trmp4 smKO mice) without detection of a significant difference in phenotype.

Version 2

Abstract

Active lymph pumping relies on the spontaneous contractions of collecting lymphatic vessels, whose contraction frequencies are exquisitely sensitive to changes in intraluminal pressure. This homeostatic and mechanosensitive mechanism, termed pressure-induced lymphatic chronotropy, enables lymph transport to be matched to the filling state of the lymphatic capillary network. The mechanistic basis of pressure-induced chronotropy was investigated using ex vivo contraction assays of mouse popliteal collecting vessels, in which contraction frequency increases >10-fold with pressure changed from 0.5 to 5 cmH2O. The contractile, electrophysiological and transcriptional similarities between lymphatic muscle cells (LMCs) and arterial smooth muscle led us to hypothesize that pressure-dependent chronotropy shares a parallel signaling process with pressure-induced arterial depolarization/constriction. Thus, we probed two major mechanisms: 1) pressure-induced activation of mechanosensitive cation channels, including TRPC6, TRPM4, PKD1/2, TRPV2 and ENaC, and 2) mechano-activation of GNAQ/GNA11-coupled G-protein receptors (GPCRs) that would generate second messengers to activate those channels. Contraction assays were combined with scRNAseq analysis of the respective targets, with maximum use made of transgenic mice to avoid non-specific effects of pharmacological inhibitors, particularly those used to block TRP channels. Our findings rule out significant roles for the above TRP channels and other putative mechanosensitive channels implicated in arterial myogenic constriction, as well as channels implicated in ionic pacemaking of other tissues. In contrast, smooth-muscle specific knock out or inhibition of ANO1 or IP3R1 significantly blunted the effect of pressure on frequency. Pressure-induced chronotropy was suppressed by ∼70-90% at all pressures in GNAQ/GNA11 double knockout vessels, but with responsiveness partially maintained at pressures above 5 cmH2O. Pressure-induced chronotropy was also suppressed after acute Gq/11 inhibition with YM254890, but was normal in vessels from GNA12/GNA13 double knock out mice. These results support a scheme whereby mechano-activation of one or more GNAQ/GNA11-coupled GPCRs generates IP3, which induces SR Ca2+ release through IP3R1 and drives depolarization through the activation of ANO1 Cl- channels. The major GPCRs expressed in LMCs were subsequently identified and ranked by scRNAseq analysis but knock out or pharmacological inhibition of each of the top 7 candidates failed to significantly affect pressure-induced chronotropy. Our results strongly implicate one or more GNAQ/GNA11-coupled GPCRs in mediating this homeostatic process; however, the specific mechanosensitive GPCRs remain to be identified.

Introduction

The lymphatic system is critically important for the maintenance of interstitial fluid balance. Because the Starling forces governing fluid movement across blood capillaries favor net filtration in most tissues (Levick & Michel, 2010), filtered fluid and protein will accumulate in the interstitium unless it is removed by the lymphatic system. Lymphatic capillaries are optimized to reabsorb that fluid, which can exceed 8 L·day-1 (Renkin, 1986). Once reabsorbed, lymph is transported by a combination of passive and active transport mechanisms through the collecting lymphatic vessel network, first to lymph nodes and eventually back into the venous system. Active lymph pumping relies on the spontaneous contractions of collecting lymphatic vessels—events that are initiated by action potentials (APs) in lymphatic muscle cells (LMCs). These contractions are relatively large in amplitude, with ejection fractions often exceeding 80% and sufficient to propel lymph through a system of one-way valves that ensure its unidirectional movement (Davis et al., 2025). Active pumping accounts for the majority of lymphatic transport in the lower legs of humans during quiet standing (Olszewski & Engeset, 1980) and facilitates lymph return against adverse hydrostatic pressure gradients that develop under gravitational loads (Davis et al., 2012; Scallan et al., 2016; Li et al., 2022). An important feature of the active lymph pump system is the exquisite sensitivity of the spontaneous contraction frequency of collecting lymphatic vessels to changes in their intraluminal pressure, enabling lymph transport to be matched to the filling state of the lymphatic capillary network (Scallan et al., 2012). This homeostatic mechanism is termed “pressure-induced lymphatic chronotropy” (Zawieja et al., 2019). In popliteal collecting vessels of the mouse hindlimb, the spontaneous contraction frequency can increase from 2 to 20 contractions·min-1 as pressure is elevated from 0.5 to 10 cmH2O, i.e., a 10-fold increase over the physiological pressure range of those vessels (Davis et al., 2020; Davis et al., 2023a; Davis et al., 2023b). Contraction amplitude increases modestly over a portion of the same pressure range (Scallan et al., 2012), but modulation of contraction frequency is the primary method for acute adjustments in active lymph transport.

Pressure-induced chronotropy reflects an underlying mechanotransduction process that ultimately involves control of an intrinsic pacemaker in LMCs by intraluminal pressure. Although the specific force sensing mechanism is unknown, contractile, electrophysiological and transcriptional similarities between LMCs and vascular smooth muscle cells (VSMCs) support the hypothesis that pressure-dependent chronotropy shares a parallel signaling process to the myogenic response of VSMCs. The myogenic response refers to the rapid increase in arterial pressure induced by distention of a small artery / arteriole, which then constricts over time until a new, stable diameter smaller than the original diameter is reached (Davis, 2012). A similar constriction often occurs in collecting lymphatic vessels coincident with pressure-induced chronotropy (Davis et al., 2009). Pressure-induced arterial constriction is preceded by VSM depolarization (Harder, 1984) and several lines of evidence support a critical role for the activation of (putative) mechanosensitive cation channels in that process—in particular the involvement of three TRP channel family members: TRPC6, TRPM4 and PKD1/2.

The first evidence for a molecularly identified ion channel involved in myogenic constriction came from the demonstration that single-channel currents in VSMCs isolated from rat cerebral arteries were activated by patch pipette suction and those currents, along with pressure-induced cerebral artery constrictions and depolarizations, were attenuated after downregulation of TRPC6 with antisense oligonucleotides (Welsh et al., 2002). TRPM4 channels were likewise implicated after inward, TRPM4-like currents in rat pial artery myocytes were found to be activated by membrane stretch, and because antisense- or siRNA-mediated knock down of TRPM4 significantly attenuated pressure-induced depolarization and myogenic tone development in those arteries (Earley et al., 2004b; Gonzales et al., 2014). A recent study confirms that SM-specific TRPM4 knockout significantly impairs myogenic tone (Zhu et al., 2026). In contrast, results from Trpc6-/- and Trpm4-/-mice tended to contradict the above studies in finding no differences or even enhanced arterial myogenic constriction in the global knock out animals (Dietrich et al., 2005; Mathar et al., 2010; Schleifenbaum et al., 2014). Evidence also supports roles for PKD2 (previously described as TRPP1) channels in myogenic constriction, but that, too, is controversial. Jaggar and colleagues reported that shRNA-mediated down-regulation of Pkd2 attenuated pressure-induced depolarization and myogenic tone in rat pial arteries (Narayanan et al., 2013) and that pressure-induced constriction was blunted in hindlimb arteries isolated from Myh11-CreERT2;Pkd2f/f (smooth muscle knockout, smKO) mice (Bulley et al., 2018). However, a study by Honoré and coworkers suggested that stretch-activated cation currents and myogenic constriction in mouse mesenteric artery smooth muscle were mediated by PKD1 channels (previously termed TRPP2), and that PKD2 channels played a counteracting role (Sharif-Naeini et al., 2009). Subsequent studies show that myogenic tone is impaired in Pkd2 smKO mice (Bulley et al., 2018) but not in Pkd1 smKO mice (Bernardelli et al., 2026).

Recent work points to G-protein-coupled receptors (GPCRs) rather than TRP or other ion channels as the primary mechanosensing elements in the arterial myogenic response. Studies of cardiomyocyctes first demonstrated that the angiotensin receptor, AT1R, a GNAQ/GNA11-coupled GPCR, could be activated by mechanical stress through a mechanism independent of its ligand (Zou et al., 2004). Gudermann and coworkers provided compelling support for this mechanism using patch-clamped HEK293 cells, co-expressing AT1R and TRPC6 channels, with hypoosmotic swelling used as a mechanical stimulus (Mederos y Schnitzler et al., 2008). Two critical findings were 1) that TRPC6 current was activated by hypoosmotic swelling only if the cells co-expressed Agtr1 (Mederos y Schnitzler et al., 2008) and 2) that the swelling-induced currents were largely prevented by inverse AT1R agonists. Agtr1 could be replaced by another GNAQ/GNA11-coupled GPCR, or TRPC6 by another DAG-sensitive TRP channel (Mederos y Schnitzler et al., 2008). Subsequent studies from multiple laboratories, using selective pharmacological inhibitors of AT1R, acute Agtr1 knockdown, and/or mice deficient in Agtr1 or other GNAQ/GNA11-coupled GPCRs, confirmed the basic principle in VSMCs: that pressure-induced constriction of many different types of arteries is transduced primarily, or exclusively, by GNAQ/GNA11-coupled GPCRs (Blodow et al., 2014; Harraz et al., 2014; Li et al., 2014b; Schleifenbaum et al., 2014; Pires et al., 2017; Bjorling et al., 2018; Cui et al., 2022). Helix 8 in the GPCR c-terminus was subsequently identified as the essential structural motif endowing mechanosensitivity (Erdogmus et al., 2019). The collective data support a model in which phospholipase C is activated downstream from GNAQ/GNA11-coupled GPCRs in response to elevated pressure, leading to increased production of DAG and IP3/Ca2+ release, which then activate TRPC6 and TRPM4 channels, respectively, to depolarize the cell and increase the open probability of L-type voltage-gated Ca2+ channels, thereby promoting global Ca2+ influx and contraction (Davis et al., 2023d). In a similar manner, TRPM4 and PKD2 channels might be activated by Ca2+ influx/release downstream from mechano-activation of a GPCR or another Ca2+-permeable ion channel such as PIEZO1 (Peyronnet et al., 2013) or TRPV4 (Swain & Liddle, 2021).

Whether similar signaling pathways are operative in LMCs is unknown. Each lymphatic contraction is preceded by an approximately linear depolarization from the “resting” potential to threshold potential, termed diastolic depolarization, which is the primary determinant of AP/contraction frequency (Zawieja et al., 2025). Thus, lymphatic pressure elevation manifests not as a stepwise increase in baseline LMC membrane potential (von der Weid et al., 2014; Davis & Zawieja, 2018) but rather as an increase in the diastolic depolarization slope and a reduction in time required to reach threshold for AP firing (Zawieja et al., 2018b; Zawieja et al., 2019). Previously, we found that regulation of the diastolic depolarization slope was mediated largely by activation of the Ca2+ activated chloride channel ANOCTAMIN 1 (ANO1, also referred to as TMEM16A). ANO1 is not thought to be intrinsically mechanosensitive, but regulated by Ca2+ and, in LMCs specifically, by IP3 receptor 1-mediated Ca2+ release from sarcoplasmic reticulum (Zawieja et al., 2019; Zawieja et al., 2023), possibly downstream from one or more mechanosensitive GPCRs. However, other mechanisms may also be involved because Ano1 deletion from smooth muscle or pharmacological inhibition of ANO1 does not completely eliminate spontaneous APs or contractions (Zawieja et al., 2019). The primary goal of the present study was to test the roles of putative VSMC mechanosensitive ion channels, other channels implicated in the regulation of pacemaking in different cell types, and GNAQ/GNA11-coupled GPCRs in pressure-induced chronotropy of popliteal lymphatic vessels. An advantage in studying this particular aspect of lymphatic vessel mechanotransduction is the high sensitivity of the spontaneous contraction frequency to pressure, increasing ∼10-fold over only a 5 cmH2O pressure range in popliteal lymphatics compared to relatively modest levels of myogenic constriction (50% change) over a 60 mmHg pressure range, even in the most myogenically reactive arteries (Davis et al., 2023d). As part of our strategy, we sought to make maximum use of transgenic mice to avoid non-specific effects of pharmacological inhibitors, particularly those used to block TRP channels.

Materials and Methods

Protocol approval

All procedures were reviewed and approved by the animal care committee at the University of Missouri and complied with the standards stated in the “Guide for the Care and Use of Laboratory Animals” (National Institutes of Health, revised 2011).

Mice

WT mice [C57Bl/6 (#000664) and SVF129 (#101043) strains], Slc8af/f (#025943) mice and AT1aR-/- (#002682) mice were purchased from The Jackson Laboratory (Bar Harbor, Maine). Myh11-CreERT2 mice were gifts from Stefan Offermans (Max Plank Institute, GDR). Sperm from Gna12-/-;Gna13f/f and Gna11-/-;Gnaqf/f mice were gifts from Stefan Offermanns, after which the respective mice were rederived at MMRC, Columbia, MO to generate Myh11-CreERT2;Gna13f/f;Gna12-/- mice and Myh11-CreERT2;Gnaqf/f;Gna11-/- mice. Slc8af/f mice were crossed with Myh11-CreERT2 mice to generate Myh11-CreERT2;Slc8af/f mice. Trpc6-/- and Trpc3-/- mice were gifts from Lutz Birnbaumer (NIH); Trpc6-/-;Trpc3-/-double KO mice were generated by crossing those two strains. Myh11-CreERT2;Kcnq4f/f (encoding Kv7.4) mice were a gift from Thomas Jentsch (Free Univ. of Berlin, GDR). Piezo1f/f mice were a gift from Ardem Pataputian (UCSD) and were bred to Myh11-CreERT2 mice to generate Myh11-CreERT2; Piezo1f/f mice. Ano1f/fmice, generated in the lab of Jonathan Jaggar (Leo et al., 2021) were bred to Myh11-CreERT2 mice to obtain Myh11-CreERT2;Ano1f/f mice. Itpr1f/f mice were obtained from Ju Chen (UCSD) and bred to Myh11-CreERT2 mice to generate Myh11-CreERT2;Itpr1f/f mice. Pkd2f/fmice were generated in the University of Maryland PKD Core by Jonathan Jaggar, as previously described (Bulley et al., 2018), and bred in the Jaggar laboratory to Myh11-CreERT2 mice to generate Myh11-CreERT2;Pkd2f/fmice. Trpm4f/f mice were generated in the laboratory of Scott Earley and bred to Myh11-CreERT2 mice to produce Myh11-CreERT2;Trpm4f/fmice. Trpv4f/f mice were generated in the laboratory of Timothy Domeier (University of Missouri) and bred to Myh11-CreERT2 mice to produce Myh11-CreERT2;Trpv4f/fmice. Rosa26mTmGf/f mice were purchased from JAX (#007676) and bred to Myh11-CreERT2 and Myh11-CreERT2;Ano1f/fmice to produce Myh11-CreERT2;Rosa26mTmGf/f and Myh11-CreERT2;Ano1f/f; Rosa26mTmGf/f mice.

For genotyping, genomic DNA was extracted from tail clips using the HotSHOT method. Genotypes were determined by PCR with 2x PCR Super Master Polymerase Mix (Catalog # B46019, Bimake, Houston, TX) according to the provider’s instructions. Mice containing Myh11-CreERT2 were induced with five consecutive daily injections of 100 mg tamoxifen (10mg/kg i.p.; in safflower oil) and the respective KO mice generated by crossing these to mice with floxed alleles are referred to as SM-specific (smKO) mice. Those mice were studied a minimum of 2 weeks after induction. All floxed control mice were subjected to the same induction protocol. Because this version of Myh11-CreERT2 used for SM-specific deletion of floxed genes is carried on the y chromosome, only male mice were used for experiments involving Myh11-CreERT2 mice or their floxed controls. For other strains, both male and female animals were used.

Vessel isolation, pressure myography, and data acquisition

Mice were anesthetized with ketamine/xylazine (100/10 mg/kg, i.p.) and placed face down on a heated tissue dissection/isolation pad. The saphenous vein was exposed by a proximal-to-distal incision along the skin of the calf and the popliteal afferent lymphatic vessels on each side of the vein were isolated as previously described (Scallan & Davis, 2013). For IALVs, the skin was cut from the top of the hip to the shoulder and the skin retracted and pinned. The IALV was often found adjacent to the thoracoepigastric vein. Each vessel was then pinned with short segments of 40 µm stainless steel wire onto the SYLGARD-coated surface of a dissection chamber filled with BSA-supplemented Krebs buffer at room temperature. Once secured, the surrounding adipose and connective tissue were removed by microdissection. An isolated popliteal lymphatic vessel or IALV was then transferred to a 3-mL observation chamber, cannulated, pressurized to 3 cmH2O using two glass micropipettes (50-60 µm outside diameter) and moved to the stage of a Zeiss inverted microscope. Lymphatic segments used in these studies typically contained a single valve. Polyethylene tubing attached to the back of each glass micropipette was connected to a two-channel computerized pressure controller (Scallan et al., 2013). To minimize diameter-tracking artifacts associated with longitudinal bowing at higher intraluminal pressures, input and output pressures were briefly set to 10 cmH2O for popliteal and 8 cmH2O for IALVs at the beginning of every experiment, and the vessel segment was stretched axially to remove longitudinal slack. The lymphatic vessel was then allowed to equilibrate at 37°C with inflow and outflow pressures set to 3 cmH2O. Constant exchange of Krebs buffer was maintained using a peristaltic pump at a rate of 0.5 mL/min. After temperature stabilized, popliteal lymphatics typically began to exhibit spontaneous contractions within 10-15 min and stabilized in a consistent contraction pattern within ∼30 minutes. Custom LabVIEW programs (National Instruments; Austin, TX) acquired real-time analog data and digital video through an A-D interface (USB-6216, National Instruments) and detected the inner diameter of the vessel at 30 fps using a Basler A641fm firewire camera (Davis, 2005). Videos of the contractile activity of lymphatic vessels were recorded for further analysis, if needed.

Assessment of ex vivo contractile function

The contractile parameters of each vessel were characterized at different levels of intraluminal pressure spanning the physiological range from 0.5 to 10 cmH2O (typically in successive steps as follows: 3, 2, 1, 0.5, 3, 5, 8, and 10 cmH2O), except IALVs which only went to 8cmH2O. Spontaneous contractions were recorded at each pressure in typical intervals of 2 minutes, although a period of 4-5 min was sometimes required at the lowest pressure to obtain multiple contractions. During the pressure response protocol, both the input and output pressures were maintained at equal levels so that there was no imposed pressure gradient for forward flow. At the end of every experiment, all vessels were equilibrated by perfusion with calcium-free Krebs buffer containing 3 mM EGTA for 30 minutes, after which passive diameters were obtained at each level of intraluminal pressure.

Contractile function parameters

Once an experiment was completed, internal diameter traces and/or 30-fps brightfield videos of spontaneous contractions were analyzed using custom-written LabVIEW programs (Davis, 2023b, a) to detect end diastolic diameter (EDD), end systolic diameter (ESD), and contraction frequency (FREQ, computed on a contraction-by-contraction basis), with each parameter averaged over a 2-5 min period. These data were used to calculate the following commonly reported parameters that characterize the contractile function of lymphatic vessels: (1) Amplitude = EDD-ESD

where DMAX represents the maximum passive diameter (obtained after incubation with calcium-free Krebs solution) at a given level of intraluminal pressure. When frequency was zero, a value for amplitude was omitted. The relative paucity of some amplitude data at low pressures in some genotypes reflects the fact that this parameter could not be determined when frequency was zero. Each of the contractile parameters represented the average of all the recorded contractions during the measurement period at each intraluminal pressure. In popliteal vessels, data obtained for the two different periods at 3 cmH2O at the beginning and middle of the protocol were averaged together.

Analysis of pressure-induced chronotropy

Quantification of the frequency-pressure (F-P) relationship for comparisons across genotypes required additional analysis steps, as illustrated in Fig. 1. Raw recordings of pressure and diameter (Fig. 1A) show the responses of a WT popliteal vessel to descending pressure steps from 5 to 0.5 cmH2O, with the average frequency stated for each pressure. The frequency data were then plotted as a function of pressure (Fig. 1B). As evident from the behavior of this representative vessel, frequency tended to plateau above 5 cmH2O, so the subset of data in which the response was nearly linear over the pressure range 0.5-5 cmH2O was fitted with a first-order polynomial to obtain the slope and intercept (Fig. 1C). As an alternative method, we considered calculating the ratio of the respective contraction frequencies at 0.5 and 5 cmH2O; however, in many cases no contractions occurred at P = 0.5 cmH2O, which would equate to a ratio of infinity. To avoid this problem, we computed the difference (ΔF) between the frequencies at 5 and 0.5 cmH2O. This approach gave consistent values that, like the slope, could detect blunting of the F-P relationship, but in essentially every case the results matched those of the curve fitting procedure; thus, only the F-P slope was used for statistical analyses and shown in subsequent figure plots. To enable a more comprehensive assessment of contractile function, full plots of amplitude, frequency, FPF and tone were also plotted as a function of pressure for each genotype (Fig. 1D).

Quantification of the F-P relationship.

Steps for quantitative assessment of pressure-induced chronotropy. A) Representative recording of pressure and diameter changes in a WT popliteal lymphatic vessel during inflow (Pin) and outflow (Pout) pressure steps between 0.5 and 5 cmH2O. Frequencies stated at the bottom of the trace were determined off-line using a custom peak detection program written in LabVIEW. B) Plot of frequency vs pressure for the vessel in A, showing how frequency reaches a plateau above 5 cmH2O. C) Fit of the frequency data with a first-order equation and calculation of the frequency difference (ΔF) between 5 and 0.5 cmH2O. D) Summary of contraction data for this representative WT vessel over the stated pressure range; see Methods for description of parameter calculations.

Effects of selective GPCR inhibitors

To screen for the effects of specific GPCR antagonists, pressures were set to 3 cmH2O and contraction frequency was determined over a 2-5 min period. The inhibitor was added to the bath (with perfusion temporarily stopped), mixed thoroughly, and after waiting for at least 2 min, the average frequency was again determined over the subsequent 2-5 min period.

FACS

Inguinal-axillary lymphatic vessels from tamoxifen-treated Myh11-CreERT2;mTmGf/f and Myh11-CreERT2;Ano1f/f;mTmGf/fmice were dissected and cleaned of fat and connective tissue as described previously (Zawieja et al., 2018a). Cleaned vessel segments were transferred to a 1-ml tube of low-Ca2+ PSS containing (in mM): 137 NaCl, 5.0 KCl, 0.1CaCl2, 1.0 MgCl2, 10 HEPES, 10 Glucose, and 1 mg/ml BSA at room temperature for 10 min. The solution was decanted and replaced with a similar solution containing 26 U/ml papain (Sigma, St. Louis, MO) and 1 mg/ml dithioerythritol. The vessels were incubated for 30 min at 37°C with occasional agitation, then transferred to a new tube containing low-Ca2+ PSS containing 25 mg/ml collagenase H (FALGPA U/ml, Sigma), 0.7 mg/ml collagenase F (Sigma), 20 mg/ml trypsin inhibitor (Sigma), 1mg/ml elastase (Worthington), and incubated for 6 min at 37°C. The dispersed cells were sedimented by centrifugation (300 g, 4 min), resuspended in 0.6 ml PSS containing 1mM Ca solution, and filtered through a 35-µm mesh nylon filter to obtain a single cell suspension. GFP+ LMCs were sorted by fluorescence-activated cell sorting (FACS) using a Beckman-Coulter MoFlo XDP instrument with excitation laser (488 nm) and emission filter (530 ± 40 nm), a 70 µm nozzle, a sheath pressure of 45 psi and sort rate of 100 events per second. Sorting was performed at the Cell and Immunobiology Core facility at the University of Missouri.

RNA isolation and quantitative, real-time PCR

Total RNA was extracted from FACS-sorted cells using the Arcturus PicoPure RNA isolation kit (ThermoFisher Scientific, Waltham, MA) with on-column DNase I treatment (Qiagen, Valencia, CA) according to manufacturer’s instructions. RNA was eluted with nuclease-free water. Purified RNA was transcribed into cDNA using High-Capacity cDNA Reverse Transcription kit (ThermoFisher Scientific, Waltham, MA). Real-time PCR (qPCR) was performed on cDNAs prepared from each sample using 2x PrimeTime Gene Expression Master Mix (IDT, Coralville, IA) with predesigned TaqMan probes as listed in Suppl. Table 1 (IDT, Coralville, IA). Real-time PCR protocols were as follows: preheating at 95°C for 3 min, 45 cycles of two-step cycling of denaturation at 95°C for 15 sec and annealing/extension steps of 30 sec at 60°C. Data collection was carried out using a Bio-Rad CFX 96 Real-Time Detection System (software version Bio-Rad CFX Manager 3.1; Bio-Rad, Hercules, CA, USA). For analysis, the result was expressed as a ratio of target gene/reference gene (β-actin).

scRNAseq analysis

We assessed G-protein and GPCR expression using our previously published scRNAseq dataset (Zawieja et al., 2025) as detailed below. This dataset was derived from a total of 10 Rosa26mTmG mice (5 males and 5 females), without Cre and without tamoxifen treatment were used for scRNAseq analysis. Inguinal-axillary lymphatic vessels (IALVs) from both sides of each mouse were isolated and cleaned of connective tissue, adipose tissue and associated capillaries and small arterioles. The pooled vessels were digested into a single cell suspension and the cells were kept on ice until all tissues had been processed. Cells from all vessels were combined and sorted for tdTomato expression to remove debris and concentrate the cells for downstream single cell 3’ RNA-Seq libraries creation with 10x Genomics Chromium Chip and Chromium Next GEM Single Cell 3’ RNA-Seq reagents. Samples were sequenced with the NovaSeq 6000 S4-PE100 flow cell. Mus musculus genome GRCm39 and annotation GTF (v106) from Ensembl (https://useast.ensembl.org/Mus_musculus/Info/Index) were used to build the reference index and the reads were processed using Cell Ranger (v7.0.1) using the default parameters. The quality control and filtering steps were performed using R (v4.2.1; https://www.r-project.org/). Ambient RNA was removed from the Cell Ranger output with SoupX (Young & Behjati, 2020). A doublet score for each cell was estimated using scDBlFinder [v1.12.0; (Germain et al., 2021)]. Non-expressed genes (sum zero across all samples) and low-quality cells (>10% mitochondrial genes, < 500 genes, < 1,000 UMIs per cell and doublet score <0.5) were removed with custom R scripts. Cells passing filtering were normalized/scaled (SCTransformation), dimensionally reduced (t-distributed stochastic neighbor embedding (t-SNE), clustered using uniform manifold approximation and projection (UMAP), and hierarchically analyzed with Seurat (Hao et al., 2021; Hao et al., 2024) using the default parameters. Marker gene expression profiles on cell clusters and gene co-expression were visualized using Seurat and the ShinyCell R app (Ouyang et al., 2021; PMID: 33774659). The raw scRNAseq dataset is available on the NIH GEO #GSE277843 (Zawieja et al., 2025).

Solutions and chemicals

Krebs buffer contained: 146.9 mM NaCl, 4.7 mM KCl, 2 mM CaCl2·2H2O, 1.2 mM MgSO4, 1.2 mM NaH2PO4·H2O, 3 mM NaHCO3, 1.5 mM Na-HEPES, and 5 mM D-glucose (pH = 7.4). An identical buffer was prepared with the addition of 0.5% bovine serum albumin (“Krebs-BSA”). During cannulation Krebs-BSA buffer was present both luminally and abluminally; however, during the experiment the abluminal solution was constantly exchanged with plain Krebs. Ca2+-free Krebs was Krebs with 3 mM EGTA replacing CaCl2·2H2O. All chemicals were obtained from Sigma-Aldrich (St. Louis, MO), with the following exceptions: BSA (US Biochemicals; Cleveland, OH), MgSO4, Na-HEPES and YM254890 (ThermoFisher Scientific; Pittsburgh, PA), ivabradine, zatebradine, S18886 (Cat. No. 5568) and losartan (Tocris). AP811, BQ123, BIBO3304 were purchased from Tocis Bioscience. YM254890, amiloride, benzamil, ivabradine, zatebradine, BaCl2, losartan, AP811 and BQ123 were dissolved in water to make stock solutions. BIBO3304, YM254890, Ani9, tranilast and SET2 were dissolved in DMSO to make stock solutions. Each inhibitor was then further diluted in Krebs solution to reach the final concentrations stated; final DMSO concentrations were well below the threshold for direct effects.

Statistical analysis

The data were analyzed in LabVIEW, compiled in Excel and statistical analyses were performed using Prism (v.10.2; Graphpad, San Diego, CA). Statistical differences in the slopes of the F-P relationships between two groups were assessed by an unpaired, two-tailed t-test, and differences between three or more groups were tested using a one-way ANOVA with Dunnett’s post-hoc tests. Comparisons between contraction parameters as a function of pressure were made using two-way or mixed-model ANOVAs with repeated measures and various post-hoc tests as indicated in the figure legends. Data are plotted as mean ± SEM. Significance levels (compared to the control group) are either stated or indicated as follows: *p<0.05; **p<0.01; ***p<0.001; ****p<0.0001; “n.s.” p≥0.05, with the color-coding of the symbols matching the respective group. In the figure legends, N refers to the number of animals and n refers to the number of vessels per group.

Results

Role of ANO1 in pressure-induced chronotropy

We maintained the previous 19 distinct cell clusters and their labels consistent with as initially identified in the scRNAseq analyses (Zawieja et al., 2025). Cells from dissected IALVs segregated into prominent clusters/subclusters of lymphatic endothelial cells (LECs), LMCs, adventitial cells (AdvCs) and immune cells (Fig. 2A) (Zawieja et al., 2025). Two small clusters reflect contamination by mammary gland ducts, which are present within the tissue in the vicinity of the IALVs. Further analysis confirmed that Ano1 was expressed at moderate levels (relative to Acta2) in >75% of LMCs and to a lesser extent in AdvCs, but not in LECs (Fig. 2A). AdvCs contained subclusters of adventitial fibroblasts with some expressing multiple cell markers of pluripotency (Zawieja et al., 2025).

Consequences of ANO1 deletion and/or inhibition.

A) UMAP plot of Ano1 expression in the various cell populations within the walls of murine IALVs that were thoroughly cleaned prior to dissociation and scRNAseq analysis, as described in (Zawieja et al., 2025). The LMC cluster, composed of at least two subclusters of 978 cells, was identified by the expression of canonical smooth muscle cell markers Acta2 (SM α-actin), Myh11 (myosin heavy chain 11) and Itga8 (integrin alpha 8). Gray color in this and subsequent figures represents Acta2 expression, red/brown colors in this panel indicate Ano1 expression. Bubble plot shows Ano1 expression in the LMC and other cell clusters in terms of percent cells in an individual cluster expressing Ano1 (size of dot) and the Ano1 expression level relative to Acta2 (color of dot). Abbreviated names of the clusters are listed on the right. B) Slope of frequency vs pressure (F-P) for WT (C57Bl/6) vessels with / without treatment by the ANO1 inhibitor Ani9 (1 μM), for Ano1f/f control vessels, for Ano1 smKO vessels and for the latter treated with Ani9. C) Summary plots of contraction parameters as a function of pressure for the 5 respective groups of vessels. Horizontal bars indicate ranges of pressures with same specified statistical significance. WT N=15, n=20; WT+Ani9 N=5, n=6; Ano1 smKO N=14, n=23; Ano1f/f N=8, n=16; Ano1 smKO+Ani9 N=4, n=4.

Previously, we reported that ANO1 inhibition or SM-specific Ano1 deletion caused a nearly complete abrogation of pressure-induced chronotropy in mouse IALVs. However, the contraction frequency of IALVs changes only ∼1.7 fold over the pressure range from 0.5 to 10 cmH2O (Zawieja et al., 2019), compared to the much larger dynamic range (≥ 10-fold) characteristic of popliteal lymphatics (Fig. 1). Thus, we used popliteal lymphatics throughout this study to assess pressure-induced chronotropy, in anticipation that we would be able to detect even subtle changes in the response after deletion of specific genes. The F-P analysis for 20 WT popliteal lymphatic vessels is shown in Fig. 2. C57Bl/6 mice served as the WT control group for many comparisons because all transgenic strains used in the present study (with one exception) were derived from, or backcrossed into, that background. The average slope (± SEM) of the F-P relationship between 0.5 and 5 cmH2O was 2.8 ± 0.3 for WT vessels (Fig. 2B), equating to an average frequency difference of 12.1 ± 1.0 contractions·min-1 over that pressure range. The average r2 value for the linear regressions was 0.93 ± 0.02, indicating a consistently high quality of the curve fits and confirming that the slope was approximately linear between 0.5 and 5 cmH2O. The role of ANO1 in pressure-induced chronotropy of mouse popliteal lymphatic vessels was then tested after pharmacologic inhibition of ANO1 with Ani9, one of the most selective small molecule inhibitors of ANO1 (Hwang et al., 2016). Initial assessment of the Ani9 concentration-response relationship (from 30 nM to 10 μM) in WT popliteal vessels revealed that 3 μM Ani9, the concentration previously used to inhibit ANO1 in IALVs (Zawieja et al., 2019; Zawieja et al., 2023), did indeed reduce the basal contraction frequency of popliteal lymphatics (in the lower pressure range), but also caused a loss of tone and marked reduction in their contraction amplitude (not shown)—suggestive of possible off-target effects; therefore, for subsequent F-P protocols we used a lower concentration (1 μM) of Ani9, which had insignificant effects on contraction amplitude and tone (Fig. 2C) but significantly reduced the average slope of the F-P relationship from 2.8 ± 0.3 to 0.3 ± 0.2 (p<0.001; Fig. 2B).

We then tested the effects of SM-specific Ano1 deletion by comparing the F-P relationship of popliteal lymphatic vessels from tamoxifen-treated Myh11-CreERT2;Ano1f/f (Ano1 smKO) mice to Ano1f/f control mice. The results are also shown in Fig. 2B. The average slope of the F-P relationship between 0.5 and 5 cmH2O for Ano1 smKO vessels was 0.4 ± 0.2, significantly lower (p<0.001) than that (2.8 ± 0.2) for Ano1f/fvessels. Complete sets of contraction data for untreated and Ani9-treated WT popliteal lymphatics, Myh11-CreERT2;Ano1f/fand Ano1f/f popliteal vessels, are shown in Fig. 2C. The frequency of Ani9-treated vessels was significantly reduced at most pressures but increased at higher pressures. Normalized contraction amplitudes were unchanged by Ani9. Tone was slightly but not significantly higher in Ani9-treated vessels. After Ano1 deletion, frequency was reduced to <3 contractions·min-1 at all pressures, in contrast to Ani9-treated WT vessels in which frequency was comparably reduced at low pressures but not at elevated pressures (Fig. 2C). Ano1 smKO vessels had significantly elevated levels of tone at 5 of 7 pressures, suggesting that Ani9 treatment may have interfered with tone development in a way that Ano1 deletion did not. Collectively, these results show that deletion / inhibition of ANO1 significantly impaired pressure-induced chronotropy in popliteal lymphatic vessels; however, SM-specific Ano1 deletion failed to completely abolish spontaneous contractions in 16 of 21 Ano1 smKO vessels. To test the possibility that residual pacemaking activity in Ano1 smKO vessels might be due to incomplete recombination by the inducible Myh11-CreERT2 driver, Ano1 mRNA levels in IALVs were quantified using qPCR analysis. Smooth muscle cells were digested from IALVs of Myh11-CreERT2;mTmGf/f or Myh11-CreERT2;mTmGf/f;Ano1f/f mice and purified by FACS using the GFP signal. Ano1 message (referenced to α-actin levels) in LMCs from Myh11-CreERT2;mTmGf/f;Ano1f/fvessels was undetectable compared to that in LMCs from Myh11-CreERT2;mTmGf/f vessels (0 ± 0 and 1006 ± 80, respectively; mean ± SD, n=3). As a further test for possible residual ANO1 activity in Ano1 smKO vessels, we treated another group of Ano1 smKO vessels with Ani9 (1 μM), and those results are shown in Fig. 2B-C. Pacemaking activity persisted even after combined deletion and inhibition of ANO1, without significant further reductions in frequency at any pressure, suggesting that residual pacemaking in Ano1 KO vessels was not due to incomplete ANO1 knock down. Examples of the contraction patterns observed in Ano1f/f and Ano1 smKO vessels are shown in Figure 3; in the KO vessel, the basal contraction frequency was reduced to a low value (to 0.6 at P=0.5 cmH2O) and increased only to 2.2 min-1 at P=5 cmH2O. Collectively, these results point to ANO1 as the major component of pressure-induced lymphatic chronotropy but also to the involvement of one or more additional mechanisms, possibly involving other LMC ion channels.

Contractions of Ano1f/f and Ano1 smKO vessels.

A) Spontaneous contractions of an Ano1f/f (control) popliteal lymphatic vessel at various pressures. Average frequencies at each pressure are listed below the diameter traces. Frequency ranged from 2.1 min-1 at 0.5 cmH2O to 15.3 min-1 at 5 cmH2O. B) Spontaneous contractions of a Myh11-CreERT1;Ano1f/f (Ano1 smKO) popliteal vessel at various pressures. In this particular example, two additional pressure steps between 0.5 and 3 cmH2O were imposed (this example represents an early experiment before the protocol was finalized). Frequency ranged from 0.6 min-1 at 0.5 cmH2O to 2.2 min-1 at 5 cmH2O.

Role of PIEZO1

PIEZO1 and PIEZO2 are two ion channels that satisfy the full set of criteria for true mechanosensitivity (Coste et al., 2010; Davis et al., 2023d). PIEZO2 is expressed primarily in neurons (Woo et al., 2014; Woo et al., 2015) whereas PIEZO1 is more widely expressed (Murthy et al., 2017). PIEZO1 is critically important for shear-stress induced Ca2+ signaling in endothelium (Li et al., 2014a), for lymphatic valve and vessel development (Nonomura et al., 2018; Choi et al., 2019; Choi et al., 2024) and for hypertension-dependent arterial wall remodeling (Retailleau et al., 2015). The role of PIEZO1 in the arterial myogenic response has not been systematically tested, except for a single report showing that SM-specific Piezo1 deletion did not alter myogenic tone development in either caudal or rostral cerebellar arteries of mice [Suppl. Figs. 2-3 in (Retailleau et al., 2015)]. The possible involvement of PIEZO1 in lymphatic vessel contractility has not been reported. Our scRNAseq analysis revealed strong expression of Piezo1 message in LECs but very weak expression in LMCs (Fig. 4A). Piezo2 expression was barely detectable in either cell type. Popliteal lymphatic vessels from Myh11-CreERT2;Piezo1f/f (Piezo1 smKO) and Piezo1f/f control mice were subjected to the protocol for assessing pressure-induced chronotropy. However, the data revealed no significant difference in the F-P slope between Piezo1 smKO vessels and Piezo1f/fcontrol vessels (Fig. 4B), suggesting that PIEZO1 channels in LMCs are not required for pressure-induced lymphatic chronotropy. As a further test, we used Nestin-Cre to generate Nestin-Cre;Piezo1f/f (Piezo1 nestinKO) mice with constitutive deletion of Piezo1 both in smooth muscle and other cells expressing intermediate filaments. An analysis of vessel responses from Piezo1 nestinKO mice indicated that there was no significant impairment in pressure-induced chronotropy (Fig. 4B). Although we did not suspect the endothelium would be involved in pressure-induced chronotropy, we also generated and tested Prox1-CreERT2;Piezo1f/f (Piezo1 lecKO) mice. As expected, there was no significant difference in the average F-P slope of vessels from Piezo1 lecKO mice compared to tamoxifen-treated Piezo1f/fcontrols (Fig. 4B). Complete sets of contraction data for popliteal vessels from Piezo1 smKO are shown in Fig. 4C and revealed no significant differences in any of the lymphatic contractile parameters at any pressure compared to their floxed controls.

Consequences of Piezo1 deletion.

A) UMAP and bubble plots of Piezo1 and Piezo2 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Piezo1f/f, Piezo1 smKO, Piezo1 nestin KO, Piezo1 lecKO mice. C) Summary plots of contraction parameters as a function of pressure for the 4 respective groups of vessels. Piezo1f/f N=9, n=17; Piezo1 smKO N=7, n=13; Piezo1 nestin KO N=5, n=10; Piezo1 lecKO N=5, n=6.

Roles of TRPC6 and TRPC3

TRPC6 is thought to be a mechanosensitive cation channel whose activation underlies pressure-induced constriction in cerebral arteries (Welsh et al., 2002). However, studies of other arteries from Trpc6-/-mice found either no differences from control vessels or even enhanced myogenic responsiveness (Dietrich et al., 2005; Schleifenbaum et al., 2014). Because global TRPC6 deletion results in compensatory upregulation of TRPC3 (Dietrich et al., 2005), a related TRPC family member with similar properties and which is also activated by DAG, it is possible that TRPC3 could compensate for loss of TRPC6. The expression patterns for TRPC6 and TRPC3, based on scRNAseq analyses, are shown in Fig. 5A. TRPC6 was expressed at modest levels in ∼50% of LMCs and was almost exclusively confined to LMCs. TRPC3 was expressed at very low levels in only a few LMCs (from control mice). We tested the effects of global TRPC6 deletion (Trpc6-/-), global TRPC3 deletion (Trpc3-/-) and global deletion of both isoforms (Trpc6-/-;Trpc3-/-) on pressure-induced lymphatic chronotropy. As these mice were all generated in the SV129 background, control measurements were performed on popliteal vessels from SV129 mice, which showed a lower average F-P slope than WT (C57Bl/6) popliteal vessels. There were no significant differences in F-P slope between SV129 controls and Trpc6-/- or Trpc3-/- vessels, but the F-P slope of Trpc6-/-;Trpc3-/-vessels was significantly higher than that of the SV129 controls (Fig. 5B). Complete sets of contraction data for Trpc6-/-;Trpc3-/-vessels, compared to their SV129 controls, are shown in Fig. 5C. Trpc6-/-;Trpc3-/-double knock-out vessels had significantly impaired contraction amplitudes at several pressures and increased frequencies and increased levels of tone at most pressures. However, relevant to the main objective of this study, there were no significant impairments in pressure-induced chronotropy in vessels deficient in TRPC6, TRPC3 or both TRPC6 and TRPC3 (Fig. 5B-C).

Consequences of Trpc6 and/or Trpc3 deletion.

A) UMAP and bubble plots of Trpc6 and Trpc3 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from SV129 controls, Trpc6-/-, Trpc3-/- and Trpc6-/-;Trpc3-/- mice. C) Summary plots of contraction parameters as a function of pressure for the 4 respective groups of vessels. SV129 control N=8, n=11; Trpc6-/- N=5, n=10; Trpc3-/- N=7, n=11; Trpc6-/-;Trpc3-/- N=5, n=10.

Role of TRPM4

Multiple studies have implicated TRPM4 channels in the arterial myogenic response (Gonzales & Earley, 2013; Earley & Brayden, 2015; Zhu et al., 2026), but to date no published studies have examined the role of TRPM4 in lymphatic vessel contractility. scRNAseq analysis revealed that Trpm4 was expressed at modest levels in only ∼10% of LMCs, and in a subset of LECs, but at higher levels in some AdvCs (Fig. 6A). We tested the role of TRPM4 in pressure-induced lymphatic chronotropy by comparing the responses of Myh11-CreERT2;Trpm4f/f (Trpm4 smKO) vessels to Trpm4f/f vessels. The lack of significant reductions in F-P slope of Trpm4 smKO vessels (Fig. 6B) suggests that TRPM4 is not required for pressure-induced lymphatic chronotropy. Contraction data comparing Trpm4 smKO vessels to Trpm4f/f popliteal vessels are shown in Fig. 6C. There were no significant differences between the contraction parameters of Trpm4 smKO and Trpm4f/f vessels at any pressure.

Consequences of Trpm4 deletion.

A) UMAP and bubble plots of Trpm4 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Trpm4f/f and Trpm4 smKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. Trpm4 smKO N=6, n=13; Trpm4f/f N=4, n=7.

Role of PKD2

Studies by Jaggar and colleagues indicate that SM-specific deletion of PKD2 attenuates pressure-induced depolarization and myogenic tone in rat pial arteries (Narayanan et al., 2013) and blunts pressure-induced constriction in hindlimb arteries (Bulley et al., 2018). The roles of PKD2/2 channels in lymphatic vessel contractility are not known. scRNAseq analysis revealed modest Pkd2 and Pkd1 expression in LMCs and AdvCs (Fig. 7A). We tested the possible involvement of PKD2 channels in pressure-induced lymphatic chronotropy by comparing the responses of Myh11-CreERT2;Pkd2f/f (Pkd2 smKO) vessels to Pkd2f/f vessels (Fig. 7B). The lack of a significant reduction in the F-P slope for Pkd2 smKO vessels suggests that PKD2 is not required for pressure-induced lymphatic chronotropy. Contraction data comparing Pkd2 smKO vessels to Pkd2f/f popliteal vessels are shown in Fig. 7C. There were no significant differences between the contraction parameters of Pkd2 smKO and Pkd2f/f vessels at any pressure.

Consequences of Pkd2 (TRPP1) deletion.

UMAP and bubble plots of Pkd2 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Pkd2 f/f and Pkd2 smKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. Pkd2 smKO N=6, n=13; Pkd2 f/f N=4, n=8.

Role of TRPV4

In the blood vasculature, TPRV4 has been studied extensively in the context of Ca2+ influx pathways in endothelium (Chen & Sonkusare, 2020). However, TRPV4 is also expressed in VSM cells of some arteries (Mercado et al., 2014) and one report indicates that pharmacological inhibition of TRPV4 channels blunts pressure-induced VSM cell depolarization and myogenic vasoconstriction of preglomerular arteries from neonatal pigs (Soni et al., 2017). Our scRNAseq analysis revealed that TRPV4 was largely undetectable in LMCs but expressed at higher levels in almost all LECs, macrophages and some AdvCs (Fig. 8A). Nonetheless, we tested the role of TRPV4 channels in pressure-induced lymphatic chronotropy by comparing the responses of Myh11-CreERT2;Trpv4f/f(Trpv4 smKO) vessels to Trpv4f/f vessels. The lack of a significant reduction in the F-P slope for Trpv4 smKO vessels suggests that TRPV4 is not required for pressure-induced lymphatic chronotropy (Fig. 8B). Contraction data comparing Trpv4 smKO and Trpv4f/fpopliteal vessels are shown in Fig. 8C. Contraction amplitude was impaired at several pressures in Trpv4 smKO vessels, suggesting some involvement of TRPV4 channels in control of contractile strength, possibly as a result of macrophage-derived products (Schulz et al., 2025). Frequency was significantly different at two lower pressures in Trpv4 smKO vessels (Fig. 8C), where it was elevated rather than reduced. We also tested the F-P relationship in Trpv4-/- (global KO) vessels and confirmed that there was no significant impairment in pressure-induced lymphatic chronotropy (average slope = 2.7 ± 0.2; n=18) compared to WT vessels (average slope = 2.8 ± 0.3; data not shown). Collectively, these results indicate that TRPV4 channels are not required for pressure-induced lymphatic chronotropy.

Consequences of Trpv4 deletion.

UMAP and bubble plots of Trpv4 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Trpv4f/f and Trpv4 smKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. TRPV4 smKO N=5, n=10; TRPV4f/f N=3, n=7.

Role of TRPV2

TRPV2 is an osmosensitive cation channel (Nilius & Droogmans, 2001). Down-regulation of TRPV2 expression in mouse aortic VSM cells reduced the amplitude of cation currents activated by hypotonic swelling (Muraki et al., 2003). In another study, application of a hypotonic bath solution stimulated TRPV2-like whole-cell cation currents and Ca2+ influx in VSMCs isolated from rat retinal arteries—responses that were attenuated by the TRPV2 inhibitor tranilast (McGahon et al., 2016). Additionally, the administration of tranilast to rat retinal arteries resulted in loss of pre-existing myogenic tone whereas pre-incubation with a TRPV2 blocking antibody prevented the development of myogenic tone (McGahon et al., 2016). Our scRNAseq analysis revealed that TRPV2 was highly expressed in 63% of Myh11-expressing LMCs (Fig. 9A). As we did not have access to TRPV2 null or floxed mice, we tested the potential role of TRPV2 channels in pressure-induced lymphatic chronotropy by bath application of tranilast (100 μM) to WT popliteal lymphatic vessels. There was no significant reduction in the F-P slope for tranilast-treated vessels (Fig. 9B). We also tested the effects of SET2, a recently reported TRPV2 inhibitor with a lower IC50 (460 nM) than tranilast. The application of SET2 (1 μM) caused a transient cessation of spontaneous contractions for 2-5 min in 5 of 6 vessels, but the contractions returned within 10 min in the continued presence of SET2. Subsequent pressure steps revealed that SET2 did not significantly alter the F-P slope (Fig. 9B). Contraction data comparing WT and tranilast- and SET2-treated WT popliteal vessels are shown in Fig. 9C. These results suggest that TRPV2 is not required for pressure-induced lymphatic chronotropy.

Consequences of Trpv2 inhibition.

UMAP and bubble plots of Trpv2 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from WT mice with/without treatment with the TRPV2 inhibitors tranilast (10 μM) and SET2 (1 μM). C) Summary plots of contraction parameters as a function of pressure for the three respective groups of vessels. WT+tranilast N=5, n=10; WT+SET2 N=3, n=6.

Role of ENaC

The epithelial Na+ channel (ENaC), comprised of a heterotrimer under the expression of the genes Scnn1a, Scnn1b, Scnn1g, Scnn1d, is expressed in both the endothelial and smooth muscle layers of arteries. αENaC is the principal isoform in the endothelium and studies of native epithelial or heterologous channels reconstituted into bilayers suggest that αENaC is intrinsically mechanosensitive (Price et al., 2000; Benos, 2004). Work from the Drummond laboratory suggests that the βENaC isoform is expressed in VSM cells and plays a critically important role in myogenic constriction (Drummond, 2012). However, our scRNAseq analysis revealed almost no expression of αENaC subunits in LMCs and only weak expression in the LEC population (Fig. 10A); βENaC was essentially undetectable. Nevertheless, we tested the possible role of ENaC in pressure-induced lymphatic chronotropy by administering two widely used ENaC inhibitors, amiloride and benzamil, to WT popliteal lymphatic vessels. The data in Fig. 10B show that neither amiloride nor benzamil, at concentrations known to inhibit arterial myogenic constriction [5 and 1 μM, respectively (Jernigan & Drummond, 2006)], caused a significant impairment in pressure-induced lymphatic chronotropy. Contraction data for amiloride- and benzamil-treated WT popliteal vessels are shown in Fig. 10C. Both ENaC inhibitors tended to impair contraction amplitude to some degree (significant only for amiloride), with benzamil also significantly increasing tone at almost all pressure levels. However, the F-P relationships were nearly identical between WT and amiloride- or benzamil-treated vessels, reinforcing the conclusion that ENaC is not involved in pressure-induced lymphatic chronotropy.

Consequences of ENaC inhibition.

UMAP and bubble plots of ENaC subunit expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from WT mice with/without treatment with the ENaC inhibitors amiloride (5 μM) and benzamil (1 μM). WT N=15, n=20, WT+amiloride N=5, n=9; WT+benzamil N=6, n=11.

Role of NCX1

The sodium-calcium exchanger (NCX) has a well-documented role in the arterial myogenic response (Zhang, 2013) and in controlling the pacemaker activity of sinoatrial nodal cells, where it couples SR calcium oscillations, termed local calcium releases (LCRs) caused by calcium movement through ryanodine channels (Vinogradova et al., 2004; Vinogradova et al., 2005), to AP ignition. LCRs may occur throughout diastole (Monfredi et al., 2013), but their role is particularly critical in late diastole where LCRs activate NCX and the ensuing addition of inward current, and positive feedback of further calcium influx through L-type channels, provides sufficient depolarization to cross the voltage threshold for initiating AP firing (Bogdanov et al., 2001; Lyashkov et al., 2018b). This process is possibly analogous to the electrophysiological pacemaking mechanism in LMCs and a previous study demonstrated functional NCX1 in lymphatic muscle in human vessels (Telinius et al., 2015). scRNAseq analysis revealed strong expression of Slc8a, which encodes NCX1, in >80% of LMCs and expression was also noted in LECs and some immune cells (Fig. 11A). We tested the role of NCX1 in pressure-induced chronotropy by comparing the responses of Myh11-CreERT2;Slc8a1f/f (Slc8a1 smKO) and Slc8a1f/f vessels. The lack of a significant reduction in the F-P slope for Slc8a1 smKO vessels suggests that NCX1 is not required for pressure-induced lymphatic chronotropy (Fig. 11B). Contraction data for Slc8a1 smKO and Slc8a1f/f popliteal vessels are shown in Fig. 11C and reveal no significant differences between the contraction parameters of Slc8a1 smKO and Slc8a1f/f vessels.

Consequences of Slc8a1 (NCX1) deletion.

UMAP and bubble plots of Slc8a1 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Slc8a1f/f and Slc8a1 smKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. Slc8a1f/f N=7, n=12; Slc8a1 smKO N=7, n=13.

Role of HCN

HCN (Hyperpolarization-activated cyclic-nucleotide-gated) channels contribute an inward “funny” current (If) to the diastolic pacemaking potential of sinoatrial node cells (Accili et al., 2002) and are also involved in the intrinsic rhythmicity of bladder and uterine smooth muscle (Alotaibi et al., 2017; Mader et al., 2018). In LMCs they could potentially contribute to diastolic depolarization. The HCN blockers CsCl, ZD7288, and ivabradine all lower the rate of spontaneous contractions of lymphatic vessels in rat diaphragm (Negrini et al., 2016), but inhibition requires substantially higher concentrations than the documented IC50 values for HCN channel blockade [see discussion in (Davis & Zawieja, 2024)]. In contrast, all three inhibitors increase the spontaneous contraction frequency of human lymphatic vessels at concentrations near or slightly higher than the IC50 values for HCN inhibition (Majgaard et al., 2022), suggesting off-target effects in some lymphatic vessels. Our scRNAseq analysis revealed essentially no expression of any of the four HCN isoforms in LMCs (Fig. 12A), however, it is possible the expression is too low to detect without deep sequencing. Due to the proven role of HCN channels in cardiac pacemaking, we tested the role of HCN channels in pressure-induced lymphatic chronotropy of WT mouse popliteal lymphatics using the HCN inhibitor ivabradine (at a concentration of 3 μM, which is slightly higher than the IC50 of 2.5 μM), and the more effective inhibitor, zatebradine, at a concentration of 3 μM (IC50 = 480 nM). The lack of significant reductions in the F-P slope for ivabradine- or zatebradine-treated WT vessels suggests that HCN channels are not required for pressure-induced lymphatic chronotropy (Fig. 12B). Contraction data for ivabradine- and zatebradine-treated WT vessels are shown in Fig. 12C and reveal no substantial differences from WT vessels, except for a single pressure in which each HCN channel inhibitor actually increased the contraction frequency. Collectively, these results recapitulate the findings of the in-depth study by Majgaard et al (Majgaard et al., 2022) and reinforce the conclusion that HCN channels are not required for pressure-induced lymphatic chronotropy.

Consequences of HCN inhibition.

UMAP and bubble plots of Hcn gene family expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from WT mice with/without treatment with the HCN inhibitors ivabradine (3 μM) and zatebradine (3 μM). WT N=15, n=20, WT+zatebradine. N=5, n=10; WT+ivabradine N=5, n=10.

Role of Kv7.4

Kv7 channels are activated at more negative membrane potentials than Kv1-family delayed rectifier K+ channels and thus are implicated in control of VSMC excitability near the resting membrane potential (Yeung & Greenwood, 2005; Zhong et al., 2010; Mani & Byron, 2011). The Kv7 blocker, XE991, enhances myogenic tone in isolated mesenteric and renal arteries and hypotonic swelling suppresses an XE991-sensitive Kv current in patch-clamped VSM cells (Schleifenbaum et al., 2014). Collectively, these results imply that mechanosensitive inhibition of Kv7.4 may increase VSM excitability. The effects of Kv7 channel inhibition on lymphatic function are unknown. Our scRNAseq analysis revealed moderately strong expression of Kcnq4 in >50% LMCs and in some AdvCs (Fig. 13A). We tested whether pressure-induced lymphatic chronotropy was altered in Myh11-CreERT2;Kcnq4f/f (Kcnq4 smKO) mice. However, the lack of a significant reduction in the F-P slope for Kcnq4 smKO vessels suggests that Kv7.4 channels are not required for pressure-induced lymphatic chronotropy (Fig. 13B). Contraction data for Kcnq4 smKO and WT vessels are compared in Fig. 13C. There were no significant differences in any of the parameters at any pressure.

Consequences of Kcnq1 deletion.

UMAP and bubble plots of Kcnq4 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from WT and Kcnq4 smKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. Kcnq4 smKO N=5, n=7.

Role of Kir channels

Inwardly rectifying K+ channels, including Kir2.1 and Kir2.2, encoded by Kcnj2 and Kcnj12 respectively, have been implicated in shear-stress-induced responses of arterial endothelium (Ahn et al., 2017) and swelling-induced currents in arterial smooth muscle cells (Sancho et al., 2019). However, their potential roles in controlling LMC membrane potential and pressure-induced lymphatic chronotropy are unknown. Our scRNAseq analysis revealed modest expression of Kcnj8 (Kir6.1), an inward rectifier which forms the pore forming unit of the KATP channel, in >50% LMCs (Fig. 14A), but very little expression of Kcnj2 or Kcnj12 in LMCs. We previously showed that deletion of Kcnj8 does not alter the F-P relationship in popliteal lymphatics (Davis et al., 2020). Here, we tested the effects of BaCl2, a selective inhibitor of Kir2.1/2.2 at a concentration of 100 μM (Longden et al., 2017; Sancho et al., 2017; Sancho et al., 2019), on the F-P relationship of WT popliteal lymphatic vessels. Kir inhibition resulted in a significantly reduced slope of the F-P relationship compared to WT vessels (Fig. 14B), but the consistent effect of BaCl2 was to flatten the slope of the F-P relationship by substantially increasing the contraction frequency at low pressures while only slightly increasing the contraction frequency at elevated pressures (Fig. 14C). This contrasts with the effect of ANO1 inhibition or deletion: in the absence of ANO1 activity, the contraction frequency was significantly reduced at all pressures, particularly at low pressures (Fig. 2C). Thus, the inhibition of Kir channels does impair pressure-induced chronotropy but does so by differentially elevating contraction frequency as a function of pressure.

Consequences of Kir2.x inhibition.

UMAP and bubble plots of Kcnj2 (Kir2.1), Kcnj12 (Kir2.2) and Kcnj8 (Kir6.1) expression in various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from WT mice with/without treatment with the Kir2 inhibitor BaCl2 (100 μM). WT N=15, n=20, WT+BaCl2 N=4, n=11.

Roles of L- and T-type Ca2+ channels

The L-type, voltage-gated Ca2+ channel (Cav1.2) carries the majority of current contributing to the action potential in LMCs (van Helden, 1993; To et al., 2020). scRNAseq analysis confirmed that >90% of LMCs expressed message for the pore forming subunit of Cav1.2 (Cacna1c), along with other accessory subunits of the channel responsible for either gating characteristics (α2δ) or membrane targeting (β) (Fig. 15A). SM-specific knockout of Cacna1, using either the Myh11-CreERT2 or Itga8-CreERT2, eliminated essentially all propulsive contractions from popliteal lymphatic vessels (Warthi et al., 2022; Davis et al., 2023e). Residual activity in some vessels was limited to small, localized diameter oscillations (<5 μm in amplitude) with irregular frequencies that were largely independent of pressure. As expected, the slope of the F-P relationship for Cacna1c smKO vessels was significantly reduced compared to Cacna1cf/f control vessels (Fig. 15B) and frequency was significantly suppressed compared to control at all pressures (Fig. 15C).

Consequences of Cacna1c (Cav1.2) and Cacna1g/Cacna1h (Cav3.1/3.2) deletion.

UMAP and bubble plots of the expression of voltage-gated Ca2+ channels and their accessory subunits in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Cacan1cf/f, Cacna1c smKO and Cacna1g-/-;Cacna1h-/- mice. C) Summary plots of contraction parameters as a function of pressure for the three respective groups of vessels. Horizontal bar indicates range of pressures with same specified statistical significance. Cacna1c: alpha-1c, Cacna2d1: α2/δ1, Cacnb2: β2, Cacnb3: β3, Cacna1d: alpha-1d (Cav1.3), Cacna1a: Cav2.1 (P/Q), Cacna1g: Cav3.1, Cacna1h: Cav3.2. Cacna1cf/f N=10, n=20; Cacna1c smKO N=14, n=24; Cacna1g-/-;Cacna1h-/-N=16, n=25.

The T-type Ca2+ channel (Cav3.x) is critical for pacemaking in SA node (Lyashkov et al., 2018a), GI smooth muscle (Sanders, 2019), and renal pelvis smooth muscle (Grainger et al., 2022). Two Cav3 isoforms are expressed in LMCs: Cacna1g and Cacna1h, encoding Cav3.1 and Cav3.2 respectively (Lee et al., 2014; To et al., 2020). One report suggests that Cav3 channels are selectively involved in regulating the contraction frequency (but not amplitude) of rat mesenteric lymphatic collectors (Lee et al., 2014). We previously tested the roles of Cav3.1 and Cav3.2 in mouse lymphatic collectors and isolated LMCs after generating Cacna1g-/-;Cacna1h-/- double KO mice (To et al., 2020). Here, we analyzed the slope of the F-P relationship in popliteal lymphatics from Cacna1g-/-;Cacna1h-/- mice but found no significant differences from either WT or Cav1.2f/f control vessels (Fig. 15B), indicating that Cav3 channels are not involved in determining the F-P relationship. No significant differences in frequency were noted between Cacna1g-/-;Cacna1h-/-and control vessels (Fig. 15C).

Upstream pathways mediating pressure-induced chronotropy

Having established that ANO1, out of all the cation and K+ channels tested, is the only channel mediating pressure-induced lymphatic chronotropy, we then turned to an investigation of mechanosensitive pathways upstream from ANO1 channel activation.

Role of IP3R in pressure-induced chronotropy

Calcium release as a consequence of IP3 generation is known to activate a number of ion channels and other proteins, including ANO1. We recently showed that genetic deletion of Itpr1 (encoding IP3R1) from mouse IALVs results in a reduction in spontaneous contraction frequency and a blunting of pressure-induced chronotropy similar to what is observed after Ano1 deletion (Zawieja et al., 2023). scRNAseq analysis confirmed expression of Itpr1 in >50% of LMCs (Fig. 16A). Here, we examined the consequences of Itpr1 deletion in popliteal lymphatics by comparing the F-P relationship for vessels from Myh11-CreERT2;Itpr1f/f(Itpr1 smKO) and Itpr1f/f mice. SM-specific deletion of Itpr1 resulted in a significant reduction in the slope of the F-P relationship from 2.6 ± 0.3 to 0.6 ± 0.1 (Fig. 16B). Contraction data for Itpr1 smKO and Itpr1f/f control vessels are shown in Fig. 16C. The most striking features are 1) the significant reductions in frequency at all pressures and 2) a significant reduction in tone at elevated pressures in Itpr1 smKO vessels. The latter effect contrasts with the increases in tone observed after ANO1 inhibition or deletion (Fig. 2C). Collectively, these results show that IP3R1 is critically important for pressure-induced lymphatic chronotropy in popliteal collecting vessels in much the same way that it was previously documented in IALVs. Deletion of Itpr1 in LMCs produces an effect on pressure-induced chronotropy similar to that of ANO1 deletion. Examples of the contraction patterns observed in Itpr1f/fand Itpr1 smKO vessels are shown in Figure 17; in the KO vessel the basal contraction frequency was reduced to zero at P=0.5 cmH2O and increased only to 1.9 min-1 at P=5 cmH2O.

Consequences of Itpr1 (IP3R1) deletion.

UMAP and bubble plots of Ip3r1 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Itpr1f/f and Itpr1 smKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. Horizontal bar indicates range of pressures with same specified statistical significance. Itpr1 smKO N=5, n=8; Itpr1f/f N=5, n=10.

Contractions of IP3R1f/f and IP3R1 smKO vessels.

A) Spontaneous contractions of an IP3R1f/f(control) popliteal lymphatic vessel at various pressures. Average frequencies at each pressure are listed below the diameter traces. Frequency ranged from 7.1 min-1 at 0.5 cmH2O to 17.9 min-1 at 5 cmH2O. B) Spontaneous contractions of an Myh11-CreERT2;IP3R1f/f (IP3R1 smKO) popliteal lymphatic vessel at various pressures. Frequency ranged from 0 min-1 at 0.5 cmH2O to 1.9 min-1 at 5 cmH2O.

Role of GNAQ/GNA11 in pressure-induced chronotropy

Evidence from multiple studies specifically points to GNAQ/GNA11-coupled GPCRs as the upstream mechanosensing elements of myogenic constriction in arteries (Mederos y Schnitzler et al., 2008; Blodow et al., 2014; Harraz et al., 2014; Li et al., 2014b; Schleifenbaum et al., 2014; Pires et al., 2017; Bjorling et al., 2018). Mechanosensitive activation of one or more GPCRs independent of ligand binding (Zou et al., 2004; Erdogmus et al., 2019) could account for pressure-induced increases in IP3 levels in LMCs. Our scRNAseq analysis revealed strong expression of Gnaq and Gna11 in ∼75% of LMCs, LECs, and AdvCs as well as very strong expression of Gnaq in some immune cell clusters (Fig. 18A). We tested the effects of SM-specific Gnaq knock out (Myh11-CreERT2;Gnaqf/f), global Gna11 knock out (Gna11-/-) and Gnaq/Gna11 double knock out (Myh11-CreERT2;Gnaqf/f;Gna11-/-) on pressure-induced lymphatic chronotropy. There was a trend for Gnaq smKO vessels to have a reduced slope in the F-P relationship in comparison to Gnaqf/f;Gna11+/+control vessels, but it did not reach statistical significance (p=0.0819). In contrast, global deletion of Gna11 significantly reduced the slope of the F-P relationship and the reduction was even more significant and pronounced in Gnaq/Gna11 double KO vessels (Fig. 18B). However, even combined knock out of both proteins did not cause an attenuation of pressure-induced chronotropy that was comparable to that observed after SM-specific Itpr1 or Ano1 deletion, particularly at pressures ≥ 3 cmH2O (compare Fig. 18B to Figs. 2B, 16B). Contraction data for Gnaq smKO, Gna11-/-and Gnaq/Gna11 double KO vessels, compared to tamoxifen-treated Gnaqf/f;Gna11+/+controls, are shown in Fig. 18C. The full F-P relationships for each genotype showed significant reductions in frequency at a few pressures for Gnaq smKO vessels compared to Gnaqf/f;Gna11+/+ controls, significant differences for Gna11-/- vessels at more pressures and significant differences at all pressures for Gnaq/Gna11 DKO vessels. Collectively, these results suggest that both GNAQ and GNA11-coupled G-proteins are involved in pressure-induced lymphatic chronotropy, with G11 being more important. However, the observation that frequency continued to increase somewhat at higher pressures when Gq/11 signaling is abrogated points to the possible contribution of other mechanisms. Examples of the lymphatic contraction patterns at each pressure in a Gnaqf/f;Gna11+/+ control vessel and a Gnaq/Gna11 DKO vessel are shown in Figure 19; in the DKO vessel the basal contraction frequency was reduced to zero at P=0.5 and 1 cmH2O and increased only to 1.3 min-1 at P=5 cmH2O.

Consequences of Gnaq and/or Gna11 deletion.

UMAP and bubble plots of Gnaq and Gna11 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Gnaqf/f;Gna11+/+, Myh11-CreERT2;Gnaqf/f (Gnaq smKO), Gna11-/- and Myh11-CreERT2;Gnaqf/f;Gna11-/- DKO mice. C) Summary plots of contraction parameters as a function of pressure for the four respective groups of vessels. Gnaqf/f;Gna11+/+ control N=3, n=6; Gnaq smKO N=4, n=8; Gna11-/- N=3, n=6; Myh11-CreERT2;Gnaqf/f;Gna11-/-N=8, n=13.

Examples of contractions of Gnaqf/f;Gna11+/+ and Myh11-CreERT2;Gnaqf/f;Gna11-/- DKO vessels.

A) Spontaneous contractions of a Gnaqf/f;Gna11+/+ (control) popliteal lymphatic vessel at various pressures. Average frequencies at each pressure are listed below the diameter traces. Frequency ranged from 1.6 min-1 at 0.5 cmH2O to 16.2 min-1 at 5 cmH2O. B) Spontaneous contractions in a Myh11-CreERT2;Gnaqf/f;Gna11-/- DKO popliteal vessel at various pressures. Frequency ranged from 0 min-1 at 0.5 cmH2O to 1.3 min-1 at 5 cmH2O. The noisy diameter signal at P=8, 10 cmH2O is caused by tracking glitches and does not represent high frequency contractions.

Role for GNA12/GNA13 in pressure-induced chronotropy

GNA12/GNA13-coupled GPCRs are known to regulate arterial smooth muscle contraction through Rho/Rho kinase (Bolz et al., 2003; Dubroca et al., 2005; Gokina et al., 2005; Jarajapu & Knot, 2005; Li & Brayden, 2017; McDuffie et al., 2024) and at least one study suggests that Rho kinase is essential for phasic lymphatic contractions (Kurtz et al., 2014). scRNAseq analysis revealed weak expression of Gna12 and Gna13 in ∼50% of LMCs, with stronger expression of Gna13 in LECs and AdvCs (Fig. 20A). Chennupati et al. (Chennupati et al., 2019) made the surprising finding that GNA12/GNA13-coupled GPCRs, rather than GNAQ/GNA11-coupled GPCRs, mediated myogenic constrictions in mesenteric and cerebral arteries of mice. We investigated the effects of deleting both Gna12 and Gna13 from LMCs (Myh11-CreT2;Gna12f/f;Gna13-/-mice) on the F-P relationship of popliteal lymphatics. In the Gna12/Gna13 double knock out, contraction amplitude was significantly blunted and tone was significantly elevated at several lower pressures, but the F-P relationship was nearly identical to that of vessels from control Gna12f/f;Gna13+/+mice (Fig. 20B) and frequencies of Gna12/Gna13 DKO vessels were nearly identical to those of controls at all pressures (Fig. 20C).

Consequences of Gna12/Gna13 deletion.

UMAP and bubble plots of Gna12 and Gna13 expression in the various IALV cell clusters. B) Slope of F-P relationship for popliteal lymphatics from Gna12f/f;Gna13+/+ and Myh11-CreERT2;Gna12f/f;Gna13-/- DKO mice. C) Summary plots of contraction parameters as a function of pressure for the two respective groups of vessels. Gna12f/f;Gna13+/+ controls N=5, n=10; Myh11-CreERT2;Gna12f/f;Gna13-/-N=7, N=15.

Roles of specific GPCRs in pressure-induced chronotropy

The findings in Fig. 18 strongly implicate Gq/11 signaling in pressure-induced lymphatic chronotropy but leave open the question of which specific GPCR(s) mediate(s) this mechanosensitive response. An scRNAseq analysis revealed that 136 GPCRs were observed at various levels in mouse LMCs with a threshold of a minimum 0.5% LMCs. We constructed a bubble plot showing the top 20 of these GPCRs, ranked according to their relative expression across LMCs (Fig. 21, lines 11-30) and included GPCRs implicated in mechanosensation in the arterial myogenic response (32-41). That list includes the NPY1 receptor (Npy1r, line 11), the endothelin type A receptor (Ednra, line 13), the atrial natriuretic peptide type C receptor (Npr3, line 19), the thromboxane A2 receptor (Tbxa2r, line 20), the histamine type1 receptor (Hrh1 line 40) and the sphingosine-1-phosphate receptor 1 (S1pr1, line 38). Several of the more highly expressed GPCRs are members of the adhesion GPCR subfamily (Adgrl1, Adgrl2, Adgre5, Adgrd1, Adgrl3, Adgra2; rows 15, 12, 17, 24-26), and two more are orphan receptors (Gprc5b, Olfr1033; rows 27,18). Both subtypes of AT1R are expressed, but only at very low levels <1% (lines 32-33).

Bubble plot representing the results of scRNAseq analysis of the most highly expressed GPCRs in the various cell types.

Rows 1-10 show the canonical markers for LMCs (rows 1-4), LECs (rows 5-7) and AdvCs (rows 8-10), respectively. Rows 11-30 show the top 20 GPCRs expressed in LMCs, ranging from 87.5% of LMCs expressing Npy1R to 36.6% LMCs expressing F2R. Putative mechanosensitive GPCRs in other cell types (e.g. arterial SM), but with very low or absent expression in LMCs, are shown in rows 34-41; AT1R subtypes are shown in rows 32-33, cysteinyl leukotriene receptor subtypes in rows 34-35 and P2Y receptor subtypes in rows 36-37. Four other GPCRs are shown in rows 38-41. G-coupling proteins are shown in rows 42-47 and β-arrestins in rows 48-49. Red arrows at left are the GPCRs that were tested with inhibitors in the present study or in previous studies (see text for references).

We proceeded to experimentally test the involvement of several of the above GPCRs for which suitable receptor antagonists were available. In lieu of evaluating the complete pressure-frequency response for each GPCR/antagonist, we determined whether blocking the GPCR would alter pacemaking by lowering the contraction frequency at P=3 cmH2O. As confirmation that this procedure was sufficiently sensitive to detect an impact on normal pacemaking frequency, we compared the frequency at 3 cmH2O for the ANO1 smKO, IP3R1 smKO and GNAQ/GNA11 DKO vessels and their respective controls (Fig. 22, gray symbols). We also performed this analysis for WT vs Ani9 vessels and for Gq+/+;G11+/+ control vessels before and after treatment with the highly selective and potent Gq/11 antagonist YM254890 (Takano et al., 2025). As expected, each of those comparisons was highly significant, indicating that the inhibitor or knock out reduced contraction frequency. Likewise, YM254890 significantly reduced the frequency of Gq+/+;G11+/+ vessels, providing further confirmation that Gq/11 signaling is critically important for the chronotropic response. However, none of the GPCR antagonists, including those for the ANP receptor, the ET1A receptor, the NPY1 receptor, the thromboxane A2 receptor, or S1PR1, produced a significant reduction in frequency (Fig. 22, red symbols). We also compared the responses of vessels from AT1aR-/- mice and from WT mice treated with losartan, which blocks both AT1aR and AT1bR; in both the KO and the WT with losartan there was actually a significant increase in contraction frequency. We conclude that none of these seven GPCRs are critical for pressure-induced chronotropy of mouse collecting lymphatic vessels.

Screening the effects of selective GPCR inhibitors.

Pacemaking frequency is compared between control and inhibitor (or knock out) at a single pressure, 3 cmH2O. Gray symbols: comparisons between Ano1f/fcontrols and Ano1 smKO, WT ± Ani9, IP3R1f/f controls and IP3R1 smKO, Gqf/f;G11+/+controls and Gq/11 DKO, Gq+/+;G11+/+ controls ± YM254890, showing that each treatment produced a very significant reduction in frequency using this test. Red symbols: comparisons between WT controls ± respective inhibitors for the various GPCR. None of the GPCR inhibitors produced a significant decrease in frequency, but the S1PR1 and AT1R inhibitors instead significantly increased frequency. Inhibitor concentrations were as follows: Gq/11 inhibitor YM254890 (100 nM); ANPR inhibitor AP811 (1 μM); ET1AR inhibitor BQ123 (1 μM); NPYR inhibitor BIBO3304 (1 μM); TXA2R inhibitor S18886 (20 nM); S1PR1 inhibitor SEW-2871 (1 μM); AT1R inhibitor losartan (10 μM). Comparisons between controls and KO vessels were made using 2-way ANOVAs with Dunnett’s post-hoc tests. Comparisons between controls ± inhibitors were made using paired t-tests. Ano1 smKO N=14, n=23; Ano1f/f N=8, n=16. WT N=15, n=20; WT ± Ani9 N=5, n=6. Itpr1 smKO N=5, n=8; Itpr1f/f N=5, n=10. Gnaqf/f;Gna11+/+control N=3, n=6; Myh11-CreERT2;Gnaqf/f;Gna11-/- N=8, n=13. Gq+/+;G11+/+ control ± YM254890 N=8, n=8. WT ± AP811 N=8, n=8. WT ± BQ123 N=7, n=7. WT ± BIBO3304 N=8, n=8. WT ± SEW-2871 N=4, n=4. WT ± losartan N=5, n=10. Agtr1a-/- N=3, n=6. S18886 N=4; n=11.

Discussion

Lack of role for mechanosensitive ion channels in pressure-induced lymphatic chronotropy

The major goal of this study was to investigate the mechanistic basis of pressure-induced lymphatic chronotropy. Because little was previously known regarding the mechanisms of mechanosensitivity underlying this phenomenon, we took cues from a well-studied mechanosensitive mechanism in vascular smooth muscle—the arterial myogenic response. In both arteries and collecting lymphatic vessels (Davis et al., 2009; Davis et al., 2023d), pressure elevation leads to vascular muscle cell depolarization and constriction, reflecting the activation of a depolarizing current that triggers Ca2+ entry through the opening of L-type, voltage-gated Ca2+ channels. Several ionic mechanisms have been previously implicated in this process for arterial smooth muscle, specifically the activation of TRPC6, TRPM4, PKD1/2, TRPV2 and ENaC cation channels (Welsh et al., 2002; Earley et al., 2004a; Jernigan & Drummond, 2005; Sharif-Naeini et al., 2009; McGahon et al., 2016; Bulley et al., 2018), with supporting evidence for each obtained from gene knockdown / knockout studies. Additional ion channels may also be involved, including the activation of channels promoting Cl- efflux (Boedtkjer et al., 2008) and the inhibition of constitutively active K+ channels (Sancho et al., 2019), but evidence for those is based largely on the use of pharmacological inhibitors and therefore subject to concerns about off-target effects (Maroto et al., 2005; Dietrich et al., 2007; Gottlieb et al., 2008; Sancho et al., 2019). In the present study, we primarily used genetic knock out approaches to test the possible roles of the major ion channels contributing to the mechanosensitive chronotropy of lymphatic vessels from the mouse. Our results strongly support the conclusion that pressure-induced lymphatic chronotropy does not involve acute activation of any of the putative mechanosensitive (and/or second-messenger gated) cation channels previously implicated in myogenic constriction (TRPC6, TRPM4, PKD2, TRPV2, ENaC), nor does it require PIEZO1, the bona fide mechanosensitive cation channel implicated in other aspects of mechanotransduction in the cardiovascular system [see references in (Davis et al., 2023d)]. Instead, our findings support a critical role for calcium dynamics regulating ANO1 in pressure-induced lymphatic chronotropy. ANO1 is a channel that is also required for the generation of arterial myogenic tone (Bulley et al., 2012). Studies of mammalian ANO1 indicate that calcium store release is the primary mode of ANO1 regulation (Xiao et al., 2011; Dulin, 2020) and our observation that IP3R1 KO has an almost identical effect on pressure-induced chronotropy as ANO1 deletion supports the conclusion that IP3-induced calcium release from SR stores through IP3R1 drives ANO1 activation in LMCs (Zawieja et al., 2023). Our further observation that Gq/11 DKO or inhibition produces nearly equivalent effects on pressure-induced chronotropy as ANO1 or IP3R1 knock out—especially at lower pressures—are consistent with pressure-induced phospholipase activation and IP3 production (Narayanan et al., 1994) that is driven by an upstream mechanosensor. Based on an extensive body of literature in cardiomyocytes (Yasuda et al., 2008) and arterial smooth muscle (Schleifenbaum et al., 2014; Cui et al., 2022), mechanotransduction of stretch/pressure is mediated by one or more GNAQ/GNA11-coupled GPCRs, leading to downstream activation of second messenger-gated channels. In the case of LMCs, mechano-activation of GNAQ/GNA11-coupled GPCRs would promote IP3 production that triggers Ca2+ release from stores through IP3R1; Ca2+ then activates ANO1 Cl- channels, which accelerate the rate of spontaneous diastolic depolarization, increasing the firing rate of action potentials that control phasic lymphatic contractions. Repolarization is controlled in part by Kv channels (Cotton et al., 1997), including Kv11 channels (Kim et al., 2023). As summarized in Table 1, knockout of each element in the Gq/11-IP3-IP3R1/Ca2+-ANO1-Cav1.2 pathway (and only those elements out of all that were tested, except Kcnj2/ Kcnj12, which can be alternatively explained) abrogated or significantly attenuated pressure-induced chronotropy.

Although we tested the requirement for 16 potential mechanosensitive ion channels in pressure-induced lymphatic chronotropy, our list was not exhaustive. Many channels have some degree of intrinsic mechanosensitivity [for review see (Davis et al., 2022)] but mechanosensitivity for most has been demonstrated only under quite artificial conditions (e.g., patch clamp + localized membrane suction or hypoosmotic swelling) in which the mechanical stimulus is unlikely to accurately mimic the magnitude and specificity of force experienced by that channel in a cell in vivo (Davis et al., 2022). The list of bona fide mechanosensitive channels is very short if one adheres to the criteria proposed by Patapoutian and colleagues (Ranade et al., 2015). Other classes of ion channels can be activated by cell volume changes, but the volume of an LMC is unlikely to substantially change during the small pressure steps that produce increased chronotropy. Nevertheless, we also tested the potential roles of selected volume activated channels implicated in arterial myogenic constriction, including TRPV2, TRPV4 and Kir (McGahon et al., 2016; Soni et al., 2017; Sancho et al., 2019), but not others reported to be important in other cell types or in other physiological contexts (Matchkov et al., 2015). Another channel known to influence LMC pacemaking that was not explicitly tested here is the voltage-gated sodium channel (VGNaC), Nav1.3. VGNaCs are present and active in LMCs, with Nav1.3 most consistently being expressed (Telinius et al., 2015; Zawieja et al., 2024; Arroyo-Ataz et al., 2025; Ruscic et al., 2025). VGNaC contributions to pacemaking appear to vary by species, in part because of differences in the resting membrane potential (Davis & Zawieja, 2024). Telinius et al. found that APs and spontaneous contractions were inhibited by the selective VGNaC blocker TTX in 60% of human mesenteric lymphatics (Telinius et al., 2015). In contrast, phasic contractions of guinea pig and sheep mesenteric lymphatics continued in the presence of TTX (Chan & von der Weid, 2003; Chan et al., 2004; Beckett et al., 2007). We previously found that TTX had no detectable effect on phasic contractions in ∼50% of mouse popliteal lymphatic vessels and only a transient inhibitory effect (lasting only 1-4 minutes) in the continued presence of the toxin in the other ∼50% (Davis et al., 2023c). In mouse IALVs, a recent study reported similar, non-detectable effects of TTX (Schulz et al., 2026) and Padera and colleagues report that spontaneous lymphatic pumping persists in popliteal lymphatics of Nav1.3 KO mice (Ruscic et al., 2025). Thus, Nav1.3 does not appear to be critical for pressure-induced chronotropy of mouse lymphatic vessels.

Roles for TRP channels in other aspects of LMC contractile function

If IP3 production in LMCs is enhanced after pressure elevation, as it is in arterial smooth muscle (Narayanan et al., 1994; Mederos y Schnitzler et al., 2008), then concomitant increases in DAG, the other product of PIP2 hydrolysis, would also be predicted to occur, and DAG should then activate TRP6C and related channels. However, if this occurs in popliteal lymphatic muscle, the activation of those channels apparently does not significantly influence the F-P relationship (Fig. 4). Likewise, the Ca2+ increase resulting from mechanosensitive activation of the Gq/11-IP3-IP3R1 signaling axis would potentially activate TRPM4 and/or PKD2 (assuming those channels are localized to the appropriate LMC microdomains), but we find that their deletion does not significantly alter the F-P relationship (Figs. 5-6). Because TRPM4, PKD2 and PIEZO1 expression levels are very low in LMCs (Figs. 3A, 4A, 5A), further activation of those channels may have little effect. It is possible that another unidentified ion channel could be mediating the pressure-driven frequency changes at higher pressures—perhaps a channel with a higher mechanosensitive threshold or requiring higher levels of IP3, DAG or Ca2+. Candidates include TRPV1, which has also been implicated in arterial myogenic constriction (Phan et al., 2022).

Although the deletion/inhibition of several TRP channel family members and other channels did not significantly impact pressure-induced chronotropy, there was sometimes an impact on other aspects of lymphatic contractile function, namely contraction amplitude and/or tone. The respective channels and their statistically significant contraction effects are summarized in Table 2. Deletion of TRPC6/TRPC3 impaired contraction amplitude by ∼40% and increased tone by ∼2.5-fold (Fig. 4C). Deletion of TRPV4 from LMCs impaired contraction amplitude by ∼45% without an effect on tone (Fig. 7C). TRPV4 channel activation has recently been implicated in the regulation of lymphatic contraction, but primarily via the production of vasoactive products from LECs and resident macrophages (Schulz et al., 2025). Inhibitors of three other ion channels also led to significant changes in amplitude and/or tone. Specifically, one of the two TRPV2 inhibitors (SET2) inhibited tone at 4 of the 5 lowest pressures. One of the two ENaC inhibitors (amiloride) inhibited amplitude by 20-30% while the other one (benzamil) enhanced tone by ∼2-fold (Fig. 9C). The Kir2 blocker Ba2+ impaired amplitude by ∼40% and increased tone by 2-fold (Fig. 13C). In the latter case the effects of Ba2+ on tone and frequency are likely to be caused by inhibition of endothelial Kir channels, which control the production of inhibitory vasoactive products (Scallan & Davis, 2013; Davis et al., 2022; DuToit et al., 2024; Davis & Bertram, 2025). Collectively, these observations suggest that the respective channels are each involved in some aspect of lymphatic contractile function (other than pressure-induce chronotropy), with the more compelling evidence being obtained after gene deletion rather than after acute application of soluble inhibitors, which may have off-target effects.

Pressure-induced lymphatic chronotropy requires intact Gq/11 signaling

Our data using mice with SM-specific knock out of Gq and/or global knock out of G11 suggest that both Gq and G11 contribute to pressure-induced chronotropy. The F-P slope of popliteal lymphatics was significantly reduced in G11-/- vessels and more so in Gq/G11 DKO vessels (Fig. 18), as well as in WT vessels treated with the specific Gq/G11 inhibitor YM254890 (Fig. 22). Although the reduction in F-P slope for Gq smKO vessels did not reach statistical significance (Fig. 18A), frequencies at 3 of 7 pressures were significantly lower than in Gqf/fcontrols (Fig. 18C). Additionally, the F-P slope for Gq/11 DKO vessels was lower than that of G11-/- vessels, so it appears that the effects of knocking out Gq and G11 are cumulative, suggesting that Gnaq deletion contributes to the lower F-P slope of Gq/11 DKO vessels. Further support for this conclusion is provided by a similar analysis of IALVs in which Gq/11 DKO vessels had significantly lower frequencies than G11-/-vessels at 3 of the 4 lowest pressures (Suppl. Fig. 1). Further inspection of Figure 18C suggests that frequency is impaired at high pressures in Gq smKO vessels whereas frequency is impaired at low pressures in G11-/-vessels, so the two G-coupled proteins may associate with GPCRs that sense different pressure ranges.

Our conclusions are consistent with evidence that Gq/11 coupled GPCRs are required for mechanotransduction of the arterial myogenic response. Pressure-induced constriction in several types of arteries is significantly attenuated or even completely blocked by inverse agonists of AT1R (Mederos y Schnitzler et al., 2008; Schleifenbaum et al., 2014; Storch et al., 2015; Hong et al., 2016) and/or by knock-out of the specific receptor subtypes AT1AR (Schleifenbaum et al., 2014; Cui et al., 2022) or AT1BR (Pires et al., 2017). However, although AT1R may be a critical mechanosensitive GPCR in many arteries, other studies have implicated additional GPCRs, including the cysteinyl leukotriene 1 receptor (Storch et al., 2015; Mederos et al., 2016) and the P2Y4 and P2Y6 receptors (Brayden et al., 2013; Forst et al., 2016; Kauffenstein et al., 2016). None of those GPCRs are likely candidates to mediate pressure-induced lymphatic chronotropy as they are all expressed at extremely low levels in only a small fraction of LMCs (Fig. 21, lines 34-37). Likewise, knock out of AT1AR did not significantly alter the lymphatic vessel F-P relationship, nor did the AT1R antagonist losartan, which inhibits both AT1AR and AT1BR subtypes (Fig. 22), indicating that AT1R is not required for this aspect of lymphatic mechanotransduction. Other Gq/11-coupled GPCRs in VSMCs that are potentially mechanosensitive include S1PR1, B2R, H1R, M5R and V1AR (Mederos y Schnitzler et al., 2008; Peter et al., 2008). We tested the S1PR1 inhibitor, SEW 2871, which did not inhibit contraction frequency but instead increased it (Fig. 22). B2R, M5R and V1AR are expressed at very low levels in LMCS, and inhibition of H1R in two other studies failed to inhibit spontaneous lymphatic contractions (Fox & von der Weid, 2002; Nizamutdinova et al., 2014).

In arteries, blockade of Gi or Gs signaling failed to inhibit mechanosensitive VSM responses (Mederos y Schnitzler et al., 2008). However, an exception to Gq/11-mediated mechanotransduction in arterial smooth muscle is a study by Chennupati et al. (Chennupati et al., 2019) who found that G12/13 but not Gq/11 signaling was critical for myogenic responses of mouse mesenteric and cerebral arteries. To our knowledge, the conflict between the latter result and the body of literature supporting a Gq/11 mechanism (Mederos et al., 2016), even with respect to specific types of arteries, has not been adequately resolved. A requirement for G12/13 signaling fits with multiple studies suggesting a critical role for Rho/Rho kinase (downstream from G12/13) in arterial myogenic constriction (Bolz et al., 2003; Dubroca et al., 2005; Gokina et al., 2005; Jarajapu & Knot, 2005; Li & Brayden, 2017) and even for phasic lymphatic contractions (Kurtz et al., 2014). However, our finding that G12/13 DKO vessels exhibit normal pressure-induced lymphatic chronotropy adds further support for a Gq/11-coupled mechanism rather than a G12/13-coupled mechanism. We cannot rule out a role for G14 participating in Gq/11 associated GPCRs as Gna14 is expressed in 28% of LMCs in our dataset (Fig. 21). Further experiments testing whether each G protein is activated at differing pressure thresholds of the pressure range, or how gene dosage of Gq/11/14 regulates pressure dependent chronotropy are required. Lastly, GPCRs are known to couple promiscuously, so there is likely crosstalk and overlap between mechanosensitive GPCRs and their coupling proteins in any given tissue (Hauser et al., 2022).

We were unable to identify a key GPCR in lymphatic muscle that mediates IP3/Ca2+-activation of ANO1 in response to pressure elevation. Our scRNAseq analysis of the GPCRs expressed in mouse LMCs identified 136 GPCRs, and we list the top 20 by % expression in LMCs (rows 11-32 in Fig. 21). We then experimentally tested 7 of the most likely candidates for which inhibitors were available by determining whether blocking the GPCR would alter pacemaking by lowering the contraction frequency at P=3 cmH2O; the results for each of those GPCRs were negative (Fig. 22). Two additional GPCRs (HT-2R, HRH1) can be ruled out by previous experiments showing that pacemaking persisted in the presence of their respective receptor blockers (Chan & von der Weid, 2003; Nizamutdinova et al., 2017). We conclude that none of these nine GPCRs (NPY1R, ETAR, NPR3, TBAX2R, HT-2R, HRH1, S1PR1, AT1AR or AT1BR), are critical for pressure-induced chronotropy in LMCs (Fig. 21, red arrows at left). Of the remaining GPCRs expressed at modest to high levels in mouse LMCs (Fig. 21), several are adhesion receptors Adgrl1, Adgrl2, Adgre5, Adgrd1, Adgrl3, Adgra2), two are orphan receptors (Gprc5b, Olfr1033), one codes for the GABA-B1 receptor (Gabbr1, with unknown function in the lymphatic system), two code for proteins in the Wnt signaling pathway (Fzd4, Fzd2), one codes for a protein in the Hedgehog signaling pathway (Smo), and one encodes a beta-amyloid binding protein (Tm2d1). The possible roles of these GPCRs, and the other 100+ GPCRs expressed at much lower levels remain to be investigated in future studies, but in many cases testing will be limited by the lack of known ligands/antagonists. We also cannot rule out a GPCR-independent mechanism involving Gq/11-coupled proteins (Dela Paz et al., 2017), although the evidence for such a mechanism is weak (Cui et al., 2022).

Lack of role of the endothelium and endothelial cell GPCRs in pressure-induced chronotropy

Is there a possible role for mechanotransduction of pressure by GPCRs in LECs? Some early vascular studies suggested that pressure-induced arterial constriction was mediated by factors released from the endothelium (Harder, 1987; Harder et al., 1989; Harder et al., 1990), however subsequent work in multiple laboratories was unable to confirm this as a general mechanism (Kuo et al., 1990; Davis et al., 2022). Perhaps the best evidence for mechanosensitive GPCRs comes from studies of the endothelium in which multiple Gq/11 coupled GPCRs, including H1R, A2AR, GPR68 and BDKRB2, each of which has been implicated in shear-stress dependent calcium signaling (Chachisvilis et al., 2006; Xu et al., 2018; Erdogmus et al., 2019), which drives production of BEC-mediated vasoactive products (Davis et al., 2022). As in VSMCs, Gi and Gs-coupled GPCRs appear not to be involved (Gudi et al., 1996; Gudi et al., 1998). Elegant experiments with cultured BECs expressing truncated or modified GPCRs point to Helix 8 in the C-terminus as the common mechanosensitive GPCR segment as GPCRs without Helix 8 are not mechanosensitive and transfer of Helix 8 to non-mechanosensitive GPCRs endows mechanosensitivity (Erdogmus et al., 2019). Interestingly, several of the mechanosensitive GPCRs in BECs listed above are also expressed in mouse LECs (Fig. 21), so is it possible that LECs are mediating pressure-induced chronotropy? Oscillatory shear stress is a known driver of several LEC signaling processes requiring GPCRs, including the adrenomedullin receptor, Calcrl, S1PR1 and Plexin D (Fritz-Six et al., 2008; Geng et al., 2025; Pang et al., 2026). Laminar shear stress responses in blood endothelium are also mediated by GPCRs, including PECAM-1, VE-Cadherin, VEGFR2 and Plexin D, but those responses control longer term processes such as cell reorientation and NF-κB activation (Tzima et al., 2005; Mehta et al., 2020). We were unable to test whether an intact lymphatic endothelium is required for pressure-induced chronotropy in mouse popliteal lymphatic vessels because it is nearly impossible to selectively remove the endothelial layer from these small vessels, which contain one-way valves, without damaging the overlying single layer of LMCs. However, studies using larger lymphatic vessels in other species or regions have previously demonstrated that lymphatic pacemaking and/or stretch-induced chronotropy remain intact after partial/complete removal or disruption of the endothelium using an air bolus (Fox & von der Weid, 2002; Ferrusi et al., 2004; DuToit et al., 2024) or thin wire (Zhang et al., 2007; von der Weid et al., 2014). Additionally, pulsatile shear stress (which is the only type of shear stress present under the conditions of our in vitro experiments) is inhibitory rather than excitatory to both the frequency of lymphatic contractions and tone (Gasheva et al., 2006; Davis & Bertram, 2025)—opposite the response needed to explain pressure-induced chronotropy. We also found that LEC-specific deletion of Piezo1, the mechanosensitive ion channel mediating shear stress-induced calcium entry in arterial endothelium (Li et al., 2014a; Rode et al., 2017), did not significantly impair pressure-induced lymphatic chronotropy (Fig. 4). Additionally, activation of Piezo1 with Yoda resulted in a nitric oxide dependent inhibition of IALV contractions (Choi et al., 2024). Nor did endothelium-specific deletion of Trpv4 (Schulz et al., 2025), a Ca2+-activated, Ca2+-permeable channel providing additional calcium entry after Piezo1 activation (Swain & Liddle, 2021; Davis et al., 2022).

Residual pacemaking and partial rescue of the F-P relationship at high pressures

Deletion/blockade of ANO1, IP3R1 or Gq/11 was associated with the complete abolition of pacemaking and phasic contractions at low pressures (0.5-1 cmH2O) in ∼50% of popliteal lymphatic vessels (Fig. 16, 18; F-P slope = 0). In other vessels, the pacemaking rate dropped to a low level that was nearly independent of pressure (Fig. 2; F-P slope <0.5) while in others, pacemaking frequency dropped to a low level but still increased to a fraction of its normal level when pressure was elevated to 8 or 10 cmH2O (F-P slope = 0.5-2). The residual pacemaking observed after removal of ANO1 activity likely represents the constitutive interactions between depolarizing current contributed by L-type Ca2+ channels and hyperpolarizing current contributed by K+ channels, as elucidated in recent models of lymphatic pacemaking (Hancock et al., 2022, 2023). These models describe coupled oscillations in the membrane voltage (the M-clock) and intracellular calcium concentration (the C-clock) in which both clocks are able to oscillate independently but with the M-clock normally driving the C-clock. Action potentials are thus triggered at a relatively constant basal rate but are subject to irregularities in pacing frequency due to small fluctuations in membrane potential or intracellular calcium release. Input from the Gq/11-IP3-IP3R1-Ca2+-ANO1 signaling axis results in a strong depolarizing input capable of accelerating the rate of oscillations (Hancock et al., 2022, 2023). The schematic in Figure 23 summarizes this proposed mechanism in light of our findings in the present study. The ability of some vessels to respond to pressure elevation after ANO1, IP3R1, or Gq/11 blockade indicates that a Gq/11 (and potentially G14) and/or ANO1-independent mechanism is operating, or compensating, at higher pressures. It is worth noting that as each LMC action potential lasts over 1 second, a reduction in any mechanism that reduces AP duration can ultimately result in an increased frequency regardless of the diastolic depolarization. Pressures > 8 cmH2O are levels are unlikely to be chronically experienced by mouse popliteal vessels (Davis, 2026) except perhaps in lymphedema.

Proposed mechanotransduction mechanism underlying pressure-induced lymphatic chronotropy.

Summary diagram depicting the primary signaling pathway through which changes in pressure regulate the frequency of the ionic pacemaker in LMCs. Gq/11-coupled GPCRs are the primary, but perhaps not the exclusive mechanosensor; a role for G14-coupled GPCRs is possible. Multiple GPCRs, which remain to be identified, may transduce different pressure ranges. ANO1 is the primary, but perhaps not the exclusive effector that regulates activation of Cav1.2 channels, but the pacemaker will cycle at a basal rate in the absence of ANO1 input. Kv11 provides at least one source of repolarizing signal (Kim et al., 2023).

Physiological Relevance

LMC function is impaired in Cantú syndrome, the only known example of primary lymphedema produced by a gene mutation (Kcnj8/Abbc9) in LMCs (Davis et al., 2023b). This impairment is evident both in lower contraction frequencies and amplitudes at multiple pressure levels.

The lower frequency is consistent with parallel observations that gain-of-function mutations in KATP channels increase metabolic stress (Kim et al., 2024), whereas the lower amplitude suggests that other aspects of LMC function are altered as a consequence of chronic KATP channel hyperactivation. It is interesting that lymphatic contraction frequency and amplitude are also impaired in secondary lymphedema, in aging, and in various models of metabolic disease [(Castorena-Gonzalez et al., 2025) and see discussion and references in (Davis et al., 2023f)]. Potential therapies for reversing LMC dysfunction under these conditions will depend on identifying the key ion channels and signaling pathways that drive pacemaking and control contraction amplitude. The present study represents a major step forward in the identification of specific LMC targets that could be used for that purpose.